---
title: "DADA2_to_Phyloseq_object"
author: "Joy O'Brien"
date: "2024-11-10"
output: html_document
---

```{r}
install.packages("dplyr")
install.packages("readxl")    
install.packages("ggplot2")   
install.packages("phyloseq")
if (!require("BiocManager", quietly = TRUE))
install.packages("BiocManager")
BiocManager::install("phyloseq")
```

```{r}
# Load packages 
library("phyloseq")
library("ggplot2")      # graphics
library("readxl")       # necessary to import the data from Excel file
library("dplyr")        # filter and reformat data frames
library("tibble")       # Needed for converting column to row names
library("vegan")        #Used for rarifying
```


```{r}
# Read the data into R from excel file incubation_16S_reseq.xlsx 
seq_tab.raw <- read_excel("~/GitHub/Masters_publication/meta_sequence_data.xlsx", sheet = "seqtab_final_combined") # fill this in with the file path for your DADA2 seqtab output
taxonomy <- read_excel("~/GitHub/Masters_publication/meta_sequence_data.xlsx", sheet = "tax_final_dada2merge", na = c("","NA")) # fill this in with the tax_final file DADA2 output
metadata <- read_excel("~/GitHub/Masters_publication/meta_sequence_data.xlsx", sheet = "metadata_final_combined") # fill this in with the metadata file
```

# Merge resequenced samples
```{r}
# Merge the resequenced samples
# Define sample pairs to combine 
sample_pairs <- list(
  Combined_Sample1 = c("1_S72_L002", "1_S1_L001"),
  Combined_Sample15 = c("15_S86_L002", "10_S10_L001"),
  Combined_Sample13 = c("13_S84_L002", "12_S12_L001"), 
  Combined_Sample40 = c("40_S111_L002", "13_S13_L001"), 
  Combined_Sample41 = c("41_S112_L002", "14_S14_L001"), 
  Combined_Sample42 = c("42_S113_L002","15_S15_L001"), 
  Combined_Sample48 = c("16_S16_L001", "48_S119_L002"), 
  Combined_Sample24 = c("24_S95_L002", "17_S17_L001"),
  Combined_Sample3 = c("3_S74_L002", "2_S2_L001"),
  Combined_Sample6 = c("3_S3_L001", "6_S77_L002"), 
  Combined_Sample32 = c("4_S4_L001", "32_S103_L002"),
  Combined_Sample7 = c("5_S5_L001", "7_S78_L002"),
  Combined_Sample8 = c("6_S6_L001", "8_S79_L002"), 
  Combined_Sample9 = c("8_S8_L001", "9_S80_L002"), 
  Combined_Sample10 = c("9_S9_L001", "10_S81_L002")
)

combined_seq_tab <- data.frame(row.names = rownames(seq_tab.raw))

# Sum specified columns for each pair and add as new combined columns
for (new_sample in names(sample_pairs)) {
  combined_seq_tab[[new_sample]] <- rowSums(seq_tab.raw[, sample_pairs[[new_sample]], drop = FALSE])
}

# View the combined ASV count table
print(combined_seq_tab)

# Saving the combined_seq_tab output because I am going to manually manipulate in excel for replacing the samples in the seq tab 
write.table(combined_seq_tab, "~/Desktop/combined_seq_tab.tsv", sep = "\t", row.names = TRUE, col.names = TRUE)

```

```{r}
# Define row names from the ASV column 
seq_tab <- seq_tab.raw %>%
  tibble::column_to_rownames("ASV_ID") 

taxonomy <- taxonomy %>%
  tibble::column_to_rownames("ASV_ID") 

metadata <- metadata %>% 
  tibble::column_to_rownames("sample_name") %>% 
  mutate(sample_blank = ifelse(test = grepl("BLANK", sample_ID), no = "Sample", yes = "BLANK")) %>%
  mutate(core_rep = gsub("_POST_SOIL", "", sample_ID)) %>%
  mutate(core_rep = gsub("_PRE_SOIL", "", core_rep))
```



```{r}
# Transform into matrices 
seq_tab <- as.matrix(seq_tab)
taxonomy <- as.matrix(taxonomy)

# Transform to phyloseq objects
ASV <- otu_table(seq_tab, taxa_are_rows = TRUE) 
TAX <- tax_table(taxonomy)
samples = sample_data(metadata)

# Create the phyloseq object
dorm <- phyloseq(ASV, TAX, samples) 

# Prune samples from the phyloseq object dorm created above, and call the pruned object dorm1
dorm1 <- prune_samples(sample_sums(dorm) > 0,dorm)
dorm1 #19617 taxa

# Remove taxa that are unassigned at the Phylum level
dorm1 <- subset_taxa(dorm1, !is.na(Phylum)) # Note that sample blanks are still included here 
dorm1 # went from 19617 to 18565 (decrease due to deleting those that are unassigned at phylum level)

# The number of taxa that were removed from being unassigned at the Phylum level: 19617-18565 = 1052
# Filtering chloroplasts and mitochondria
dorm1 <- subset_taxa(dorm1, Family != "Mitochondria")
dorm1 <- subset_taxa(dorm1, Order != "Chloroplast") 
# check to see if mitochondria and chloroplasts were removed
dorm1  #10721 taxa
# Number of taxa removed from mitochondria and chloroplasts: 18565-10721 = 7,844 # calculate as % reads for publication
```

```{r}
# Remove sequence blanks from the data set 
dorm_noblank <- subset_samples(dorm1, sample_blank == "Sample")

# Check to see if samples were removed (should be 5 for the re-seq round)
dorm_noblank # this checks out, 5 samples were removed
```
```{r}
# Extract the OTU table from the phyloseq object
otu_table <- otu_table(dorm1)

# Convert the OTU table to a matrix if it's not already
otu_matrix <- as(otu_table, "matrix")

# Calculate the sum of ASV counts per sample
asv_sums <- colSums(otu_matrix)

# Print the results
print(asv_sums)
hist(asv_sums)

```

```{r}

# Print the sums per sample
print(asv_sums)
# Rarefying the data 
otu_rarify <- as.data.frame(dorm_noblank@otu_table)
rarecurve(otu_rarify, step = 50, cex = 0.5, label = FALSE)
dorm_noblank.rarefied <- rarefy_even_depth(dorm_noblank, rngseed = 1, sample.size = 4000, replace = FALSE)

# Now let's check the rarifaction plot after rarifying 
rarify.check <- as.data.frame(dorm_noblank.rarefied@otu_table)
rarecurve(rarify.check, step = 50, cex = 0.5, label = FALSE)

# Visualize with a histogram
hist(sort(colSums(otu_table(dorm_noblank)))) 

# Create the rarified OTU table
total_otu_table_noblank <- otu_table(dorm_noblank.rarefied)
```

```{r}
# Transform the OTU table 
dorm_transformed <- t(sqrt(total_otu_table_noblank))

# Create a distance matrix using vegan and bray
dm_dorm <- vegdist(dorm_transformed, method = "bray")
# I did the following to check how many samples made it through rarifaction
dm_mat <- as.matrix(dm_dorm)

# Run an NMDS (stress values indicate how well the variation is represented, stress less than 0.05 is good, below 0.3 is poor, below 0.2 is okay) pp: NIce comment, here's the link for that info: https://jonlefcheck.net/2012/10/24/nmds-tutorial-in-r/
dorm.nmds <- metaMDS(dm_dorm, k = 2, trymax = 1500) ## "k" is the number of axes, sometimes 2 is not enough, "trymax" is the number of times it tries to find solutions
print(dorm.nmds)
plot(dorm.nmds)
```


```{r}
plot_ordination(
  physeq = dorm_noblank.rarefied,
  ordination = dorm.nmds,
  color = "site",
  shape = "pre_post_thaw",
  #title = "NMDS of Permafrost Bacterial Communities"
) +
  geom_point(aes(color = site), alpha = 0.7, size = 4) +
  geom_point(size = 2)+
  geom_text(aes(label = core), size = 4, color = "black") +
  theme_classic() +
  theme(axis.text = element_text(size = 18)) +
  scale_color_discrete("Site") +
  scale_shape_discrete("Treatment") +
  scale_shape_manual("Treatment", labels = c("Pre-thaw", "Post-thaw"), values = c(2, 1))

# Visualize with a histogram
hist(sort(colSums(otu_table(dorm_noblank)))) #H: this is nice! I'll in

```
```{r}
# For publication
library(ggplot2)
library(RColorBrewer)

plot_ordination(
  physeq = dorm_noblank.rarefied,
  ordination = dorm.nmds,
  color = "site",
  shape = "pre_post_thaw"
) +
  geom_point(aes(color = site), alpha = 0.7, size = 4) +
  geom_point(size = 2) +
  theme_classic() +
  theme(axis.text = element_text(size = 18)) +
  scale_color_brewer(palette = "Set1", name = "Site") +
  scale_shape_discrete("Treatment") +
  scale_shape_manual("Treatment", labels = c("Pre-thaw", "Post-thaw"), values = c(2, 1))




```
```



```{r}
# Save the phyloseq object to yourfolder if this is your final object (save individual parts of the phyloseq object)
saveRDS(dorm_noblank.rarefied,"~/GitHub/Masters_publication/Masters_publication_R/dorm1rarefied_reseq.rds")
saveRDS(dorm_noblank.rarefied@otu_table,"~/GitHub/Masters_publication/Masters_publication_R/dorm1rarefied_OTU_merged.rds")
saveRDS(dorm_noblank.rarefied@tax_table,"~/GitHub/Masters_publication/Masters_publication_R/dorm1rarefied_tax_table_merged.rds" )
saveRDS(dorm_noblank.rarefied@sam_data,"~/GitHub/Masters_publication/Masters_publication_R/dorm1rarefied_sam_data_merged.rds")
```


```{r}
