---
title: "2.Biotic_Abiotic_soil_properties"
author: "Joy O'Brien"
date: "2025-03-03"
output: html_document
---
THIS IS THE ANOVAS FOR SITE COMPARISON SCIPT
Data we are working with here: We will be running ANOVAs on DOC data post thaw, TDN data post thaw, 
copy numbers pre thaw and copy numbers post thaw and cum respiration across sites

The reason we are NOT including pre thaw data here is because the pre thaw data is not properly replicated (due to lack of soil to replicate). 
#```{r setup, include=FALSE}
#knitr::opts_knit$set(root.dir = "Users/joyobrien/GitHub/thaw.response")
#```

```{r}
# Load libraries
library(vegan)
library(readxl)
library(dplyr)
library(FSA) #installed for the dunntest
library(multcompView) # used for the tukey test
library(ggplot2) # to make pretty figs 
library(rstatix) 
library(multcompView)
library(rcompanion) # attempted to use this to get the significance codes from KW test
library(agricolae) # used to get the significance codes from the KW test
```
1. Load data
```{r}
# Read in the data (this is all accessible in the incubation_physical_chemical.xlsx sheet on Ernakovich Lab Box Folder)
resp <- read_excel("~/GitHub/thaw.response/Data/thaw_physical_chemical_data.xlsx", sheet = "Respiration_cum")
pH <- read_excel("~/GitHub/thaw.response/Data/thaw_physical_chemical_data.xlsx", sheet = "pH")
EC <- read_excel("~/GitHub/thaw.response/Data/thaw_physical_chemical_data.xlsx", sheet = "EC")
TC_TN <- read_excel("~/GitHub/thaw.response/Data/thaw_physical_chemical_data.xlsx", sheet = "TC_TN")
GWC <- read_excel("~/GitHub/thaw.response/Data/thaw_physical_chemical_data.xlsx", sheet = "GWC")
resp_tctn <- read_excel("~/GitHub/thaw.response/Data/thaw_physical_chemical_data.xlsx", sheet = "Resp_TC_TN")
cpnumb <- read_excel("~/GitHub/thaw.response/Data/thaw_physical_chemical_data.xlsx", sheet = "copynumber_all") 
```

PART 1. COMPARING PRE-THAW AND POST-THAW ABIOTIC PROPERTIES

1. pH
```{r}
# Subset data into pH for pre- and post-thaw samples
pH_pre_thaw <- pH[pH$pre_post_thaw == "pre", ]
pH_post_thaw <- pH[pH$pre_post_thaw == "post", ]

# Checking normality 
shapiro.test(pH_pre_thaw$pH) # 0.3505 means that the data is normally distributed so we can run ANOVA

# Run ANOVA
phpreaov <- aov(pH ~ site, data = pH_pre_thaw)
print(phpreaov)
summary(phpreaov)

#> summary(phpreaov)
#            Df Sum Sq Mean Sq F value Pr(>F)  
#site         2 0.3742  0.1871   16.56 0.0239 *
#Residuals    3 0.0339  0.0113 

# Run TUKEY
phpretukey<- TukeyHSD(phpreaov, conf.level = 0.95)
print(phpretukey)

# Obtain significance codes for TUKEY
tukey_cld_pre_ph <-multcompLetters4(phpreaov, phpretukey)
print(tukey_cld_pre_ph)
# $site
#diff        lwr          upr     p adj
#FL-CRREL        -0.155 -0.5992086  0.289208596 0.4216976
#Utqiagvik-CRREL -0.590 -1.0342086 -0.145791404 0.0233858
#Utqiagvik-FL    -0.435 -0.8792086  0.009208596 0.0527796

# pH post-thaw ANOVA 

# Checking normality
shapiro.test(pH_post_thaw$pH) # p-value = 0.05193 WHEW
# ANOVA
phpostaov <- aov(pH ~ site, data = pH_post_thaw)
print(phpostaov)
summary(phpostaov)

# Run TUKEY
phposttukey<- TukeyHSD(phpostaov, conf.level = 0.95)
print(phposttukey)
  #$site
  #diff        lwr         upr     p adj
  #FL-CRREL        -0.140 -0.3624316  0.08243164 0.1510352
  #Utqiagvik-CRREL -0.855 -1.0774316 -0.63256836 0.0011239
  #Utqiagvik-FL    -0.715 -0.9374316 -0.49256836 0.0018593
```


2. EC
```{r setup, include=FALSE}
# NOW WE WILL LOOK AT ELECTRICAL CONDUCTIVITY (EC)
# Subset the data 
EC_pre_thaw <- EC[EC$pre_post_thaw == "pre", ]
EC_post_thaw <- EC[EC$pre_post_thaw == "post", ]

# Checking for normality
shapiro.test(EC_pre_thaw$EC) # p-value = 0.8672 
# ANOVA
ecpreaov <- aov(EC ~ site, data = EC_pre_thaw)
summary(ecpreaov)

# Not significant so we don't need to run a Tukey

# EC post-thaw ANOVA 

# Checking for normality
shapiro.test(EC_post_thaw$EC) # p-value = 0.4292
# ANOVA
ecpostaov <- aov(EC ~ site, data = EC_post_thaw)
summary(ecpostaov)
# TUKEY
ECposttukey <- TukeyHSD(ecpostaov, conf.level = 0.95)
print(ECposttukey)

# This library allows us to view the significance codes associated with TUKEY
library(multcompView)

# Viewing results with significance codes
tukey_cld <-multcompLetters4(ecpostaov, ECposttukey)
print(tukey_cld)
```

3. %C and %N
```{r setup, include=FALSE}
# Comparing % C across site
# Checking for normality
shapiro.test(TC_TN$C) # p = 0.2355
TCaov <- aov(C ~ site, data = TC_TN)
summary(TCaov)
# Tukey
TCtukey <- TukeyHSD(TCaov, conf.level = 0.95)
print(TCtukey)

# obtaining results with significance codes
tukey_tc_tld <-multcompLetters4(TCaov, TCtukey)
print(tukey_tc_tld)

# Comparing % N across site 
# Checking for normality
shapiro.test(TC_TN$N) # p = 0. 3077 
# ANOVA
TNaov <- aov(N ~ site, data = TC_TN)
summary(TNaov)
# Tukey
TNtukey <- TukeyHSD(TNaov, conf.level = 0.95)
print(TNtukey)
# obtaining results with significance codes
tukey_tn <-multcompLetters4(TNaov, TNtukey)
```

4. GWC
```{r setup, include=FALSE}
# Now we are going to look at GWC across sites pre and post thaw 
GWC_pre_thaw <- GWC[GWC$pre_post_thaw == "pre", ]
GWC_post_thaw <- GWC[GWC$pre_post_thaw == "post", ]
# Pre-thaw GWC
# Checking for normality
shapiro.test(GWC_pre_thaw$GWC) # p-value = 0.6917
# ANOVA
GWC_preaov <- aov(GWC ~ site, data = GWC_pre_thaw)
summary(GWC_preaov) # Not significant across sites, so I will not run a Tukey

# Post-thaw GWC
# normality
shapiro.test(GWC_post_thaw$GWC) # p-value = 0.4414
# ANOVA
GWC_postaov <- aov(GWC ~ site, data = GWC_post_thaw)
summary(GWC_postaov) 

# Comparing pre and post thaw using the original GWC dataset
# Checking if normality is met for both pre and post thaw
shapiro.test(GWC$GWC) # Test for overall normality

# ANOVA to compare GWC between pre and post thaw across sites
GWC_change <- aov(GWC ~ pre_post_thaw * site, data = GWC)
summary(GWC_change)  # Main effects for thaw treatment and site, and interaction

# If significant results, perform Tukey's HSD test to compare levels of "pre_post_thaw" 
gwctukey <- TukeyHSD(GWC_change)
print(gwctukey)  # Tukey's post-hoc results
```


PART 2: COMPARING PRE AND POST THAW BIOTIC SOIL PROPERTIES

1. Cumulative Respiration
```{r}
# Let's look at the differences in cumulative respiration across sites 
shapiro.test(resp$Cumulative_Respiration) #this is not normally distributed W = 0.89709, p-value = 0.01868
# Because of that, lets look at a Kruskal test
kruskal.test(Cumulative_Respiration ~ site, data = resp) # Kruskal-Wallis chi-squared = 16.805, df = 2, p-value = 0.0002243
kruskal(resp$Cumulative_Respiration, resp$site, group=TRUE, p.adj = "bonferroni")$groups # use this to get significance codes for KW test
# CUMMULATIVE RESPIRATION IS STATISTICALLY SIGNIFICANT ACROSS SITES

# Now let's visualize with a box plot 
#boxplot(Cumulative_Respiration ~ site, data = resp)
library(ggplot2)
ggplot(resp, aes(x = site, y = Cumulative_Respiration)) + 
  geom_boxplot() +
  theme_classic() +
  xlab("Site") +
  ylab("Cumulative Respiration (µg C-CO2 g-1 dry soil)") +
  theme(text = element_text(size = 12))
# Let's run a dunn test for respiration and site 
dunn.resp <- dunnTest(Cumulative_Respiration ~ site, data = resp, method = 'bh')
```

```{r}
# Making the plot prettier for publication
library(ggsignif)
# Re-order the sites in the plot to tell the story
resp$site <- factor(resp$site, levels = c("CRREL", "Utqiagvik", "FL"))

# Get the statistical significance codes

ggplot(resp, aes(x = site, y = Cumulative_Respiration, fill = site)) + 
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +  # Remove outliers for clarity
  geom_jitter(width = 0.2, alpha = 0.6, size = 2) + # Add jittered points
  scale_fill_manual(values = c("#1b9e77", "#d95f02", "#7570b3")) +  # Custom colors
  theme_classic() +
  labs(x = "Site", 
       y = "Cumulative Respiration (µg C-CO2 g-1 dry soil)") +
  theme(
    text = element_text(size = 16),  
    axis.text = element_text(size = 12, color = "black"),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    legend.position = "none"  # Remove legend if unnecessary
  ) +
    geom_signif(
    comparisons = list(c("CRREL", "FL"), c("CRREL", "Utqiagvik")),  # Specify the groups to compare
    map_signif_level = TRUE,  # Automatically display significance codes like "*", "**", etc.
    annotations = c("**", "*"),  # You can manually annotate with custom significance labels
    y_position = c(1560,1560),  # Adjust y-position of the annotations
    tip_length = 0.03,
    textsize = 8                   # Adjust the length of the annotation line
  )
```

3. Copy number analysis pre and post thaw

```{r}
# Subset the data first 
pre_thaw_copy <- cpnumb[cpnumb$pre_post_thaw == "pre", ]
post_thaw_copy <- cpnumb[cpnumb$pre_post_thaw == "post", ]

# Run the Shapiro-Wilk test on the copy_number_per_gsoil for pre-thaw
shapiro.test(pre_thaw_copy$copy_numb) #p-value = 0.007, W = 0.86772

# Run the Shapiro-Wilk test on the copy number for post-thaw 
shapiro.test(post_thaw_copy$copy_numb) #p-value = 6.147e06, W = 0.65905

# Kruskal on pre thaw copy number and site
kruskal.test(copy_numb ~ site, data = pre_thaw_copy) # Kruskal-Wallis chi-squared = 2.2237, df = 2, p-value = 0.3289

# Kruskal on post thaw copy number and site
kruskal.test(copy_numb ~ site, data = post_thaw_copy) # Kruskal-Wallis chi-squared = 7.8511, df = 2, p-value = 0.01973

#Visualizing pre thaw copy number by site
boxplot(copy_numb ~ site, data = cpnumb)
# Visualizing post thaw copy number by site 
boxplot(copy_numb ~ site, data = post_thaw_copy)
```




4. GLM with C and N
```{r}
# Trying GLM since this data is normally distributed 
# C
linearmodel_Cresp <- lm(Cumulative_Respiration ~ C, data = resp_tctn)

print(linearmodel_Cresp)
summary(linearmodel_Cresp)

# Define custom site colors (same as in boxplot)
site_colors <- c("CRREL" = "#1b9e77", "FL" = "#d95f02", "Utqiagvik" = "#7570b3")

ggplot(resp_tctn, aes(x = C, y = Cumulative_Respiration)) + 
  geom_point(aes(color = site), size = 3.5, alpha = 0.7) +  # Adjust transparency to match jitter points
  scale_color_manual(values = site_colors, name = "Site") +  # Custom site colors
  theme_classic() +
  labs(
    x = "% C", 
    y = "Cumulative Respiration (µg C-CO2 g-1 dry soil)"
  ) +
  geom_smooth(method = "lm", color = "dark gray", se = TRUE) +  # Regression line
  theme(
    text = element_text(size = 16),  
    axis.text = element_text(size = 12, color = "black"),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5)
  )

```


```{r}
# N
linearmodel_Nresp <- lm(Cumulative_Respiration ~ N, data = resp_tctn)
summary(linearmodel_Nresp)
print(linearmodel_Nresp)


shapiro.test(resp_tctn$N)
Nanova <- aov(N ~ Cumulative_Respiration, data = resp_tctn)
summary(Nanova)


# Define custom site colors (same as in boxplot)
site_colors <- c("CRREL" = "#1b9e77", "FL" = "#d95f02", "Utqiagvik" = "#7570b3")

ggplot(resp_tctn, aes(x = N, y = Cumulative_Respiration)) + 
  geom_point(aes(color = site), size = 3.5, alpha = 0.7) +  # Adjust transparency to match jitter points
  scale_color_manual(values = site_colors, name = "Site") +  # Custom site colors
  theme_classic() +
  labs(
    x = "% N", 
    y = "Cumulative Respiration (µg C-CO2 g-1 dry soil)"
  ) +
  geom_smooth(method = "lm", color = "dark gray", se = TRUE) +  # Regression line
  theme(
    text = element_text(size = 16),  
    axis.text = element_text(size = 12, color = "black"),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5)
  )
```



```{r}
# Checking the relationship using all of the replicates (Jessica asked to make sure that this relationship is strong)
respcheck <- read_excel("~/Desktop/incubation_physical_chemical.xlsx", sheet = "Resp_TC_TN_check")

shapiro.test(respcheck$Cumulative_Respiration) # not normally distributed
resp_C_corr_test <- cor.test(x =respcheck$C, y = respcheck$Cumulative_Respiration, method = "spearman")
print(resp_C_corr_test)


ggplot(respcheck, aes(x = C, y = Cumulative_Respiration)) + 
  geom_point(aes(color = site), size = 2.5) +
  theme_classic() +
  xlab("% C") +
  ylab("Cumulative Respiration (µg C-CO2 g-1 dry soil)") +
  geom_smooth(method = "lm", color = "dark gray", se = TRUE) +
  theme(text = element_text(size = 12)) +
  scale_color_discrete("Site") #changing the name of the legend title

linearmodel_Crespcheck <- lm(Cumulative_Respiration ~ C, data = respcheck)
print(linearmodel_Crespcheck)
summary(linearmodel_Crespcheck)


```




