---
title: "2.Biotic_Abiotic_soil_properties"
author: "Joy O'Brien"
date: "2025-03-03"
output: html_document
---
 
Set wd
```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "Users/joyobrien/GitHub/thaw.response")
```

(set theme)
```{r}
library(paletteer)
library(ggplot2)
theme_set(theme_classic() +
  theme(axis.title = element_text(size = 16),
        axis.title.x = element_text(margin = margin(t = 15, b = 15)),
        axis.title.y = element_text(margin = margin(l = 15, r = 15)),
        axis.text = element_text(size = 13),
        axis.text.x = element_text(margin = margin(t = 5)),
        axis.text.y = element_text(margin = margin(r = 5)),
        #axis.line.x = element_line(linewidth = 1),
        #axis.line.y = element_line(linewidth = 1),
        axis.ticks.y.right = element_line(linewidth = 1),
        axis.text.y.right = element_text(size = 13, margin = margin(l = 5)),
        axis.line.y.right = element_line(linewidth = 0.1),
        axis.ticks.x.top = element_line(linewidth = 1),
        axis.text.x.top = element_text(size = 13, margin = margin(b = 5)),
        axis.line.x.top = element_line(linewidth = 0.1),
        axis.line.x = element_blank(),
        axis.line.y = element_blank(),
        axis.ticks.x = element_line(linewidth = 1),
        axis.ticks.y = element_line(linewidth = 1),
        axis.ticks.length = unit(.1, "in"),
        panel.border = element_rect(color = "grey20", fill = NA, linewidth = 1.5),
        legend.text = element_text(size = 14),
        strip.text = element_text(size = 14),
        strip.background = element_blank()
        )) 
colors <-paletteer_d("ggthemes::Miller_Stone")
```
1. Load libraries
```{r}
# Load libraries
library(vegan)
library(readxl)
library(dplyr)
library(FSA) #installed for the dunntest
library(multcompView) # used for the tukey test
library(ggplot2) # to make pretty figs 
library(rstatix) 
library(multcompView)
library(rcompanion) # attempted to use this to get the significance codes from KW test
library(agricolae) # used to get the significance codes from the KW test
```
2. Load data
```{r}
# Read in the data (this is all accessible in the incubation_physical_chemical.xlsx sheet
resp <- read_excel("~/GitHub/thaw.response/Data/thaw_physical_chemical_data.xlsx", sheet = "Respiration_cum") #cumulative respiration
pH <- read_excel("~/GitHub/thaw.response/Data/thaw_physical_chemical_data.xlsx", sheet = "pH") #pH
EC <- read_excel("~/GitHub/thaw.response/Data/thaw_physical_chemical_data.xlsx", sheet = "EC") # electrical conductivity
TC_TN <- read_excel("~/GitHub/thaw.response/Data/thaw_physical_chemical_data.xlsx", sheet = "TC_TN_raw_edit") # total carbon and nitrogen
GWC <- read_excel("~/GitHub/thaw.response/Data/thaw_physical_chemical_data.xlsx", sheet = "GWC") #gravimentric water content
resp_tctn <- read_excel("~/GitHub/thaw.response/Data/thaw_physical_chemical_data.xlsx", sheet = "Resp_TC_TN") # cumulative respiration with total carbon and total nitrogen
cpnumb <- read_excel("~/GitHub/thaw.response/Data/thaw_physical_chemical_data.xlsx", sheet = "copynumber_all")  #16S rRNA copy number
```

PART 1. COMPARING PRE-THAW AND POST-THAW ABIOTIC PROPERTIES

1. pH
```{r}
# Subset data into pH for pre- and post-thaw samples
pH_pre_thaw <- pH[pH$pre_post_thaw == "pre", ]
pH_post_thaw <- pH[pH$pre_post_thaw == "post", ]

# Checking normality 
shapiro.test(pH_pre_thaw$pH) # 0.3505 means that the data is normally distributed so we can run ANOVA

# Run ANOVA
phpreaov <- aov(pH ~ site, data = pH_pre_thaw)
print(phpreaov)
summary(phpreaov)

#> summary(phpreaov)
#            Df Sum Sq Mean Sq F value Pr(>F)  
#site         2 0.3742  0.1871   16.56 0.0239 *
#Residuals    3 0.0339  0.0113 

# Run TUKEY
phpretukey<- TukeyHSD(phpreaov, conf.level = 0.95)
print(phpretukey)

# Obtain significance codes for TUKEY
tukey_cld_pre_ph <-multcompLetters4(phpreaov, phpretukey)
print(tukey_cld_pre_ph)
# $site
#diff        lwr          upr     p adj
#FL-CRREL        -0.155 -0.5992086  0.289208596 0.4216976
#Utqiagvik-CRREL -0.590 -1.0342086 -0.145791404 0.0233858
#Utqiagvik-FL    -0.435 -0.8792086  0.009208596 0.0527796

# pH post-thaw ANOVA 

# Checking normality
shapiro.test(pH_post_thaw$pH) # p-value = 0.05193 WHEW
# ANOVA
phpostaov <- aov(pH ~ site, data = pH_post_thaw)
print(phpostaov)
summary(phpostaov)

# Run TUKEY
phposttukey<- TukeyHSD(phpostaov, conf.level = 0.95)
print(phposttukey)
  #$site
  #diff        lwr         upr     p adj
  #FL-CRREL        -0.140 -0.3624316  0.08243164 0.1510352
  #Utqiagvik-CRREL -0.855 -1.0774316 -0.63256836 0.0011239
  #Utqiagvik-FL    -0.715 -0.9374316 -0.49256836 0.0018593
```

2. EC
```{r}
# NOW WE WILL LOOK AT ELECTRICAL CONDUCTIVITY (EC)
# Subset the data 
EC_pre_thaw <- EC[EC$pre_post_thaw == "pre", ]
EC_post_thaw <- EC[EC$pre_post_thaw == "post", ]

# Checking for normality
shapiro.test(EC_pre_thaw$EC) # p-value = 0.8672 
# ANOVA
ecpreaov <- aov(EC ~ site, data = EC_pre_thaw)
summary(ecpreaov)

# Not significant so we don't need to run a Tukey

# EC post-thaw ANOVA 

# Checking for normality
shapiro.test(EC_post_thaw$EC) # p-value = 0.4292
# ANOVA
ecpostaov <- aov(EC ~ site, data = EC_post_thaw)
summary(ecpostaov)
# TUKEY
ECposttukey <- TukeyHSD(ecpostaov, conf.level = 0.95)
print(ECposttukey)

# This library allows us to view the significance codes associated with TUKEY
library(multcompView)

# Viewing results with significance codes
tukey_cld <-multcompLetters4(ecpostaov, ECposttukey)
print(tukey_cld)
```

3. %C stats
```{r}
# Comparing % C across site
# Checking for normality
shapiro.test(TC_TN$C) # p = 0.001
TC.kruskal <- kruskal.test(C ~ site, data = TC_TN)
print(TC.kruskal) #p = 0.001
dunnTest(C ~ site, data = TC_TN, method = 'bh')
```

3A. %N stats
```{r}
# Now for N
shapiro.test(TC_TN$N) # p = 0.004 so this is not normally distributed data
TN.kruskal <- kruskal.test(N ~ site, data = TC_TN)
print(TN.kruskal) #p = 0.001
dunnTest(N ~ site, data = TC_TN, method = 'bh')
```

4. GWC
```{r}
# Now we are going to look at GWC across sites pre and post thaw 
GWC_pre_thaw <- GWC[GWC$pre_post_thaw == "pre", ]
GWC_post_thaw <- GWC[GWC$pre_post_thaw == "post", ]
# Pre-thaw GWC
# Checking for normality
shapiro.test(GWC_pre_thaw$GWC) # p-value = 0.6917
# ANOVA
GWC_preaov <- aov(GWC ~ site, data = GWC_pre_thaw)
summary(GWC_preaov) # Not significant across sites, so I will not run a Tukey

# Post-thaw GWC
# normality
shapiro.test(GWC_post_thaw$GWC) # p-value = 0.4414
# ANOVA
GWC_postaov <- aov(GWC ~ site, data = GWC_post_thaw)
summary(GWC_postaov) 

# Comparing pre and post thaw using the original GWC dataset
# Checking if normality is met for both pre and post thaw
shapiro.test(GWC$GWC) # Test for overall normality

# ANOVA to compare GWC between pre and post thaw across sites
GWC_change <- aov(GWC ~ pre_post_thaw * site, data = GWC)
summary(GWC_change)  # Main effects for thaw treatment and site, and interaction

# If significant results, perform Tukey's HSD test to compare levels of "pre_post_thaw" 
gwctukey <- TukeyHSD(GWC_change)
print(gwctukey)  # Tukey's post-hoc results
```


PART 2: COMPARING PRE AND POST THAW BIOTIC SOIL PROPERTIES

1. Cumulative Respiration
```{r}
# Renaming sites in Resp dataset 
library(dplyr)

resp <- resp %>%
  mutate(site = recode(site,
                       "CRREL" = "PT",
                       "Utqiagvik" = "BEO",
                       "FL" = "FL"))

# Let's look at the differences in cumulative respiration across sites 
shapiro.test(resp$Cumulative_Respiration) #this is not normally distributed W = 0.89709, p-value = 0.01868
# Because of that, lets look at a Kruskal test
kruskal.test(Cumulative_Respiration ~ site, data = resp) # Kruskal-Wallis chi-squared = 16.805, df = 2, p-value = 0.0002243
kruskal(resp$Cumulative_Respiration, resp$site, group=TRUE, p.adj = "bonferroni")$groups # use this to get significance codes for KW test
# CUMMULATIVE RESPIRATION IS STATISTICALLY SIGNIFICANT ACROSS SITES

# Now let's visualize with a box plot 
#boxplot(Cumulative_Respiration ~ site, data = resp)
library(ggplot2)
ggplot(resp, aes(x = site, y = Cumulative_Respiration)) + 
  geom_boxplot() +
  theme_classic() +
  xlab("Site") +
  ylab("Cumulative Respiration (µg C-CO2 g-1 dry soil)") +
  theme(text = element_text(size = 12))
# Let's run a dunn test for respiration and site 
dunn.resp <- dunnTest(Cumulative_Respiration ~ site, data = resp, method = 'bh')
print(dunn.resp)
```
1A. Plotting cumulative respiration
```{r}
# Making the plot prettier for publication
library(ggsignif)
# Re-order the sites in the plot to tell the story
resp$site <- factor(resp$site, levels = c("PT", "BEO", "FL"))

# Get the statistical significance codes

ggplot(resp, aes(x = site, y = Cumulative_Respiration, fill = site)) + 
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +  # Remove outliers for clarity
  geom_jitter(width = 0.2, alpha = 0.6, size = 2) + # Add jittered points
  scale_fill_manual(values = c("#1b9e77", "#7570b3", "#d95f02")) +  # Custom colors
  theme_classic() +
  scale_y_continuous(limits = c(0,1900)) +
  labs(x = "Site", 
        y = expression("Cumulative Respiration (" * mu * g~CO[2]*"-C"~g^-1~"dry soil)")) +
  theme(
    text = element_text(size = 12),  
    axis.text = element_text(size = 12, color = "black"),
    #axis.title = element_text(size = 12, face = "bold"),
    #plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    legend.position = "none", 
        #plot.margin = margin(50, 50, 50, 50)  # top, right, bottom, left (in points)
    # Remove legend if unnecessary
 ) +
    geom_signif(
    comparisons = list(c("PT", "FL"), c("PT", "BEO")),  # Specify the groups to compare
    map_signif_level = TRUE,  # Automatically display significance codes like "*", "**", etc.
    annotations = c("***", "**"),  # Manually annotate with custom significance labels
    y_position = c(1680,1730),  # Adjust y-position of the annotations
    tip_length = 0.03,
    textsize = 8                   # Adjust the length of the annotation line
  )
```
3. Copy number analysis pre and post thaw
```{r}
# Subset the data first 
pre_thaw_copy <- cpnumb[cpnumb$pre_post_thaw == "pre", ]
post_thaw_copy <- cpnumb[cpnumb$pre_post_thaw == "post", ]

# Run the Shapiro-Wilk test on the copy_number_per_gsoil for pre-thaw
shapiro.test(pre_thaw_copy$copy_numb) #p-value = 0.007, W = 0.86772

# Run the Shapiro-Wilk test on the copy number for post-thaw 
shapiro.test(post_thaw_copy$copy_numb) #p-value = 6.147e06, W = 0.65905

# Kruskal on pre thaw copy number and site
kruskal.test(copy_numb ~ site, data = pre_thaw_copy) # Kruskal-Wallis chi-squared = 2.2237, df = 2, p-value = 0.3289

# Kruskal on post thaw copy number and site
kruskal.test(copy_numb ~ site, data = post_thaw_copy) # Kruskal-Wallis chi-squared = 7.8511, df = 2, p-value = 0.01973

#Visualizing pre thaw copy number by site
boxplot(copy_numb ~ site, data = cpnumb)
# Visualizing post thaw copy number by site 
boxplot(copy_numb ~ site, data = post_thaw_copy)
```

3A. Looking at change in copy number pre_post thaw
```{r}
# First remove the PRE_SOIL and POST_SOIL from the names of the samples in pre_thaw copy and post_thaw copy
post_thaw_copy$sample_ID <- sub("_POST_SOIL", "", post_thaw_copy$sample_ID)
pre_thaw_copy$sample_ID <- sub("_PRE_SOIL", "", pre_thaw_copy$sample_ID)
merged<- merge(pre_thaw_copy, post_thaw_copy, by = "sample_ID")

# Keep all columns from pre_thaw_copy, and just the columns needed from post_thaw_copy

post <- post_thaw_copy[, c("sample_ID", "copy_numb")]
colnames(post)[colnames(post) == "copy_numb"] <- "post_copy"

pre <- pre_thaw_copy
colnames(pre)[colnames(pre) == "copy_numb"] <- "pre_copy"

# Merge, keeping one copy of shared metadata
merged <- merge(pre, post, by = "sample_ID")

merged$change_in_copy <- (merged$post_copy - merged$pre_copy)

# Remove outlier
merged <- merged[-c(9), ]

# Rename the sites for publication
library(dplyr)

merged <- merged %>%
  mutate(site = recode(site,
                       "CRREL" = "PT",
                       "Utqiagvik" = "BEO",
                       "FL" = "FL"))
```
3B. Testing for significant differences of change in copy number
```{r}
kruskal.test(change_in_copy ~ site, data = merged)  # non-parametric
dunnTest(change_in_copy ~ site, data = merged, method = 'bh')
```

3C.Change in copy number figure
```{r}
# Changing the order that they appear
merged$site <- factor(merged$site, levels = c("PT", "BEO", "FL"))
ggplot(merged, aes(x = site, y = change_in_copy, fill = site)) + 
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +  # Remove outliers for clarity
  geom_jitter(width = 0.2, alpha = 0.6, size = 2) + # Add jittered points
  scale_fill_manual(values = c("#1b9e77", "#7570b3", "#d95f02")) +  # Custom colors
  theme_classic() +
  scale_y_continuous(limits = c(0, 4500000)) +
  labs(x = "Site", 
       y = "Change in 16S copy number (copy number per g soil)") +
  theme(
    text = element_text(size = 12),  
    axis.text = element_text(size = 12, color = "black"),
    #axis.title = element_text(size = 12, face = "bold"),
    #plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    legend.position = "none") +  # Remove legend 
    geom_signif(
    comparisons = list(c("PT", "FL"), c("PT", "BEO")),  # Specify the groups to compare
    map_signif_level = TRUE,  # Automatically display significance codes like "*", "**", etc.
    annotations = c("*", "*"),  # Manually annotate with custom significance labels
    y_position = c(4000000,4095000),  # Adjust y-position of the annotations
    tip_length = 0.03,
    textsize = 8)


```
3C. Change in copy number and site stats
```{r}
kruskal.test(change_in_copy ~ site, data = merged)  # non-parametric
dunnTest(change_in_copy ~ site, data = merged, method = 'bh')

```


C and N: Plotting Raw data on the top (do box plot, I don't think it makes sense to correlate cumulative resp with pre-thaw %C and %N)
```{r}
# Rename the sites for publication
library(dplyr)

TC_TN <- TC_TN %>%
  mutate(site = recode(site,
                       "CRREL" = "PT",
                       "Utqiagvik" = "BEO",
                       "FL" = "FL"))
# Reordering appearance of site
TC_TN$site <- factor(TC_TN$site, levels = c("PT", "BEO", "FL"))

ggplot(TC_TN, aes(x = site, y = C, fill = site)) + 
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +  # Remove outliers for clarity
  geom_jitter(width = 0.2, alpha = 0.6, size = 2) + # Add jittered points
  scale_fill_manual(values = c("#1b9e77", "#7570b3", "#d95f02")) +  # Custom colors
  theme_classic() +
  scale_y_continuous(limits = c(0,47)) +
  labs(x = "Site", 
       y = "% C") +
  theme(
    text = element_text(size = 12),  
    axis.text = element_text(size = 12, color = "black"),
    legend.position = "none") +  # Remove legend if unnecessary
    geom_signif(
    comparisons = list(c("PT", "FL"), c("PT", "BEO")),  # Specify the groups to compare
    map_signif_level = TRUE,  # Automatically display significance codes like "*", "**", etc.
    annotations = c("***", "*"),  # Manually annotate with custom significance labels
    y_position = c(42,43),  # Adjust y-position of the annotations
    tip_length = 0.03,
    textsize = 8)    
```

Boxplot for N
```{r}
ggplot(TC_TN, aes(x = site, y = N, fill = site)) + 
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +  # Remove outliers for clarity
  geom_jitter(width = 0.2, alpha = 0.6, size = 2) + # Add jittered points
  scale_fill_manual(values = c("#1b9e77", "#7570b3", "#d95f02")) +  # Custom colors
  theme_classic() +
  scale_y_continuous(limits = c(0, 2.6)) +
  labs(x = "Site", 
       y = "% N") +
  theme(
    text = element_text(size = 12),  
    axis.text = element_text(size = 12, color = "black"),
    legend.position = "none") +  # Remove legend if unnecessary
    geom_signif(
    comparisons = list(c("PT", "FL"), c("PT", "BEO")),  # Specify the groups to compare
    map_signif_level = TRUE,  # Automatically display significance codes like "*", "**", etc.
    annotations = c("***", "*"),  # Manually annotate with custom significance labels
    y_position = c(2.35,2.4),  # Adjust y-position of the annotations
    tip_length = 0.03,
    textsize = 8)  
```

OLD CODE (C and N linear models):
4. GLM with C and N
```{r}
# Trying GLM since this data is normally distributed 
# C
linearmodel_Cresp <- lm(Cumulative_Respiration ~ C, data = resp_tctn)

print(linearmodel_Cresp)
summary(linearmodel_Cresp)

# Define custom site colors (same as in boxplot)
site_colors <- c("CRREL" = "#1b9e77", "FL" = "#d95f02", "Utqiagvik" = "#7570b3")

ggplot(resp_tctn, aes(x = C, y = Cumulative_Respiration)) + 
  geom_point(aes(color = site), size = 3.5, alpha = 0.7) +  # Adjust transparency to match jitter points
  scale_color_manual(values = site_colors, name = "Site") +  # Custom site colors
  theme_classic() +
  labs(
    x = "% C", 
    y = "Cumulative Respiration (µg C-CO2 g-1 dry soil)"
  ) +
  geom_smooth(method = "lm", color = "dark gray", se = TRUE) +  # Regression line
  theme(
    text = element_text(size = 16),  
    axis.text = element_text(size = 12, color = "black"),
    #axis.title = element_text(size = 12, face = "bold"),
    #plot.title = element_text(size = 12, face = "bold", hjust = 0.5)
  )

```

```{r}
# N
linearmodel_Nresp <- lm(Cumulative_Respiration ~ N, data = resp_tctn)
summary(linearmodel_Nresp)
print(linearmodel_Nresp)


shapiro.test(resp_tctn$N)
Nanova <- aov(N ~ Cumulative_Respiration, data = resp_tctn)
summary(Nanova)


# Define custom site colors (same as in boxplot)
site_colors <- c("CRREL" = "#1b9e77", "FL" = "#d95f02", "Utqiagvik" = "#7570b3")

ggplot(resp_tctn, aes(x = N, y = Cumulative_Respiration)) + 
  geom_point(aes(color = site), size = 3.5, alpha = 0.7) +  # Adjust transparency to match jitter points
  scale_color_manual(values = site_colors, name = "Site") +  # Custom site colors
  theme_classic() +
  labs(
    x = "% N", 
    y = "Cumulative Respiration (µg C-CO2 g-1 dry soil)"
  ) +
  geom_smooth(method = "lm", color = "dark gray", se = TRUE) +  # Regression line
  theme(
    text = element_text(size = 16),  
    axis.text = element_text(size = 12, color = "black"),
    #axis.title = element_text(size = 12, face = "bold"),
    #plot.title = element_text(size = 12, face = "bold", hjust = 0.5)
  )
```

```{r}
# Checking the relationship using all of the replicates (Jessica asked to make sure that this relationship is strong)
respcheck <- read_excel("~/Desktop/incubation_physical_chemical.xlsx", sheet = "Resp_TC_TN_check")

shapiro.test(respcheck$Cumulative_Respiration) # not normally distributed
resp_C_corr_test <- cor.test(x =respcheck$C, y = respcheck$Cumulative_Respiration, method = "spearman")
print(resp_C_corr_test)


ggplot(respcheck, aes(x = C, y = Cumulative_Respiration)) + 
  geom_point(aes(color = site), size = 2.5) +
  theme_classic() +
  xlab("% C") +
  ylab("Cumulative Respiration (µg C-CO2 g-1 dry soil)") +
  geom_smooth(method = "lm", color = "dark gray", se = TRUE) +
  theme(text = element_text(size = 12)) +
  scale_color_discrete("Site") #changing the name of the legend title

linearmodel_Crespcheck <- lm(Cumulative_Respiration ~ C, data = respcheck)
print(linearmodel_Crespcheck)
summary(linearmodel_Crespcheck)

"#4169e1"
"#6a7bd7"
```