---
title: "3. Microbial diversity analysis"
author: "Joy O'Brien"
date: "2025-03-19"
output: html_document
---
Previously micro_stats.R (April 29, 2022)
# This script encompassess statistical tests that were performed on the phyloseq object

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("/Users/joyobrien/GitHub/thaw.response")
```

```{r}
# Load the necessary packages (this is from the phyloseq tutorial)
library("phyloseq")
library("ggplot2")      # graphics
library("readxl")       # necessary to import the data from Excel file
library("dplyr")        # filter and reformat data frames
library("tibble")       # Needed for converting column to row names
library("vegan")
library("viridis")
library("RColorBrewer")
library("colorRamps")
library("colorspace")
library("ggpubr")
library("readxl")
library("FSA")
library("agricolae") # used to get significance codes 
```

```{r}
# Load in the phyloseq object
thaw <- readRDS("~/GitHub/thaw.response/Data/thaw_phyloseq.rds")
```
0. Looking at top phylum and classes for fun
```{r}

# TOP 5 PHYLA
# Load necessary libraries
library(phyloseq)
library(ggplot2)

# Summarize taxa at the Phylum level
phyla_abundance <- tax_glom(thaw, taxrank = "Phylum")

# Convert phyloseq object to a dataframe
phyla_df <- as.data.frame(tax_table(phyla_abundance))
phyla_df$Abundance <- taxa_sums(phyla_abundance)

# Summarize and sort by abundance
phyla_df <- phyla_df[order(-phyla_df$Abundance), ]  # Sort in descending order
top_5_phyla <- phyla_df[1:5, ]  # Select top 5 phyla

# Plot using ggplot2
ggplot(top_5_phyla, aes(x = reorder(Phylum, -Abundance), y = Abundance, fill = Phylum)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Top 5 Most Abundant Phyla",
       x = "Phylum",
       y = "Abundance") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
```{r}
# TOP 10 PHYLA

top_10_phyla <- phyla_df[1:10, ]  # Select top 10 phyla

# Plot using ggplot2
ggplot(top_10_phyla, aes(x = reorder(Phylum, -Abundance), y = Abundance, fill = Phylum)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Top 10 Most Abundant Phyla",
       x = "Phylum",
       y = "Abundance") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

0. Looking at the top phyla and classes for fun
```{r}
# TOP 5 CLASSES
# Load necessary libraries
library(phyloseq)
library(ggplot2)

# Summarize taxa at the Phylum level
class_abundance <- tax_glom(thaw, taxrank = "Class")

# Convert phyloseq object to a dataframe
class_df <- as.data.frame(tax_table(class_abundance))
class_df$Abundance <- taxa_sums(class_abundance)

# Summarize and sort by abundance
class_df <- class_df[order(-class_df$Abundance), ]  # Sort in descending order
top_5_class <- class_df[1:5, ]  # Select top 5 classes

# Plot using ggplot2
ggplot(top_5_class, aes(x = reorder(Class, -Abundance), y = Abundance, fill = Class)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Top 5 Most Abundant Class",
       x = "Class",
       y = "Abundance") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r}
# TOP 10 MOST ABUNDANT CLASS
top_10_class <- class_df[1:10, ]  # Select top 10 classes

# Plot using ggplot2
ggplot(top_10_class, aes(x = reorder(Class, -Abundance), y = Abundance, fill = Class)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Top 10 Most Abundant Class",
       x = "Class",
       y = "Abundance") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


1. Coming back to this after thaw response analysis--need to know if the common thaw responders make up a significant part of the pre thaw? 
1A. Top 5 phyla pre-thaw 
```{r}
# Okay we made the thaw.abs phyloseq object in script 1, so that we can look at absolute abundance

# TOP 5 PHYLA pre-thaw
presamples <- subset_samples(thaw.abs, pre_post_thaw == "pre")
# Summarize taxa at the Phylum level
prephyla_abundance <- tax_glom(presamples, taxrank = "Phylum")

# Convert phyloseq object to a dataframe
prephyla_df <- as.data.frame(tax_table(prephyla_abundance))
prephyla_df$Abundance <- taxa_sums(prephyla_abundance)

# Summarize and sort by abundance
prephyla_df <- prephyla_df[order(-prephyla_df$Abundance), ]  # Sort in descending order
pretop_5_phyla <- prephyla_df[1:5, ]  # Select top 5 phyla

# Plot using ggplot2
ggplot(pretop_5_phyla, aes(x = reorder(Phylum, -Abundance), y = Abundance, fill = Phylum)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Top 5 Most Abundant Phyla PRE-THAW",
       x = "Phylum",
       y = "Abundance") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


1B. Top 5 phyla post thaw 
```{r}
postsamples <- subset_samples(thaw.abs, pre_post_thaw == "post")
# Summarize taxa at the Phylum level
postphyla_abundance <- tax_glom(postsamples, taxrank = "Phylum")

# Convert phyloseq object to a dataframe
postphyla_df <- as.data.frame(tax_table(postphyla_abundance))
postphyla_df$Abundance <- taxa_sums(postphyla_abundance)

# Summarize and sort by abundance
postphyla_df <- postphyla_df[order(-postphyla_df$Abundance), ]  # Sort in descending order
posttop_5_phyla <- postphyla_df[1:5, ]  # Select top 5 phyla

# Plot using ggplot2
ggplot(posttop_5_phyla, aes(x = reorder(Phylum, -Abundance), y = Abundance, fill = Phylum)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Top 5 Most Abundant Phyla POST-THAW",
       x = "Phylum",
       y = "Abundance") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```



1C. Top 5 genus pre-thaw 

```{r}
# TOP 5 PHYLA pre-thaw
presamples <- subset_samples(thaw.abs, pre_post_thaw == "pre")
# Summarize taxa at the Phylum level
pregenus_abundance <- tax_glom(presamples, taxrank = "Genus")

# Convert phyloseq object to a dataframe
pregenus_df <- as.data.frame(tax_table(pregenus_abundance))
pregenus_df$Abundance <- taxa_sums(pregenus_abundance)

# Summarize and sort by abundance
pregenus_df <- pregenus_df[order(-pregenus_df$Abundance), ]  # Sort in descending order
pretop_5_genus <- pregenus_df[1:5, ]  # Select top 5 genus

# Plot using ggplot2
ggplot(pretop_5_genus, aes(x = reorder(Genus, -Abundance), y = Abundance, fill = Genus)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Top 5 Most Abundant Genus PRE-THAW",
       x = "Genus",
       y = "Abundance") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


1D. Top 5 genus post-thaw

```{r}
# TOP 5 PHYLA pre-thaw
postamples <- subset_samples(thaw.abs, pre_post_thaw == "post")
# Summarize taxa at the Phylum level
postgenus_abundance <- tax_glom(postsamples, taxrank = "Genus")

# Convert phyloseq object to a dataframe
postgenus_df <- as.data.frame(tax_table(postgenus_abundance))
postgenus_df$Abundance <- taxa_sums(postgenus_abundance)

# Summarize and sort by abundance
postgenus_df <- postgenus_df[order(-postgenus_df$Abundance), ]  # Sort in descending order
posttop_5_genus <- postgenus_df[1:5, ]  # Select top 5 genus

# Plot using ggplot2
ggplot(posttop_5_genus, aes(x = reorder(Genus, -Abundance), y = Abundance, fill = Genus)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Top 5 Most Abundant Genus POST-THAW",
       x = "Genus",
       y = "Abundance") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Should this be assessbed by site now?
CRREL
```{r}
# TOP 5 Genus pre-thaw -- CRREL
presamples_crrel <- subset_samples(thaw.abs, pre_post_thaw == "pre" & site == "CRREL")
# Summarize taxa at the Phylum level
pregenus_abundance_crrel <- tax_glom(presamples_crrel, taxrank = "Genus")

# Convert phyloseq object to a dataframe
pregenus_df_crrel <- as.data.frame(tax_table(pregenus_abundance_crrel))
pregenus_df_crrel$Abundance <- taxa_sums(pregenus_abundance_crrel)

# Summarize and sort by abundance
pregenus_df_crrel <- pregenus_df_crrel[order(-pregenus_df_crrel$Abundance), ]  # Sort in descending order
pretop_5_genus_crrel <- pregenus_df_crrel[1:10, ]  # Select top 5 genus/top 10

# Plot using ggplot2
ggplot(pretop_5_genus_crrel, aes(x = reorder(Genus, -Abundance), y = Abundance, fill = Genus)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Top 10 Most Abundant Genus PRE-THAW CRREL",
       x = "Genus",
       y = "Abundance") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


# TOP 5 Genus post-thaw -- CRREL
postsamples_crrel <- subset_samples(thaw.abs, pre_post_thaw == "post" & site == "CRREL")
# Summarize taxa at the Phylum level
postgenus_abundance_crrel <- tax_glom(postsamples_crrel, taxrank = "Genus")

# Convert phyloseq object to a dataframe
postgenus_df_crrel <- as.data.frame(tax_table(postgenus_abundance_crrel))
postgenus_df_crrel$Abundance <- taxa_sums(postgenus_abundance_crrel)

# Summarize and sort by abundance
postgenus_df_crrel <- postgenus_df_crrel[order(-postgenus_df_crrel$Abundance), ]  # Sort in descending order
posttop_5_genus_crrel <- postgenus_df_crrel[1:10, ]  # Select top 5 genus

# Plot using ggplot2
ggplot(posttop_5_genus_crrel, aes(x = reorder(Genus, -Abundance), y = Abundance, fill = Genus)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Top 10 Most Abundant Genus POST-THAW",
       x = "Genus",
       y = "Abundance") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

CRREL top 10, phylum legend

```{r}
# Subset and summarize
#presamples_crrel <- subset_samples(thaw.abs, pre_post_thaw == "pre" & site == "CRREL")#Epregenus_abundance_crrel <- tax_glom(presamples_crrel, taxrank = "Genus")

# Convert phyloseq object to dataframe
pregenus_df_crrel <- as.data.frame(tax_table(pregenus_abundance_crrel))
pregenus_df_crrel$Abundance <- taxa_sums(pregenus_abundance_crrel)
#pregenus_df_crrel$Genus <- rownames(pregenus_df_crrel)

# Select top 10
pregenus_df_crrel <- pregenus_df_crrel[order(-pregenus_df_crrel$Abundance), ]
pretop_10_genus_crrel <- pregenus_df_crrel[1:10, ]

# Plot
ggplot(pretop_10_genus_crrel, aes(x = reorder(Genus, -Abundance), y = Abundance, fill = Phylum)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Top 10 Most Abundant Genus PRE-THAW CRREL",
       x = "Genus",
       y = "Abundance",
       fill = "Phylum") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

POST THAW CRREL
```{r}
# Subset and summarize
#presamples_crrel <- subset_samples(thaw.abs, pre_post_thaw == "pre" & site == "CRREL")#Epregenus_abundance_crrel <- tax_glom(presamples_crrel, taxrank = "Genus")

# Convert phyloseq object to dataframe
pregenus_df_crrel <- as.data.frame(tax_table(postgenus_abundance_crrel))
pregenus_df_crrel$Abundance <- taxa_sums(postgenus_abundance_crrel)
#pregenus_df_crrel$Genus <- rownames(pregenus_df_crrel)

# Select top 10
postgenus_df_crrel <- postgenus_df_crrel[order(-postgenus_df_crrel$Abundance), ]
posttop_10_genus_crrel <- postgenus_df_crrel[1:10, ]

# Plot
ggplot(posttop_10_genus_crrel, aes(x = reorder(Genus, -Abundance), y = Abundance, fill = Phylum)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Top 10 Most Abundant Genus POST-THAW CRREL",
       x = "Genus",
       y = "Abundance",
       fill = "Phylum") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


FL

```{r}
# Subset and summarize
# presamples_fl <- subset_samples(thaw.abs, pre_post_thaw == "pre" & site == "FL")
# pregenus_abundance_fl <- tax_glom(presamples_fl, taxrank = "Genus")

# Convert phyloseq object to dataframe
pregenus_df_fl <- as.data.frame(tax_table(pregenus_abundance_fl))
pregenus_df_fl$Abundance <- taxa_sums(pregenus_abundance_fl)
# pregenus_df_fl$Genus <- rownames(pregenus_df_fl)

# Select top 10
pregenus_df_fl <- pregenus_df_fl[order(-pregenus_df_fl$Abundance), ]
pretop_10_genus_fl <- pregenus_df_fl[1:10, ]

# Plot
ggplot(pretop_10_genus_fl, aes(x = reorder(Genus, -Abundance), y = Abundance, fill = Phylum)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Top 10 Most Abundant Genus PRE-THAW FL",
       x = "Genus",
       y = "Abundance",
       fill = "Phylum") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Convert phyloseq object to dataframe
postgenus_df_fl <- as.data.frame(tax_table(postgenus_abundance_fl))
postgenus_df_fl$Abundance <- taxa_sums(postgenus_abundance_fl)
# postgenus_df_fl$Genus <- rownames(postgenus_df_fl)

# Select top 10
postgenus_df_fl <- postgenus_df_fl[order(-postgenus_df_fl$Abundance), ]
posttop_10_genus_fl <- postgenus_df_fl[1:10, ]

# Plot
ggplot(posttop_10_genus_fl, aes(x = reorder(Genus, -Abundance), y = Abundance, fill = Phylum)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Top 10 Most Abundant Genus POST-THAW FL",
       x = "Genus",
       y = "Abundance",
       fill = "Phylum") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))



```

```{r}
# TOP 5 Genus pre-thaw -- FL
presamples_fl <- subset_samples(thaw.abs, pre_post_thaw == "pre" & site == "FL")
# Summarize taxa at the Phylum level
pregenus_abundance_fl <- tax_glom(presamples_fl, taxrank = "Genus")

# Convert phyloseq object to a dataframe
pregenus_df_fl <- as.data.frame(tax_table(pregenus_abundance_fl))
pregenus_df_fl$Abundance <- taxa_sums(pregenus_abundance_fl)

# Summarize and sort by abundance
pregenus_df_fl <- pregenus_df_fl[order(-pregenus_df_fl$Abundance), ]  # Sort in descending order
pretop_5_genus_fl <- pregenus_df_fl[1:5, ]  # Select top 5 genus

# Plot using ggplot2
ggplot(pretop_5_genus_fl, aes(x = reorder(Genus, -Abundance), y = Abundance, fill = Genus)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Top 5 Most Abundant Genus PRE-THAW FL",
       x = "Genus",
       y = "Abundance") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


# TOP 5 Genus post-thaw -- FL
postsamples_fl <- subset_samples(thaw.abs, pre_post_thaw == "post" & site == "FL")
# Summarize taxa at the Phylum level
postgenus_abundance_fl <- tax_glom(postsamples_fl, taxrank = "Genus")

# Convert phyloseq object to a dataframe
postgenus_df_fl <- as.data.frame(tax_table(postgenus_abundance_fl))
postgenus_df_fl$Abundance <- taxa_sums(postgenus_abundance_fl)

# Summarize and sort by abundance
postgenus_df_fl <- postgenus_df_fl[order(-postgenus_df_fl$Abundance), ]  # Sort in descending order
posttop_5_genus_fl <- postgenus_df_fl[1:5, ]  # Select top 5 genus

# Plot using ggplot2
ggplot(posttop_5_genus_fl, aes(x = reorder(Genus, -Abundance), y = Abundance, fill = Genus)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Top 5 Most Abundant Genus POST-THAW FL",
       x = "Genus",
       y = "Abundance") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

UT

```{r}
# Subset and summarize (if needed)
# presamples_ut <- subset_samples(thaw.abs, pre_post_thaw == "pre" & site == "UT")
# pregenus_abundance_ut <- tax_glom(presamples_ut, taxrank = "Genus")

# Convert phyloseq object to dataframe
pregenus_df_ut <- as.data.frame(tax_table(pregenus_abundance_ut))
pregenus_df_ut$Abundance <- taxa_sums(pregenus_abundance_ut)
# pregenus_df_ut$Genus <- rownames(pregenus_df_ut)  # Uncomment if genus is in rownames

# Select top 10 genera
pregenus_df_ut <- pregenus_df_ut[order(-pregenus_df_ut$Abundance), ]
pretop_10_genus_ut <- pregenus_df_ut[1:10, ]

# Plot
ggplot(pretop_10_genus_ut, aes(x = reorder(Genus, -Abundance), y = Abundance, fill = Phylum)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Top 10 Most Abundant Genus PRE-THAW UT",
       x = "Genus",
       y = "Abundance",
       fill = "Phylum") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))





# Subset and summarize (if needed)
# postsamples_ut <- subset_samples(thaw.abs, pre_post_thaw == "post" & site == "UT")
# postgenus_abundance_ut <- tax_glom(postsamples_ut, taxrank = "Genus")

# Convert phyloseq object to dataframe
postgenus_df_ut <- as.data.frame(tax_table(postgenus_abundance_ut))
postgenus_df_ut$Abundance <- taxa_sums(postgenus_abundance_ut)
# postgenus_df_ut$Genus <- rownames(postgenus_df_ut)  # Uncomment if genus is in rownames

# Select top 10 genera
postgenus_df_ut <- postgenus_df_ut[order(-postgenus_df_ut$Abundance), ]
posttop_10_genus_ut <- postgenus_df_ut[1:10, ]

# Plot
ggplot(posttop_10_genus_ut, aes(x = reorder(Genus, -Abundance), y = Abundance, fill = Phylum)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Top 10 Most Abundant Genus POST-THAW UT",
       x = "Genus",
       y = "Abundance",
       fill = "Phylum") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```
Facet by site -- UT
Okay now we need to clean up (color by phylum across all figures, facet by site, 
```{r}

# Add timepoint column to each dataframe
pretop_10_genus_ut$Timepoint <- "Pre-Thaw"
posttop_10_genus_ut$Timepoint <- "Post-Thaw"


#combined_ut$Timepoint <- factor(combined_ut$Timepoint, levels = c("Pre-Thaw", "Post-Thaw"))


# Combine into one dataframe
combined_ut <- rbind(pretop_10_genus_ut, posttop_10_genus_ut)
combined_ut$Timepoint <- factor(as.character(combined_ut$Timepoint), levels = c("Pre-Thaw", "Post-Thaw"))
# Make sure Genus is a factor and levels are consistent
combined_ut$Genus <- factor(combined_ut$Genus, levels = unique(combined_ut$Genus))

# Plot
ggplot(combined_ut, aes(x = reorder(Genus, -Abundance), y = Abundance, fill = Phylum)) +
  geom_bar(stat = "identity") +
  facet_wrap(~Timepoint, scales = "free_x") +
  scale_y_log10() +
  theme_minimal() +
  labs(title = "Top 10 Most Abundant Genus - UT",
       x = "Genus",
       y = "Abundance",
       fill = "Phylum") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Facet by site --FL
```{r}
# Add timepoint column to each dataframe
pretop_10_genus_fl$Timepoint <- "Pre-Thaw"
posttop_10_genus_fl$Timepoint <- "Post-Thaw"

# Combine into one dataframe
combined_fl <- rbind(pretop_10_genus_fl, posttop_10_genus_fl)
combined_fl$Timepoint <- factor(as.character(combined_fl$Timepoint), levels = c("Pre-Thaw", "Post-Thaw"))

# Make sure Genus is a factor and levels are consistent
combined_fl$Genus <- factor(combined_fl$Genus, levels = unique(combined_fl$Genus))

# Plot
ggplot(combined_fl, aes(x = reorder(Genus, -Abundance), y = Abundance, fill = Phylum)) +
  geom_bar(stat = "identity") +
  facet_wrap(~Timepoint, scales = "free_x") +
  scale_y_log10() +
  theme_minimal() +
  labs(title = "Top 10 Most Abundant Genus - FL",
       x = "Genus",
       y = "Abundance",
       fill = "Phylum") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```
Facet by site--CRREL
```{r}
# Add timepoint column to each dataframe
pretop_10_genus_crrel$Timepoint <- "Pre-Thaw"
posttop_10_genus_crrel$Timepoint <- "Post-Thaw"

# Combine into one dataframe
combined_crrel <- rbind(pretop_10_genus_crrel, posttop_10_genus_crrel)
combined_crrel$Timepoint <- factor(as.character(combined_crrel$Timepoint), levels = c("Pre-Thaw", "Post-Thaw"))

# Make sure Genus is a factor and levels are consistent
combined_crrel$Genus <- factor(combined_crrel$Genus, levels = unique(combined_crrel$Genus))

# Plot
ggplot(combined_crrel, aes(x = reorder(Genus, -Abundance), y = Abundance, fill = Phylum)) +
  geom_bar(stat = "identity") +
  facet_wrap(~Timepoint, scales = "free_x") +
  scale_y_log10() +
  theme_minimal() +
  labs(title = "Top 10 Most Abundant Genus - CRREL",
       x = "Genus",
       y = "Abundance",
       fill = "Phylum") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```



```{r}
# TOP 5 Genus pre-thaw -- UT
presamples_ut <- subset_samples(thaw.abs, pre_post_thaw == "pre" & site == "Utqiagvik")
# Summarize taxa at the Phylum level
pregenus_abundance_ut <- tax_glom(presamples_ut, taxrank = "Genus")

# Convert phyloseq object to a dataframe
pregenus_df_ut <- as.data.frame(tax_table(pregenus_abundance_ut))
pregenus_df_ut$Abundance <- taxa_sums(pregenus_abundance_ut)

# Summarize and sort by abundance
pregenus_df_ut <- pregenus_df_ut[order(-pregenus_df_ut$Abundance), ]  # Sort in descending order
pretop_5_genus_ut <- pregenus_df_ut[1:5, ]  # Select top 5 genus

# Plot using ggplot2
ggplot(pretop_5_genus_ut, aes(x = reorder(Genus, -Abundance), y = Abundance, fill = Genus)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Top 5 Most Abundant Genus PRE-THAW UT",
       x = "Genus",
       y = "Abundance") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


# TOP 5 Genus post-thaw -- UT
postsamples_ut <- subset_samples(thaw.abs, pre_post_thaw == "post" & site == "Utqiagvik")
# Summarize taxa at the Phylum level
postgenus_abundance_ut <- tax_glom(postsamples_ut, taxrank = "Genus")

# Convert phyloseq object to a dataframe
postgenus_df_ut <- as.data.frame(tax_table(postgenus_abundance_ut))
postgenus_df_ut$Abundance <- taxa_sums(postgenus_abundance_ut)

# Summarize and sort by abundance
postgenus_df_ut <- postgenus_df_ut[order(-postgenus_df_ut$Abundance), ]  # Sort in descending order
posttop_5_genus_ut <- postgenus_df_ut[1:5, ]  # Select top 5 genus

# Plot using ggplot2
ggplot(posttop_5_genus_ut, aes(x = reorder(Genus, -Abundance), y = Abundance, fill = Genus)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Top 5 Most Abundant Genus POST-THAW UT",
       x = "Genus",
       y = "Abundance") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
Phyla as legend
```{r}
# Merge tax_table info and abundance into one dataframe
pregenus_df_ut$Genus <- rownames(pregenus_df_ut)  # Keep Genus names
pregenus_df_ut$Phylum <- pregenus_df_ut[, "Phylum"]  # Extract Phylum
pregenus_df_ut$Genus <- factor(pregenus_df_ut$Genus, levels = pregenus_df_ut$Genus)

# Plot: Genus on x-axis, Phylum as fill
ggplot(pretop_5_genus_ut, aes(x = reorder(Genus, -Abundance), y = Abundance, fill = Phylum)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Top 5 Most Abundant Genus PRE-THAW UT",
       x = "Genus",
       y = "Abundance",
       fill = "Phylum") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))



# Add Genus and Phylum info
postgenus_df_ut$Genus <- rownames(postgenus_df_ut)
postgenus_df_ut$Phylum <- postgenus_df_ut[, "Phylum"]
postgenus_df_ut$Genus <- factor(postgenus_df_ut$Genus, levels = postgenus_df_ut$Genus)

# Plot
ggplot(posttop_5_genus_ut, aes(x = reorder(Genus, -Abundance), y = Abundance, fill = Phylum)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Top 5 Most Abundant Genus POST-THAW UT",
       x = "Genus",
       y = "Abundance",
       fill = "Phylum") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```



Top 5 Phylum pre-thaw 
```{r}
# TOP 5 PHYLA pre-thaw
presamples <- subset_samples(thaw, pre_post_thaw == "pre")
# Summarize taxa at the Phylum level
prephyla_abundance <- tax_glom(presamples, taxrank = "Phylum")

# Convert phyloseq object to a dataframe
prephyla_df <- as.data.frame(tax_table(prephyla_abundance))
prephyla_df$Abundance <- taxa_sums(prephyla_abundance)

# Summarize and sort by abundance
prephyla_df <- prephyla_df[order(-prephyla_df$Abundance), ]  # Sort in descending order
pretop_5_phyla <- prephyla_df[1:5, ]  # Select top 5 phyla

# Plot using ggplot2
ggplot(pretop_5_phyla, aes(x = reorder(Phylum, -Abundance), y = Abundance, fill = Phylum)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Top 5 Most Abundant Phyla PRE-THAW",
       x = "Phylum",
       y = "Abundance") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Top 5 phylum post-thaw
```{r}
# TOP 5 PHYLA pre-thaw
postsamples <- subset_samples(thaw, pre_post_thaw == "post")
# Summarize taxa at the Phylum level
postphyla_abundance <- tax_glom(postsamples, taxrank = "Phylum")

# Convert phyloseq object to a dataframe
postphyla_df <- as.data.frame(tax_table(postphyla_abundance))
postphyla_df$Abundance <- taxa_sums(postphyla_abundance)

# Summarize and sort by abundance
postphyla_df <- postphyla_df[order(-postphyla_df$Abundance), ]  # Sort in descending order
posttop_5_phyla <- postphyla_df[1:5, ]  # Select top 5 phyla

# Plot using ggplot2
ggplot(posttop_5_phyla, aes(x = reorder(Phylum, -Abundance), y = Abundance, fill = Phylum)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Top 5 Most Abundant Phyla POST-THAW",
       x = "Phylum",
       y = "Abundance") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

1. Shannon and Simpson Diversity metrics
```{r}
# Creating a table of Simpson and Shannon diversity
shansimp <- estimate_richness(thaw, measures = c("Simpson", "Shannon"))

# Visualize shannon
plot_richness(thaw, x = "site", color = "pre_post_thaw", measures = c("Shannon")) +
  geom_boxplot() +
  theme_classic() +
  xlab("Site") +
  theme(text = element_text(size =12)) +
  scale_color_discrete("Treatment")

# Visualize Simpson
plot_richness(thaw, x = "site", color = "pre_post_thaw", measures = c("Simpson")) +
  geom_boxplot() +
  theme_classic() +
  xlab("Site") +
  theme(text = element_text(size =12)) +
  scale_color_discrete("Treatment")

  
```


A. Obtain Shannon and Simpson metrics
```{r}
# Creating a table of Simpson and Shannon diversity and run stats
shan.simp <- as.data.frame(shansimp)

# Add metadata to this dataframe
metadata_df <- as.data.frame(sample_data(thaw))
shansimp.meta <- merge(metadata_df, shan.simp, by = "row.names", all.x = TRUE)

# Rename the first column after the merge
#colnames(shansimp.meta)[1] <- "sample_ID"

# View the merged dataframe
head(shansimp.meta)

# Save the table as a CSV file
#write.csv(shansimp.meta, "~/GitHub/thaw.response/Data/thaw_diversity_metrics.csv", row.names = FALSE)

```
B. Check for normality of the metrics
```{r}
# Before running stats, check to see if data is normally distributed
# Checking normal distribution: SHANNON
hist(shansimp.meta$Shannon, main = "Shannon index", xlab = "")
shapiro.test(shansimp.meta$Shannon) # not normally distributed W = 0.89436, p-value = 0.0008473

# Checking normal distribution: SIMPSON
hist(shansimp.meta$Simpson, main = "Simpson index", xlab = "")
shapiro.test(shansimp.meta$Simpson) # not normally distributed W = 0.57395, p-value = 5.77e-10
```

C. Testing for differences in Shannon and Simpson by site
```{r}
# Check to see if Shannon differs significantly between sites regardless of treatment
# Recall that Shannon is not normally distributed
kruskal.test(shansimp.meta$Shannon ~ site, data = shansimp.meta)
dunnTest(shansimp.meta$Shannon ~ site, data = shansimp.meta)

# Check to see if Simpson differs significantly between sites
kruskal.test(shansimp.meta$Simpson ~ site, data = shansimp.meta)
dunnTest(shansimp.meta$Simpson ~ site, data = shansimp.meta)

# Check to see if it (Shannon) differs based on pre post thaw # use wilcoxon because we are comparing two groups
wilcox.test(Shannon ~ pre_post_thaw, data = shansimp.meta)

wilcox.test(Simpson ~ pre_post_thaw, data = shansimp.meta)

# To plot with significance codes?
kruskal(ssmeta$Shannon, ssmeta$site, group=TRUE, p.adj="bonferroni")$groups #using this to get the letter significance codes 
dunnTest(ssmeta$Shannon ~ site, data = ssmeta)


by(shansimp.meta$Shannon, shansimp.meta$Site, 
   function(x) wilcox.test(x ~ shansimp.meta$pre_post_thaw[shansimp.meta$Site == unique(shansimp.meta$Site)]))

```


D. Min and Max values for Shannon and Simpson
```{r}
# Load dplyr for grouping and summarizing the data
library(dplyr)

# Calculate min and max Shannon by treatment
shannon_by_treatment <- shansimp.meta %>%
  group_by(pre_post_thaw) %>%
  summarise(
    min_shannon = min(Shannon),
    max_shannon = max(Shannon)
  )

# Calculate min and max Simpson by treatment
simpson_by_treatment <- shansimp.meta %>%
  group_by(pre_post_thaw) %>%
  summarise(
    min_simpson = min(Simpson),
    max_simpson = max(Simpson)
  )

# Print the results
print(shannon_by_treatment)
print(simpson_by_treatment)
```

E. Mean and SD for Shannon and Simpson
```{r}
# Obtain average Shannon values for each site with SD
# Load dplyr for data manipulation
library(dplyr)

# Calculate min, max, and standard error for Shannon by site
shannon_by_site <- shansimp.meta %>%
  group_by(site) %>%
  summarise(
    avg_shannon = mean(Shannon),
    se_shannon = sd(Shannon) / sqrt(n())
  )

# Calculate min, max, and standard error for Simpson by site
simpson_by_site <- shansimp.meta %>%
  group_by(site) %>%
  summarise(
    avg_simpson = mean(Simpson),
    se_simpson = sd(Simpson) / sqrt(n())
  )

# Print the results
print(shannon_by_site)
print(simpson_by_site)

```

2. Shannon and Simpson by site 
A. Subset Shannon and Simpson by site
```{r}
# Subset data by site and check for normality
sscrrel <- subset(shansimp.meta, site =="CRREL")
ssfl <- subset(shansimp.meta, site =="FL")
ssut <- subset(shansimp.meta, site =="Utqiagvik")

# Testing normality for each site: SHANNON
shapiro.test(sscrrel$Shannon) # p = 0.9954 normally distributed, use anova 
shapiro.test(ssfl$Shannon) # p = 0.02876 not normally distributed, use kruskal
shapiro.test(ssut$Shannon) # p = 0.001398 not normally distributed, use kruskal 

# Testing normality for each site: SIMPSON
shapiro.test(sscrrel$Simpson) # p = 0.03035 # use kruskal to evaluate 
shapiro.test(ssfl$Simpson) # p = 0.00119 # use kruskal
shapiro.test(ssut$Simpson) # p = 3.407e-05 not normally distributed, use kruskal
```

B. Test for differences between pre and post thaw samples from the same site SHANNON
```{r}
# ANOVA for CRREL SHANNON
crrelanova <- aov(Shannon ~ pre_post_thaw, data = sscrrel)
print(crrelanova)
summary(crrelanova)

# Estimated effects may be unbalanced
table(sscrrel$pre_post_thaw)
# run test for homogeneity of variance
library(car)
leveneTest(Shannon ~ pre_post_thaw, data = sscrrel) #indicates that homogenetiy of variances is met and we can proceed with anova

# Run shapiro to check normality of residuals
shapiro.test(resid(crrelanova))  # Shapiro-Wilk test, p-value indicates that residuals are normal

# Visualize
qqnorm(resid(crrelanova))
qqline(resid(crrelanova))
hist(resid(crrelanova), main = "Histogram of Residuals", breaks = 10)

summary(crrelanova)
TukeyHSD(crrelanova)

# Obtain significance codes for CRREL Shannon
# Extract p-value from ANOVA summary
crrel_anova_summary <- summary(crrelanova)
crrel_shannon_p_value <- crrel_anova_summary[[1]]$`Pr(>F)`[1]  # Extract p-value

```

```{r}
# Running Kruskal wallis on the rest of the sites since their Shannon was not normally distributed

# Farmers Loop shannon
kruskal.fl <- kruskal.test(ssfl$Shannon ~ pre_post_thaw, data = ssfl) #Kruskal-Wallis chi-squared = 2.0408, df = 1, p-value = 0.1531
print(kruskal.fl)
# Obtain significance codes for FL Shannon
# Extract p-value from the Kruskal-Wallis test result
fl_shannon_p_value <- kruskal.fl$p.value
```


```{r}
# Utqiagvik shannon
kruskal.ut<- kruskal.test(ssut$Shannon ~ pre_post_thaw, data = ssut) #Kruskal-Wallis chi-squared = 8.6471, df = 1, p-value = 0.003276
# Extract p-value:
ut_shannon_p_value <- kruskal.ut$p.value
```

C. Testing for differences pre-post thaw Simpson by site
```{r}
# CRREL Simpson
kruskal.simp.crrel <- kruskal.test(sscrrel$Simpson ~ pre_post_thaw, data = sscrrel) # Kruskal-Wallis chi-squared = 8.8167, df = 1, p-value = 0.002985
# obtain p-value
crrel_simp_p_value <- kruskal.simp.crrel$p.value

# Farmers Loop Simpson
kruskal.simp.fl<- kruskal.test(ssfl$Simpson ~ pre_post_thaw, data = ssfl) # Kruskal-Wallis chi-squared = 5.898, df = 1, p-value = 0.01516
# obtain p-value
fl_simp_p_value <- kruskal.simp.fl$p.value

# Utqiagvik shannon
kruskal.simp.ut <- kruskal.test(ssut$Simpson ~ pre_post_thaw, data = ssut) # Kruskal-Wallis chi-squared = 7.4559, df = 1, p-value = 0.006323
# obtain p-value
ut_simp_p_value <- kruskal.simp.ut$p.value

```
3. Obtaining average changes in shannon and simpson diversity pre to post thaw with SD
```{r}
# Calculate mean Shannon value for each condition
mean_shannon <- aggregate(Shannon ~ pre_post_thaw, data = sscrrel, mean)

# Compute the difference between conditions
change_in_shannon <- diff(mean_shannon$Shannon)

# Print results
print(mean_shannon)  # Mean values for each condition
print(change_in_shannon)  # Difference in means
```

Shannon Differences-CRREL
```{r}
library(dplyr)

# Calculate mean and standard error (SE) for each condition in CRREL
crrel_summary_shannon <- sscrrel %>%
  group_by(pre_post_thaw) %>%
  summarise(
    mean_shannon = mean(Shannon, na.rm = TRUE),
    se_shannon = sd(Shannon, na.rm = TRUE) / sqrt(n())  # Standard Error (SE)
  )

# Compute difference in means
change_in_shannon <- crrel_summary_shannon$mean_shannon[crrel_summary_shannon$pre_post_thaw == "post"] - 
                     crrel_summary_shannon$mean_shannon[crrel_summary_shannon$pre_post_thaw == "pre"]

# Compute combined standard error (SE) using SE formula for difference of two means
se_change <- sqrt(sum(crrel_summary_shannon$se_shannon^2))

# Print results
print(crrel_summary_shannon)  # Mean and SE for each condition
cat("Average change in Shannon value (Post - Pre):", change_in_shannon, "±", se_change, "\n")

```
Shannon Differences-FL
```{r}
library(dplyr)

# Calculate mean and standard error (SE) for each condition in FL
fl_summary_shannon <- ssfl %>%
  group_by(pre_post_thaw) %>%
  summarise(
    mean_shannon = mean(Shannon, na.rm = TRUE),
    se_shannon = sd(Shannon, na.rm = TRUE) / sqrt(n())  # Standard Error (SE)
  )

# Compute difference in means
fl_change_in_shannon <- fl_summary_shannon$mean_shannon[fl_summary_shannon$pre_post_thaw == "post"] - 
                     fl_summary_shannon$mean_shannon[fl_summary_shannon$pre_post_thaw == "pre"]

# Compute combined standard error (SE) using SE formula for difference of two means
se_change <- sqrt(sum(fl_summary_shannon$se_shannon^2))

# Print results
print(crrel_summary_shannon)  # Mean and SE for each condition
cat("Average change in Shannon value (Post - Pre):", fl_change_in_shannon, "±", se_change, "\n")

```
Shannon Differences - UT
```{r}
library(dplyr)

# Calculate mean and standard error (SE) for each condition in UT
UT_summary_shannon <- ssut %>%
  group_by(pre_post_thaw) %>%
  summarise(
    mean_shannon = mean(Shannon, na.rm = TRUE),
    se_shannon = sd(Shannon, na.rm = TRUE) / sqrt(n())  # Standard Error (SE)
  )

# Compute difference in means
UT_change_in_shannon <- UT_summary_shannon$mean_shannon[UT_summary_shannon$pre_post_thaw == "post"] - 
                     UT_summary_shannon$mean_shannon[UT_summary_shannon$pre_post_thaw == "pre"]

# Compute combined standard error (SE) using SE formula for difference of two means
se_change <- sqrt(sum(crrel_summary_shannon$se_shannon^2))

# Print results
print(UT_summary_shannon)  # Mean and SE for each condition
cat("Average change in Shannon value (Post - Pre):", UT_change_in_shannon, "±", se_change, "\n")
```

Simpson Differences - CRREL
```{r}
library(dplyr)

# Calculate mean and standard error (SE) for each condition in CRREL
crrel_summary_simpson <- sscrrel %>%
  group_by(pre_post_thaw) %>%
  summarise(
    mean_simpson = mean(Simpson, na.rm = TRUE),
    se_simpson = sd(Simpson, na.rm = TRUE) / sqrt(n())  # Standard Error (SE)
  )

# Compute difference in means
crrel_change_in_simpson <- crrel_summary_simpson$mean_simpson[crrel_summary_simpson$pre_post_thaw == "post"] - 
                     crrel_summary_simpson$mean_simpson[crrel_summary_simpson$pre_post_thaw == "pre"]

# Compute combined standard error (SE) using SE formula for difference of two means
se_change <- sqrt(sum(crrel_summary_simpson$se_simpson^2))

# Print results
print(crrel_summary_simpson)  # Mean and SE for each condition
cat("Average change in Simpson value (Post - Pre):", crrel_change_in_simpson, "±", se_change, "\n")


```
Simpson Differences - FL
```{r}
library(dplyr)

# Calculate mean and standard error (SE) for each condition in CRREL
fl_summary_simpson <- ssfl %>%
  group_by(pre_post_thaw) %>%
  summarise(
    mean_simpson = mean(Simpson, na.rm = TRUE),
    se_simpson = sd(Simpson, na.rm = TRUE) / sqrt(n())  # Standard Error (SE)
  )

# Compute difference in means
fl_change_in_simpson <- fl_summary_simpson$mean_simpson[fl_summary_simpson$pre_post_thaw == "post"] - 
                     fl_summary_simpson$mean_simpson[fl_summary_simpson$pre_post_thaw == "pre"]

# Compute combined standard error (SE) using SE formula for difference of two means
se_change <- sqrt(sum(fl_summary_simpson$se_simpson^2))

# Print results
print(fl_summary_simpson)  # Mean and SE for each condition
cat("Average change in Simpson value (Post - Pre):", fl_change_in_simpson, "±", se_change, "\n")

```
Simpson Differences - UT
```{r}
library(dplyr)

# Calculate mean and standard error (SE) for each condition in CRREL
ut_summary_simpson <- ssut %>%
  group_by(pre_post_thaw) %>%
  summarise(
    mean_simpson = mean(Simpson, na.rm = TRUE),
    se_simpson = sd(Simpson, na.rm = TRUE) / sqrt(n())  # Standard Error (SE)
  )

# Compute difference in means
ut_change_in_simpson <- ut_summary_simpson$mean_simpson[ut_summary_simpson$pre_post_thaw == "post"] - 
                     ut_summary_simpson$mean_simpson[ut_summary_simpson$pre_post_thaw == "pre"]

# Compute combined standard error (SE) using SE formula for difference of two means
se_change <- sqrt(sum(ut_summary_simpson$se_simpson^2))

# Print results
print(ut_summary_simpson)  # Mean and SE for each condition
cat("Average change in Simpson value (Post - Pre):", ut_change_in_simpson, "±", se_change, "\n")

```




C1. Plotting data for publication (manuvering around legend re-ordering)--MOVE this later
```{r}
library(dplyr)
library(ggplot2)
# Change the column name of pre-post-thaw to thaw_type
colnames(shansimp.meta)[which(names(shansimp.meta) == "pre_post_thaw")] <- "thaw_type"

```


```{r}
shansimp.fig <- shansimp.meta
# Make a copy of the original dataframe
shansimp.meta1 <- as.data.frame(shansimp.meta)

# Replace 'pre' with 'no_thaw' and 'post' with 'thaw'
#shansimp.meta1$thaw_type <- recode(shansimp.meta1$thaw_type, "pre" = "no_thaw", "post" = "thaw")
shansimp.meta2 <- shansimp.meta1 %>%
  mutate(thaw_type = case_when(
    thaw_type == "pre" ~ "no_thaw",
    thaw_type == "post" ~ "thaw",
    TRUE ~ thaw_type # Keep other values unchanged
  ))
```


```{r}
# Now that this is changed, we can plot 
ggplot(shansimp.meta2, aes(x = site, y = Shannon, color = thaw_type)) +
  geom_boxplot() + 
  theme_classic() +
  xlab("Site") +
  ylab("Alpha Diversity Measure") +
  scale_color_discrete("Treatment", labels=c('Pre-thaw', 'Post-thaw')) +
  theme(text = element_text(size = 12)) 

# Repeat above code with simpson diversity

ggplot(shansimp.meta2, aes(x = site, y = Simpson, color = thaw_type)) +
  geom_boxplot() +
  theme_classic() +
  xlab("Site") +
  ylab("Alpha Diversity Measure") +
  scale_color_discrete("Treatment", labels=c('Pre-thaw', 'Post-thaw')) +
  theme(text = element_text(size =12))

```

4. Shannon and Simpson figure for publication

```{r}
# for publicaiton and matching aesthetics
# Define custom colors for thaw_type
thaw_colors <- c("no_thaw" = "#1b9e77", "thaw" = "#d95f02")  

# Shannon Diversity Plot
ggplot(shansimp.meta2, aes(x = site, y = Shannon, color = thaw_type)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +  # Remove outliers for clarity
  geom_jitter(width = 0.2, alpha = 0.6, size = 2) + # Add jittered points
  scale_color_manual(values = thaw_colors, labels=c('Pre-thaw', 'Post-thaw')) +  # Custom colors
  theme_classic() +
  xlab("Site") +
  theme_minimal() +
  ylab("Alpha Diversity Measure") +
  theme(text = element_text(size = 12)) +
  stat_compare_means(comparisons = list(c("thaw", "no_thaw", label = "p.signif", method = "t.test", ref.group = ".all.")))
  #stat_compare_means(comparisons = list(c("thaw", "no_thaw"), 
                     #label = "p.signif") +
```


```{r}
# Simpson Diversity Plot
ggplot(shansimp.meta2, aes(x = site, y = Simpson, color = thaw_type)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +  
  geom_jitter(width = 0.2, alpha = 0.6, size = 2) +  
  scale_color_manual(values = thaw_colors, labels=c('Pre-thaw', 'Post-thaw')) +  
  theme_classic() +
  xlab("Site") +
  ylab("Alpha Diversity Measure") +
  theme(text = element_text(size =12))
```

```{r}
# Facet wrap
# Reshape data to long format for faceting
library(tidyr)
library(ggplot2)

shansimp_long <- shansimp.meta2 %>%
  pivot_longer(cols = c(Shannon, Simpson), 
               names_to = "Diversity_Metric", 
               values_to = "Value")

# Define custom colors
thaw_colors <- c("no_thaw" = "#1b9e77", "thaw" = "#d95f02")  

# Create faceted plot
ggplot(shansimp_long, aes(x = site, y = Value, color = thaw_type)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +  # Remove outliers for clarity
  geom_jitter(width = 0.2, alpha = 0.6, size = 2) +  # Add jittered points
  scale_color_manual(values = thaw_colors, labels = c('Pre-thaw', 'Post-thaw')) +  
  theme_classic() +
  xlab("Site") +
  ylab("Alpha Diversity Measure") +
  facet_wrap(~Diversity_Metric, scales = "free_y") +  # Separate facets for Shannon and Simpson
  theme(text = element_text(size = 12)) +
  labs(color = "Treatment") + # Change legend title
  #stat_compare_means(comparisons = list(c("thaw", "no_thaw")
                     #label = "p.signif") +
  theme_minimal()

```
```{r}
library(ggplot2)
library(ggpubr)

# Create the plot with asterisks based on p-values
ggplot(shansimp_long, aes(x = site, y = Value, color = thaw_type)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +  
  geom_jitter(width = 0.2, alpha = 0.6, size = 2) +  
  scale_color_manual(values = thaw_colors, labels = c('Pre-thaw', 'Post-thaw')) +  
  theme_classic() +
  xlab("Site") +
  ylab("Alpha Diversity Measure") +
  facet_wrap(~Diversity_Metric, scales = "free_y") +  
  theme(text = element_text(size = 12)) +
  labs(color = "Treatment") +
  
  # Add significance asterisks (p.signif) and specify the test method
  stat_compare_means(comparisons = list(c("no_thaw", "thaw")),
                     label = "p.signif", 
                     method = "t.test") +  # You can change the method to t.test or others
  theme_minimal()

```



```{r}
# Plotting with significance codes
p_values <- data.frame(
  Diversity_Metric = c("Shannon", "Simpson"),
  site = c("CRREL", "FL", "Utqiagvik"),  # Adjust to match your x-axis categories
  p_value = c(0.04, 0.0029, 0.001)  # Example p-values
)
```

```{r}
# Load required libraries
library(tidyr)
library(ggplot2)
library(ggpubr)  # For stat_compare_means
library(ggsignif)  # For adding significance bars

# Reshape data to long format for faceting
shansimp_long <- shansimp.meta2 %>%
  pivot_longer(cols = c(Shannon, Simpson), 
               names_to = "Diversity_Metric", 
               values_to = "Value")

# Define custom colors
thaw_colors <- c("no_thaw" = "#6a7bd7", "thaw" = "#e94b3c")  

# Create faceted plot
ggplot(shansimp_long, aes(x = site, y = Value, color = thaw_type, shape = thaw_type)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +  # Remove outliers for clarity
  geom_jitter(width = 0.2, alpha = 0.6, size = 2) +  # Add jittered points
  scale_color_manual(values = thaw_colors, labels = c('Pre-thaw', 'Post-thaw')) +  
  scale_shape_manual(values = c("no_thaw" = 17, "thaw" = 16),
                     labels = c("Pre-thaw", "Post-thaw")) +  # Triangle and circle
  theme_classic() +
  xlab("Site") +
  ylab("Alpha Diversity Measure") +
  facet_wrap(~Diversity_Metric, scales = "free_y") +  # Separate facets for Shannon and Simpson
  theme(text = element_text(size = 12)) +
  labs(color = "Treatment") +  # Change legend title
  # #stat_compare_means(comparisons = list(c("no_thaw", "thaw")), 
  #                    label = "p.signif", method = "t.test") +  # Add p-values
  # geom_signif(comparisons = list(c("no_thaw", "thaw")),
  #             annotations = c("***"),  # Customize asterisks
  #             y_position = max(shansimp_long$Value, na.rm = TRUE) + 0.2, 
  #             tip_length = 0.02) +  # Adjust position
  theme_classic()
```
```{r}
# Define custom colors and shapes
thaw_colors <- c("no_thaw" = "#6a7bd7", "thaw" = "#e94b3c")
thaw_shapes <- c("no_thaw" = 17, "thaw" = 16)  # triangle, circle
thaw_labels <- c("no_thaw" = "Pre-thaw", "thaw" = "Post-thaw")

# Create faceted plot
ggplot(shansimp_long, aes(x = site, y = Value, color = thaw_type, shape = thaw_type)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.6, size = 2) +
  scale_color_manual(values = thaw_colors, labels = thaw_labels) +
  scale_shape_manual(values = thaw_shapes, labels = thaw_labels) +
  xlab("Site") +
  ylab("Alpha Diversity Measure") +
  facet_wrap(~Diversity_Metric, scales = "free_y") +
  theme_classic() +
  theme(text = element_text(size = 12)) +
  labs(color = "Treatment", shape = "Treatment") +
  
  
  
  
  
     geom_signif(
    comparisons = list(c("CRREL", "FL"), c("CRREL", "Utqiagvik")),  # Specify the groups to compare
    map_signif_level = TRUE,  # Automatically display significance codes like "*", "**", etc.
    annotations = c("***", "*"),  # You can manually annotate with custom significance labels
    y_position = c(42,43),  # Adjust y-position of the annotations
    tip_length = 0.03,
    textsize = 8)  
  #annotate( # unified legend title

```
```{r}
library(ggpubr)

# Filter only for Shannon
shannon_only <- shansimp_long %>% 
  filter(Diversity_Metric == "Shannon")

shannon.plot <- ggplot(shannon_only, aes(x = site, y = Value, color = thaw_type, shape = thaw_type)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7, position = position_dodge(width = 0.75)) +
  geom_jitter(alpha = 0.6, size = 4, position = position_dodge(width = 0.75)) +
  stat_compare_means(
    aes(group = thaw_type), 
    method = "wilcox.test", 
    label = "p.signif",
    label.y = max(shannon_only$Value, na.rm = TRUE) - 0.1, # adjust as needed
    hide.ns = TRUE, 
    size = 8, 
    show.legend = FALSE
  ) +
  scale_color_manual(values = thaw_colors, labels = thaw_labels) +
  scale_shape_manual(values = thaw_shapes, labels = thaw_labels) +
  xlab("Site") +
  ylab("Shannon Diversity") +
  theme_classic() +
  theme(text = element_text(size = 12)) +
  labs(color = "Treatment", shape = "Treatment")

# Remove legend for plotting with the other simpson plot
shannon.plot.noleg <- shannon.plot + theme(legend.position = "none")

```
Simpson plot
```{r}
library(ggpubr)

# Filter only for Shannon
simpson_only <- shansimp_long %>% 
  filter(Diversity_Metric == "Simpson")

simpson.plot <- ggplot(simpson_only, aes(x = site, y = Value, color = thaw_type, shape = thaw_type)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7, position = position_dodge(width = 0.75)) +
  geom_jitter(alpha = 0.6, size = 4, position = position_dodge(width = 0.75)) +
  scale_y_continuous(limits = c(0, 1.1)) +
  stat_compare_means(
    aes(group = thaw_type), 
    method = "wilcox.test", 
    label = "p.signif",
    label.y = max(simpson_only$Value, na.rm = TRUE) + 0.1, # adjust as needed
    hide.ns = TRUE, 
    size = 8, 
    show.legend = FALSE
  ) +
  scale_color_manual(values = thaw_colors, labels = thaw_labels) +
  scale_shape_manual(values = thaw_shapes, labels = thaw_labels) +
  xlab("Site") +
  ylab("Simpson Diversity") +
  theme_classic() +
  theme(text = element_text(size = 12)) +
  labs(color = "Treatment", shape = "Treatment")

print(simpson.plot)
```

```{r}
# Plot shannon and simpson plots together 
library(ggplot2)
library(cowplot)

# Step 3: Combine all plots 
alpha.combined_plots <- ggdraw() +
  draw_plot(shannon.plot.noleg, x = 0,   y = 0, width = 1/2, height = 1) +  # Left image
  draw_plot(simpson.plot,  x = 1/2, y = 0, width = 1/2, height = 1) +  # Center plot
  theme(plot.background = element_rect(fill= "white", color = NA)) #+
  #draw_plot_label(label = c("(A)", "(B)"), size = 16,
                  #x = c(0.05, 0.38), y = c(1, 1))

print(alpha.combined_plots)
ggsave("~/GitHub/thaw.response/Figures/shan.simp.combo.png", alpha.combined_plots, width = 12, height = 5)
```











```{r}
scale_shape_manual(values = c("no_thaw" = 17, "thaw" = 16)) +  # Triangle and circle
  theme_classic() +
  xlab("Site") +
  ylab("Alpha Diversity Measure") +
  facet_wrap(~Diversity_Metric, scales = "free_y") +
  theme(text = element_text(size = 12)) +
  labs(shape = "Treatment")  # Legend title for shape

```
Left off: Trying to get significance codes on the plot but I just can't seem to figure it out, may need to ask co-authors figure out another time, need to move on 

OLD CODE
```{r}
# coding to reorder pre and post in the boxplot figure 
# what you might have to do is work with the shan simp file instead of the phyloseq object and then use ggplot with 
# the reorder function & the rest of the code , or edit the phyloseq object so that it's no thaw and thaw so that pre comes before post 

# use the mutate function to re-name the pre and post in the phyloseeq object 
# example 
metadata <- data.frame("tissue type"=c("soil", "leaves", "leaf"), check.names = FALSE)
newdata <- dorm1rarefied %>% mutate(pre_post_thaw = replace(pre_post_thaw, pre== "before"))

plot <- plot_richness(dorm1rarefied, x = "site", color = "pre_post_thaw", measures = c("Shannon")) 
plot2 <- as.data.frame(plot)

# okay for shannon and simpson box plots, this is the code you have to use: 
# also before you dive in with the new data, you need to obtain the shannon and simpson values, put them in an excel sheet with metadata, 
# rename the pre_post_thaw column to thaw_type, and then change pre to no_thaw, and post to thaw
# then run using this code 

ggplot(shan_simp, aes(x = site, y = Shannon, color = thaw_type)) +
  geom_boxplot() + 
  theme_classic() +
  xlab("Site") +
  ylab("Alpha Diversity Measure") +
  scale_color_discrete("Treatment", labels=c('Pre-thaw', 'Post-thaw')) +
  theme(text = element_text(size = 12))

# Repeat above code with simpson diversity?

ggplot(shannon, aes(x = "site", y = "shannon", color = "pre_post_thaw")) +
  geom_boxplot() 
  theme_classic() +
  xlab("Site") +
  theme(text = element_text(size =12))

geom_boxplot(shannon, aes(x = "site", color = "pre_post_thaw"))

# Plot Simpson Diversity
plot_richness(dorm1rarefied, x = "site", color = "pre_post_thaw", measures = c("Simpson")) +
  geom_boxplot() +
  theme_classic() +
  xlab("Site") +
  theme(text = element_text(size =18)) +
  scale_color_discrete("Treatment")

# Replotting Simpson

plot_richness(dorm1rarefied, x = "site", measures = c("Simpson"), color = "pre_post_thaw", ) +
  geom_boxplot(aes(color = "pre_post_thaw")) +# need to reorder so that pre comes before post 
  geom_point(size = 0.25) +
  theme_classic() +
  xlab("Site") +
  theme(text = element_text(size =18)) +
  scale_color_discrete("Treatment")

plot_richness(dorm1rarefied, x = "site", color = "pre_post_thaw", measures = c("Simpson")) + 
  geom_boxplot()

# Make a boxplot of the number of OTUs and Shannon entropy 
plot_richness(dorm1rarefied, x = "site", measures = c("Observed", "Shannon")) + geom_boxplot()

# END
```


For publication: 
Plotting as a black and white figure, denoting pre as triangles and post as circles
```{r}
ggplot(shansimp_long, aes(x = site, y = Value, shape = thaw_type, color = thaw_type)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +  # Black boxplots
  geom_jitter(width = 0.2, alpha = 0.6, size = 2) +  # Points with shape aesthetic
  scale_shape_manual(values = c("no_thaw" = 17, "thaw" = 16)) +  # Triangle and circle
  theme_classic() +
  xlab("Site") +
  ylab("Alpha Diversity Measure") +
  facet_wrap(~Diversity_Metric, scales = "free_y") +
  theme(text = element_text(size = 12)) +
  labs(shape = "Treatment")  # Legend title for shape

```


Beta dispersion analysis: Permutest
```{r}
# Working with distance matrices 
# Creating a bray curtis distance matrix from our phyloseq object
dist.bc <- distance(thaw, method = "bray")
print(dist.bc)
ord <- ordinate(thaw, method = "PCoA", distance = dist.bc)
# Visualize
plot_ordination(thaw, ord, color = "site", shape = "pre_post_thaw") +
  stat_ellipse(aes(group = site))

# Calculate bray curtis distance matrix
thaw_bray <- phyloseq::distance(thaw, method = "bray")

# Making a data frame from the phyloseq object 
thawOTU <- thaw@otu_table %>%
  data.frame()

# Saving matrix
# write.csv(dormOTU,"~/Desktop/dormOTU.csv")

# Need to transform so species are columns and samples are rows (for vegan)
thawOTUtransform <- t(sqrt(thawOTU))
thawOTU_bray <- vegdist(thawOTUtransform, method = "bray")

# Make a data frame from the sample_data
sampledf <- data.frame(sample_data(thaw))

# Adonis test (PERMANOVA)
permanova1 <- adonis2(thawOTU_bray ~ pre_post_thaw, data = sampledf)
summary(permanova)
print(permanova)

# Adonis but get significance for pre_post_thaw and site
permanova2 <- adonis2(thawOTU_bray ~ site + pre_post_thaw, data = sampledf, permutations = 999, by = "margin")
print(permanova2)
summary(permanova2)



# Now check to see if we violate key assumption of permanova 
site_bd <- betadisper(thawOTU_bray, sampledf$site)
permutest(site_bd) # this is significant, which could indicate that permanova is influenced by unequal dispersion
plot(site_bd) 
TukeyHSD(site_bd)


# Now checking by treatment 
thaw_bd <-betadisper(thawOTU_bray, sampledf$pre_post_thaw)
permutest(thaw_bd) # not significant 
plot(thaw_bd)

# Homogeneity of dispersion test for site
beta <- betadisper(thawOTU_bray, thaw@sam_data$site) # testing the dispersion between groups (sites)
print(beta) # dispersion is how scattered the points are

#adonis2(thawOTU_bray ~ site, data=datadorm)

# Homogeneity of dispersion test for pre-post thaw 
beta_pp <- betadisper(thawOTU_bray, thaw@sam_data$pre_post_thaw)
print(beta_pp) 
permutest(beta_pp)

# Homogeneity of dispersion test for core
beta_core <- betadisper(thawOTU_bray, thaw_phyloseq@sam_data$core)
print(beta_core)

# Homogeneity of dispersion test for pre post thaw 
beta_pp <- betadisper(thawOTU_bray, thaw@sam_data$pre_post_thaw)
print(beta_pp)
# Are these distances (dispersion/distance to the median) significant?

# Site 
permutest(beta)
TukeyHSD(beta)

# Core 
#permutest(beta_core)

# Pre post thaw 
permutest(beta_pp)

# Tukey for site 
tukey_beta <- TukeyHSD(beta)
plot(tukey_beta) 
print(tukey_beta)

# Tukey for core 
# tukey_beta_core <- TukeyHSD(beta_core)
# print(tukey_beta_core)

# Tukey for pre post thaw  # not necessary since it wasnt significant
TukeyHSD(beta_pp)
print(tukey_beta_pp)


betadisper(thawOTU_bray, thaw@sam_data$pre_post_thaw)
```
```{r}
# Permanova to quantify and test how much of the variation is explained by site and thaw 
library(vegan)
permav
```

```{r}
# Making a data frame for PERMANOVA analysis
thaw.df <- data.frame(sample_data(thaw)) 

# site 
perms <- with(thaw.df, how(nperm = 1000, blocks = site))
# trying this with core 
# perms_core <- with(datadorm, how(nperm = 1000, blocks = core))

# pre post thaw? 
perms_pp <- with(thaw.df, how(nperm = 1000, blocks = pre_post_thaw))

# Permanova assessing composition by site and pre-post thaw 
sitethaw.pnova <- vegan::adonis2(thaw_bray ~ site + pre_post_thaw, data = thaw.df)
print(sitethaw.pnova)

# Permanova assessing composition by core and pre post thaw 
corethaw.pnova <- vegan::adonis2(thaw_bray ~ core + pre_post_thaw, data = thaw.df)
print(corethaw.pnova)

# Old code
#sitethaw.strata <- vegan::adonis2(thaw_bray ~ pre_post_thaw, permutations = perms, data = thaw.df)
#print(sitethaw.strata)

# SAVE
#saveRDS(dorm1rarefied, "~/Desktop/dorm1.rds")
#saveRDS(dorm_bray, "~/Desktop/dormbray.rds")
```

OLD and extra code
```{r}
#Alpha Diversity (this code was not used for analysis)
min_lib <- min(sample_sums(dorm1rarefied))

# Initialize matrices to store richness and evenness estimates
nsamp = nsamples(dorm1rarefied)
trials = 100

richness <- matrix(nrow = nsamp, ncol = trials)
row.names(richness) <- sample_names(dorm1rarefied)

evenness <- matrix(nrow = nsamp, ncol = trials)
row.names(evenness) <- sample_names(dorm1rarefied)

# It is always important to set a seed when you subsample so your result is replicable 
set.seed(3)

for (i in 1:100) {
  # Subsample
  r <- rarefy_even_depth(dorm1rarefied, sample.size = min_lib, verbose = FALSE, replace = TRUE)
  
  # Calculate richness
  rich <- as.numeric(as.matrix(estimate_richness(r, measures = "Observed")))
  richness[ ,i] <- rich
  
  # Calculate evenness
  even <- as.numeric(as.matrix(estimate_richness(r, measures = "InvSimpson")))
  evenness[ ,i] <- even
}
```


```{r}


# Using the base Kruskal test we find that it's the same as above with the FSA package Kruskal test 
#stats::kruskal.test(ssmeta$Simpson ~ site, data = ssmeta)
# Checking to see if Simpson is normally distributed
shapiro.test(ssmeta$Simpson) # this is not normally distributed, p = 1.541e-07

# Simpson by site 
kruskal.test(ssmeta$Simpson ~ site, data = ssmeta)
kruskal(ssmeta$Simpson, ssmeta$site, group=TRUE, p.adj="bonferroni")$groups
dunnTest(ssmeta$Simpson ~ site, data = ssmeta)
# Simpson by pre- post-thaw 
kruskal.test(ssmeta$Simpson ~ pre_post_thaw, data = ssmeta) #chi-squared = 18.564, df = 1, p-value = 1.643e-05
kruskal(ssmeta$Simpson, ssmeta$pre_post_thaw, group=TRUE, p.adj="bonferroni")$groups


```


```{r}

#Alpha diversity take two (https://micca.readthedocs.io/en/latest/phyloseq.html)
# Alpha diversity: plotting Shannon and Chao1
plot_richness(dorm1rarefied, measures = c("Chao1", "Shannon"))
shannon<- merge(ssmeta,shan)

# THIS NEEDS TO BE FIXED FOR PUBLICATION, YOU DON'T WANT POST AND PRE TO HAVE THE SAME BOX
# Plot Shannon diversity/making the plot prettier
plot_richness(dorm1rarefied, x = "site", measures = c("Shannon"), color = "pre_post_thaw") +
  geom_boxplot(aes(color = "pre_post_thaw")) + # need to reorder so that pre comes before post +
  theme_classic() +
  xlab("Site") +
  theme(text = element_text(size =18)) +
  scale_color_discrete("Treatment")


# Code for box plots above (trying to fix the above code)
plot_richness(dorm1rarefied, x = "site", measures = c("Shannon"), color = "pre_post_thaw") +
  geom_boxplot() +
  theme_classic() +
  xlab("Site") +
  theme(text = element_text(size =12)) +
  scale_color_discrete("Treatment")
```



Extra code that was not used:
```{r}
#********************************************************************************
# BELOW IS EXTRA CODE (THAT HAS NOT BEEN USED) FOR MAKING A BAR PLOT
#Stacked bar plot (https://deneflab.github.io/MicrobeMiseq/demos/mothur_2_phyloseq.html)
dorm_phylum <- dorm1rarefied %>%
  tax_glom(taxrank = "Phylum") %>%                     # agglomerate at phylum level
  transform_sample_counts(function(x) {x/sum(x)} ) %>% # Transform to rel. abundance
  psmelt() %>%                                         # Melt to long format
  filter(Abundance > 0.02) %>%                         # Filter out low abundance taxa
  arrange(Phylum)                                      # Sort data frame alphabetically by phylum

phylum_colors <- c(
  "#CBD588", "#5F7FC7", "orange","#DA5724", "#508578", "#CD9BCD",
  "#AD6F3B", "#673770","#D14285", "#652926", "#C84248", 
  "#8569D5", "#5E738F","#D1A33D", "#8A7C64", "#599861"
)


# Plot with ggplot 
ggplot(dorm_phylum, aes(x = pre_post_thaw, y = Abundance, fill = Phylum)) + 
  facet_grid(site~.) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = phylum_colors) +
  scale_x_discrete(
    breaks = c("7/8", "8/4", "9/2", "10/6"),
    labels = c("Jul", "Aug", "Sep", "Oct"), 
    drop = FALSE
  ) +
  # Remove x axis title
  theme(axis.title.x = element_blank()) + 
  #
  guides(fill = guide_legend(reverse = TRUE, keywidth = 1, keyheight = 1)) +
  ylab("Relative Abundance (Phyla > 2%) \n") 
  #ggtitle("Phylum Composition of Lake Erie \n Bacterial Communities by Sampling Site") 

# Scale reads to even depth 
dorm_scale <- dorm1rarefied %>%
  scale_reads(round = "round") 

# Ordination 
# Ordinate via PCOA
dorm_pcoa <- ordinate(
  physeq = dorm1rarefied, 
  method = "PCoA", 
  distance = "bray"
)

# Plot the PCOA
plot_ordination(
  physeq = dorm1rarefied,
  ordination = dorm_pcoa,
  color = "site",
  shape = "pre_post_thaw",
  title = "PCoA of permafrost bacterial communities pre- and post thaw"
) + 
  scale_color_manual(values = c("#a65628", "red", "#ffae19",
                                "#4daf4a", "#1919ff", "darkorchid3", "magenta")
  ) +
  geom_point(aes(color = site), alpha = 0.7, size = 4) +
  geom_point(colour = "grey90", size = 1.5)
```
More extra code that was not used:
```{r}
# Create a new dataframe to hold the means and standard deviations of richness estimates
SampleID <- row.names(richness)
mean <- apply(richness, 1, mean)
sd <- apply(richness, 1, sd)
measure <- rep("Richness", nsamp)
rich_stats <- data.frame(SampleID, mean, sd, measure)

# Create a new dataframe to hold the means and standard deviations of evenness estimates
SampleID <- row.names(evenness)
mean <- apply(evenness, 1, mean)
sd <- apply(evenness, 1, sd)
measure <- rep("Inverse Simpson", nsamp)
even_stats <- data.frame(SampleID, mean, sd, measure)

alpha <- rbind(rich_stats, even_stats)
s <- data.frame(sample_data(dorm1rarefied))
alphadiv <- merge(alpha, s) 

ggplot(alphadiv, aes(x = site, y = mean, color = site, group = site, shape = site)) +
  geom_point(size = 2) + 
  geom_line(size = 0.8) +
  facet_wrap(~measure, ncol = 1, scales = "free") +
  scale_color_manual(values = c("#E96446", "#302F3D", "#87CEFA")) +
  scale_x_discrete(
    breaks = c("7/8", "8/4", "9/2", "10/6"),
    labels = c("Jul", "Aug", "Sep", "Oct"), 
    drop = FALSE
  ) +
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank()
  )

```

