---
title: "3. Microbial diversity analysis"
author: "Joy O'Brien"
date: "2025-03-19"
output: html_document
---
Previously micro_stats.R (April 29, 2022)
# This script encompassess statistical tests that were performed on the phyloseq object

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("/Users/joyobrien/GitHub/thaw.response")
```

```{r}
# Load the necessary packages (this is from the phyloseq tutorial)
library("phyloseq")
library("ggplot2")      # graphics
library("readxl")       # necessary to import the data from Excel file
library("dplyr")        # filter and reformat data frames
library("tibble")       # Needed for converting column to row names
library("vegan")
library("viridis")
library("RColorBrewer")
library("colorRamps")
library("colorspace")
library("ggpubr")
library("readxl")
library("FSA")
library("agricolae") # used to get significance codes 
```
```{r}
# Load in the phyloseq object
thaw <- readRDS("~/GitHub/thaw.response/Data/thaw_phyloseq.rds")
```
0. Looking at top phylum and classes for fun
```{r}
# TOP 5 PHYLA
# Load necessary libraries
library(phyloseq)
library(ggplot2)

# Summarize taxa at the Phylum level
phyla_abundance <- tax_glom(thaw, taxrank = "Phylum")

# Convert phyloseq object to a dataframe
phyla_df <- as.data.frame(tax_table(phyla_abundance))
phyla_df$Abundance <- taxa_sums(phyla_abundance)

# Summarize and sort by abundance
phyla_df <- phyla_df[order(-phyla_df$Abundance), ]  # Sort in descending order
top_5_phyla <- phyla_df[1:5, ]  # Select top 5 phyla

# Plot using ggplot2
ggplot(top_5_phyla, aes(x = reorder(Phylum, -Abundance), y = Abundance, fill = Phylum)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Top 5 Most Abundant Phyla",
       x = "Phylum",
       y = "Abundance") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
```{r}
# TOP 10 PHYLA

top_10_phyla <- phyla_df[1:10, ]  # Select top 10 phyla

# Plot using ggplot2
ggplot(top_10_phyla, aes(x = reorder(Phylum, -Abundance), y = Abundance, fill = Phylum)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Top 10 Most Abundant Phyla",
       x = "Phylum",
       y = "Abundance") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

0. Looking at the top phyla and classes for fun
```{r}
# TOP 5 CLASSES
# Load necessary libraries
library(phyloseq)
library(ggplot2)

# Summarize taxa at the Phylum level
class_abundance <- tax_glom(thaw, taxrank = "Class")

# Convert phyloseq object to a dataframe
class_df <- as.data.frame(tax_table(class_abundance))
class_df$Abundance <- taxa_sums(class_abundance)

# Summarize and sort by abundance
class_df <- class_df[order(-class_df$Abundance), ]  # Sort in descending order
top_5_class <- class_df[1:5, ]  # Select top 5 classes

# Plot using ggplot2
ggplot(top_5_class, aes(x = reorder(Class, -Abundance), y = Abundance, fill = Class)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Top 5 Most Abundant Class",
       x = "Class",
       y = "Abundance") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
```{r}
# TOP 10 MOST ABUNDANT CLASS
top_10_class <- class_df[1:10, ]  # Select top 10 classes

# Plot using ggplot2
ggplot(top_10_class, aes(x = reorder(Class, -Abundance), y = Abundance, fill = Class)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Top 10 Most Abundant Class",
       x = "Class",
       y = "Abundance") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Top 5 Phylum pre-thaw 
```{r}
# TOP 5 PHYLA pre-thaw
presamples <- subset_samples(thaw, pre_post_thaw == "pre")
# Summarize taxa at the Phylum level
prephyla_abundance <- tax_glom(presamples, taxrank = "Phylum")

# Convert phyloseq object to a dataframe
prephyla_df <- as.data.frame(tax_table(prephyla_abundance))
prephyla_df$Abundance <- taxa_sums(prephyla_abundance)

# Summarize and sort by abundance
prephyla_df <- prephyla_df[order(-prephyla_df$Abundance), ]  # Sort in descending order
pretop_5_phyla <- prephyla_df[1:5, ]  # Select top 5 phyla

# Plot using ggplot2
ggplot(pretop_5_phyla, aes(x = reorder(Phylum, -Abundance), y = Abundance, fill = Phylum)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Top 5 Most Abundant Phyla PRE-THAW",
       x = "Phylum",
       y = "Abundance") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Top 5 phylum post-thaw
```{r}
# TOP 5 PHYLA pre-thaw
postsamples <- subset_samples(thaw, pre_post_thaw == "post")
# Summarize taxa at the Phylum level
postphyla_abundance <- tax_glom(postsamples, taxrank = "Phylum")

# Convert phyloseq object to a dataframe
postphyla_df <- as.data.frame(tax_table(postphyla_abundance))
postphyla_df$Abundance <- taxa_sums(postphyla_abundance)

# Summarize and sort by abundance
postphyla_df <- postphyla_df[order(-postphyla_df$Abundance), ]  # Sort in descending order
posttop_5_phyla <- postphyla_df[1:5, ]  # Select top 5 phyla

# Plot using ggplot2
ggplot(posttop_5_phyla, aes(x = reorder(Phylum, -Abundance), y = Abundance, fill = Phylum)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Top 5 Most Abundant Phyla POST-THAW",
       x = "Phylum",
       y = "Abundance") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

1. Shannon and Simpson Diversity metrics
```{r}
# Creating a table of Simpson and Shannon diversity
shansimp <- estimate_richness(thaw, measures = c("Simpson", "Shannon"))

# Visualize shannon
plot_richness(thaw, x = "site", color = "pre_post_thaw", measures = c("Shannon")) +
  geom_boxplot() +
  theme_classic() +
  xlab("Site") +
  theme(text = element_text(size =12)) +
  scale_color_discrete("Treatment")

# Visualize Simpson
plot_richness(thaw, x = "site", color = "pre_post_thaw", measures = c("Simpson")) +
  geom_boxplot() +
  theme_classic() +
  xlab("Site") +
  theme(text = element_text(size =12)) +
  scale_color_discrete("Treatment")

  
```


```{r}
# coding to reorder pre and post in the boxplot figure 
# what you might have to do is work with the shan simp file instead of the phyloseq object and then use ggplot with 
# the reorder function & the rest of the code , or edit the phyloseq object so that it's no thaw and thaw so that pre comes before post 

# use the mutate function to re-name the pre and post in the phyloseeq object 
# example 
metadata <- data.frame("tissue type"=c("soil", "leaves", "leaf"), check.names = FALSE)
newdata <- dorm1rarefied %>% mutate(pre_post_thaw = replace(pre_post_thaw, pre== "before"))

plot <- plot_richness(dorm1rarefied, x = "site", color = "pre_post_thaw", measures = c("Shannon")) 
plot2 <- as.data.frame(plot)

# okay for shannon and simpson box plots, this is the code you have to use: 
# also before you dive in with the new data, you need to obtain the shannon and simpson values, put them in an excel sheet with metadata, 
# rename the pre_post_thaw column to thaw_type, and then change pre to no_thaw, and post to thaw
# then run using this code 

ggplot(shan_simp, aes(x = site, y = Shannon, color = thaw_type)) +
  geom_boxplot() + 
  theme_classic() +
  xlab("Site") +
  ylab("Alpha Diversity Measure") +
  scale_color_discrete("Treatment", labels=c('Pre-thaw', 'Post-thaw')) +
  theme(text = element_text(size = 12))

# Repeat above code with simpson diversity?

ggplot(shannon, aes(x = "site", y = "shannon", color = "pre_post_thaw")) +
  geom_boxplot() 
  theme_classic() +
  xlab("Site") +
  theme(text = element_text(size =12))

geom_boxplot(shannon, aes(x = "site", color = "pre_post_thaw"))

# Plot Simpson Diversity
plot_richness(dorm1rarefied, x = "site", color = "pre_post_thaw", measures = c("Simpson")) +
  geom_boxplot() +
  theme_classic() +
  xlab("Site") +
  theme(text = element_text(size =18)) +
  scale_color_discrete("Treatment")

# Replotting Simpson

plot_richness(dorm1rarefied, x = "site", measures = c("Simpson"), color = "pre_post_thaw", ) +
  geom_boxplot(aes(color = "pre_post_thaw")) +# need to reorder so that pre comes before post 
  geom_point(size = 0.25) +
  theme_classic() +
  xlab("Site") +
  theme(text = element_text(size =18)) +
  scale_color_discrete("Treatment")

plot_richness(dorm1rarefied, x = "site", color = "pre_post_thaw", measures = c("Simpson")) + 
  geom_boxplot()

# Make a boxplot of the number of OTUs and Shannon entropy 
plot_richness(dorm1rarefied, x = "site", measures = c("Observed", "Shannon")) + geom_boxplot()

# END
```





A. Obtain metrics
```{r}
# Creating a table of Simpson and Shannon diversity and run stats
shan.simp <- as.data.frame(shansimp)

# Add metadata to this dataframe
metadata_df <- as.data.frame(sample_data(thaw))
shansimp.meta <- merge(metadata_df, shan.simp, by = "row.names", all.x = TRUE)

# Rename the first column after the merge
#colnames(shansimp.meta)[1] <- "sample_ID"

# View the merged dataframe
head(shansimp.meta)

# Save the table as a CSV file
#write.csv(shansimp.meta, "~/GitHub/thaw.response/Data/thaw_diversity_metrics.csv", row.names = FALSE)

```
B. Check for normality
```{r}
# Before running stats, check to see if data is normally distributed
# Checking normal distribution: SHANNON
hist(shansimp.meta$Shannon, main = "Shannon index", xlab = "")
shapiro.test(shansimp.meta$Shannon) # not normally distributed W = 0.89436, p-value = 0.0008473

# Checking normal distribution: SIMPSON
hist(shansimp.meta$Simpson, main = "Simpson index", xlab = "")
shapiro.test(shansimp.meta$Simpson) # not normally distributed W = 0.57395, p-value = 5.77e-10
```

C. Testing for differences in Shannon and Simpson by site
```{r}
# Check to see if Shannon differs significantly between sites regardless of treatment
# Recall that Shannon is not normally distributed
kruskal.test(shansimp.meta$Shannon ~ site, data = shansimp.meta)
dunnTest(shansimp.meta$Shannon ~ site, data = shansimp.meta)

# Check to see if Simpson differs significantly between sites
kruskal.test(shansimp.meta$Simpson ~ site, data = shansimp.meta)
dunnTest(shansimp.meta$Simpson ~ site, data = shansimp.meta)
```

```{r}
library(dplyr)

shansimp.meta1 <- shansimp.meta  # Make a copy of the original dataframe

# Rename the column 'pre_post_thaw' to 'thaw_type'
colnames(shansimp.meta1)[colnames(shansimp.meta1) == "pre_post_thaw"] <- "thaw_type"

# Replace 'pre_thaw' with 'no_thaw' and 'post_thaw' with 'thaw'
shansimp.meta2 <- shansimp.meta1 %>%
  mutate(thaw_type = recode(thaw_type, "pre" = "no_thaw", "post" = "thaw"))

# Now that this is changed, we can plot 
ggplot(shansimp.meta2, aes(x = site, y = Shannon, color = thaw_type)) +
  geom_boxplot() + 
  theme_classic() +
  xlab("Site") +
  ylab("Alpha Diversity Measure") +
  scale_color_discrete("Treatment", labels=c('Pre-thaw', 'Post-thaw')) +
  theme(text = element_text(size = 12)) 

# Repeat above code with simpson diversity

ggplot(shansimp.meta2, aes(x = site, y = Simpson, color = thaw_type)) +
  geom_boxplot() +
  theme_classic() +
  xlab("Site") +
  ylab("Alpha Diversity Measure") +
  scale_color_discrete("Treatment", labels=c('Pre-thaw', 'Post-thaw')) +
  theme(text = element_text(size =12))
```


```{r}
# for publicaiton and matching aesthetics
# Define custom colors for thaw_type
thaw_colors <- c("no_thaw" = "#1b9e77", "thaw" = "#d95f02")  

# Shannon Diversity Plot
ggplot(shansimp.meta2, aes(x = site, y = Shannon, color = thaw_type)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +  # Remove outliers for clarity
  geom_jitter(width = 0.2, alpha = 0.6, size = 2) + # Add jittered points
  scale_color_manual(values = thaw_colors, labels=c('Pre-thaw', 'Post-thaw')) +  # Custom colors
  theme_classic() +
  xlab("Site") +
  ylab("Alpha Diversity Measure") +
  theme(text = element_text(size = 12))

# Simpson Diversity Plot
ggplot(shansimp.meta2, aes(x = site, y = Simpson, color = thaw_type)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +  
  geom_jitter(width = 0.2, alpha = 0.6, size = 2) +  
  scale_color_manual(values = thaw_colors, labels=c('Pre-thaw', 'Post-thaw')) +  
  theme_classic() +
  xlab("Site") +
  ylab("Alpha Diversity Measure") +
  theme(text = element_text(size =12))
```
```{r}
# Facet wrap
# Reshape data to long format for faceting
library(tidyr)
library(ggplot2)

shansimp_long <- shansimp.meta2 %>%
  pivot_longer(cols = c(Shannon, Simpson), 
               names_to = "Diversity_Metric", 
               values_to = "Value")

# Define custom colors
thaw_colors <- c("no_thaw" = "#1b9e77", "thaw" = "#d95f02")  

# Create faceted plot
ggplot(shansimp_long, aes(x = site, y = Value, color = thaw_type)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +  # Remove outliers for clarity
  geom_jitter(width = 0.2, alpha = 0.6, size = 2) +  # Add jittered points
  scale_color_manual(values = thaw_colors, labels = c('Pre-thaw', 'Post-thaw')) +  
  theme_classic() +
  xlab("Site") +
  ylab("Alpha Diversity Measure") +
  facet_wrap(~Diversity_Metric, scales = "free_y") +  # Separate facets for Shannon and Simpson
  theme(text = element_text(size = 12))

```

C. Subset Shannon and Simpson by site
```{r}
# Subset data by site and check for normality
sscrrel <- subset(shansimp.meta, site =="CRREL")
ssfl <- subset(shansimp.meta, site =="FL")
ssut <- subset(shansimp.meta, site =="Utqiagvik")

# Testing normality for each site: SHANNON
shapiro.test(sscrrel$Shannon) # p = 0.9954 normally distributed, use anova 
shapiro.test(ssfl$Shannon) # p = 0.02876 not normally distributed, use kruskal
shapiro.test(ssut$Shannon) # p = 0.001398 not normally distributed, use kruskal 

# Testing normality for each site: SIMPSON
shapiro.test(sscrrel$Simpson) # p = 0.03035 # use kruskal to evaluate 
shapiro.test(ssfl$Simpson) # p = 0.00119 # use kruskal
shapiro.test(ssut$Simpson) # p = 3.407e-05 not normally distributed, use kruskal
```


D. Min and Max values for Shannon and Simpson
```{r}
# Load dplyr for grouping and summarizing the data
library(dplyr)

# Calculate min and max Shannon by treatment
shannon_by_treatment <- shansimp.meta %>%
  group_by(pre_post_thaw) %>%
  summarise(
    min_shannon = min(Shannon),
    max_shannon = max(Shannon)
  )

# Calculate min and max Simpson by treatment
simpson_by_treatment <- shansimp.meta %>%
  group_by(pre_post_thaw) %>%
  summarise(
    min_simpson = min(Simpson),
    max_simpson = max(Simpson)
  )

# Print the results
print(shannon_by_treatment)
print(simpson_by_treatment)
```
E. Mean and SD for Shannon and Simpson
```{r}
# Obtain average Shannon values for each site with SD
# Load dplyr for data manipulation
library(dplyr)

# Calculate min, max, and standard error for Shannon by site
shannon_by_site <- shansimp.meta %>%
  group_by(site) %>%
  summarise(
    avg_shannon = mean(Shannon),
    se_shannon = sd(Shannon) / sqrt(n())
  )

# Calculate min, max, and standard error for Simpson by site
simpson_by_site <- shansimp.meta %>%
  group_by(site) %>%
  summarise(
    avg_simpson = mean(Simpson),
    se_simpson = sd(Simpson) / sqrt(n())
  )

# Print the results
print(shannon_by_site)
print(simpson_by_site)

```

```{r}

```

```{r}
# Shannon by site
kruskal.test(ssmeta$Shannon ~ site, data = ssmeta)# chi-squared = 13.765, df = 2, p-value = 0.001026
kruskal(ssmeta$Shannon, ssmeta$site, group=TRUE, p.adj="bonferroni")$groups #using this to get the letter significance codes 
dunnTest(ssmeta$Shannon ~ site, data = ssmeta)

# Shannon by pre thaw 
kruskal.test(ssmeta$Shannon ~ pre_post_thaw, data = ssmeta) # chi-squared = 14.702, df = 1, p-value = 0.0001259
kruskal(ssmeta$Shannon, ssmeta$pre_post_thaw, group=TRUE, p.adj="bonferroni")$groups # to get significance codes

# Shannon by pre- post-thaw by site 
# CRREL
kruskal.test(ssmeta$Shannon ~ pre_post_thaw, site == "CRREL", data = ssmeta) # chi-squared = 6.5455, df = 1, p-value = 0.01052
kruskal(ssmeta$Shannon, ssmeta$pre_post_thaw, site == "CRREL", group=TRUE, p.adj="bonferroni") #$groups

# Farmers Loop 
kruskal.test(ssmeta$Shannon ~ pre_post_thaw, site == "FL", data = ssmeta) # chi-squared = 6, df = 1, p-value = 0.01431

# Utqiagvik
kruskal.test(ssmeta$Shannon ~ pre_post_thaw, site == "Utqiagvik", data = ssmeta) # chi-squared = 8.0769, df = 1, p-value = 0.004483

# Subsetting by site manually 
crrel_ss <- subset(ssmeta,subset =  site == "CRREL")
kruskal.test(crrel_ss$Shannon ~ pre_post_thaw, data = crrel_ss)

#dunnTest(ssmeta$Shannon ~ pre_post_thaw, data = ssmeta) 
```




```{r}
# Working with distance matrices 
# Creating a bray curtis distance matrix from our phyloseq object
dist.bc <- distance(dorm1rarefied, method = "bray")
print(dist.bc)
# Unifrac
dist.uf <- distance(dorm1rarefied, method = "unifrac")
ord <- ordinate(dorm1rarefied, method = "MDS", distance = "bray")
# Visualize
plot_ordination(dorm1rarefied, ord, color = "site", shape = "pre_post_thaw") +
  stat_ellipse(aes(group = site))
# Calculate bray curtis distance matrix
dorm_bray <- phyloseq::distance(dorm1rarefied, method = "bray")

# Making a data frame from the phyloseq object 
dormOTU <- dorm1rarefied_OTU %>%
  data.frame()

# Saving matrix
# write.csv(dormOTU,"~/Desktop/dormOTU.csv")

# Need to transform so species are columns and samples are rows (for vegan)
dormOTUtransform <- t(sqrt(dormOTU))
dorm_bray <- vegdist(dormOTUtransform, method = "bray")

# Make a data frame from the sample_data
# sampledf <- data.frame(sample_data(dorm1rarefied))

# Adonis test (PERMANOVA)
# adonis(dorm_bray ~ pre_post_thaw, data = sampledf)

# Homogeneity of dispersion test for site
beta <- betadisper(dorm_bray, dorm1rarefied_sam_data$site) # testing the dispersion between groups (sites)
print(beta) # dispersion is how scattered the points are

adonis2(dorm_bray ~ site, data=datadorm)

# Homogeneity of dispersion test for core
beta_core <- betadisper(dorm_bray, dorm1rarefied_sam_data$core)
print(beta_core)

# Homogeneity of dispersion test for pre post thaw 
beta_pp <- betadisper(dorm_bray, dorm1rarefied_sam_data$pre_post_thaw)
print(beta_pp)
# Are these distances (dispersion/distance to the median) significant?

# Site 
permutest(beta)

# Core 
# permutest(beta_core)

# Pre post thaw 
permutest(beta_pp)

# Tukey for site 
tukey_beta <- TukeyHSD(beta)
plot(tukey_beta) 
print(tukey_beta)

# Tukey for core 
# tukey_beta_core <- TukeyHSD(beta_core)
# print(tukey_beta_core)

# Tukey for pre post thaw  # not necessary since it wasnt significant
tukey_beta_pp <- TukeyHSD(beta_pp)
print(tukey_beta_pp)
```

```{r}
# Making a data frame for PERMANOVA analysis
datadorm <- data.frame(sample_data(dorm1rarefied)) 

# site 
perms <- with(datadorm, how(nperm = 1000, blocks = site))
# trying this with core 
# perms_core <- with(datadorm, how(nperm = 1000, blocks = core))

# pre post thaw? 
perms_pp <- with(datadorm, how(nperm = 1000, blocks = pre_post_thaw))

# Permanova assessing composition by site and pre-post thaw 
sitethaw.pnova <- vegan::adonis2(dorm_bray ~ site + pre_post_thaw, data = datadorm)
print(sitethaw.pnova)

# Permanova assessing composition by core and pre post thaw 
corethaw.pnova <- vegan::adonis2(dorm_bray ~ core + pre_post_thaw, data = datadorm)
print(corethaw.pnova)

sitethaw.strata <- vegan::adonis2(dorm_bray ~ pre_post_thaw, permutations = perms, data = datadorm)
print(sitethaw.strata)

# SAVE
saveRDS(dorm1rarefied, "~/Desktop/dorm1.rds")
saveRDS(dorm_bray, "~/Desktop/dormbray.rds")
 
#Alpha Diversity (this code was not used for analysis)
min_lib <- min(sample_sums(dorm1rarefied))

# Initialize matrices to store richness and evenness estimates
nsamp = nsamples(dorm1rarefied)
trials = 100

richness <- matrix(nrow = nsamp, ncol = trials)
row.names(richness) <- sample_names(dorm1rarefied)

evenness <- matrix(nrow = nsamp, ncol = trials)
row.names(evenness) <- sample_names(dorm1rarefied)

# It is always important to set a seed when you subsample so your result is replicable 
set.seed(3)

for (i in 1:100) {
  # Subsample
  r <- rarefy_even_depth(dorm1rarefied, sample.size = min_lib, verbose = FALSE, replace = TRUE)
  
  # Calculate richness
  rich <- as.numeric(as.matrix(estimate_richness(r, measures = "Observed")))
  richness[ ,i] <- rich
  
  # Calculate evenness
  even <- as.numeric(as.matrix(estimate_richness(r, measures = "InvSimpson")))
  evenness[ ,i] <- even
}
```


```{r}
# Simpson pre post thaw by site (obtaining the significance codes)

# CRREL
kruskal.test(sscrrel$Simpson ~ pre_post_thaw, data = sscrrel)
kruskal(sscrrel$Simpson, sscrrel$pre_post_thaw, group=TRUE, p.adj="bonferroni")$groups
# FL
kruskal.test(ssfl$Simpson ~ pre_post_thaw, data = ssfl)
kruskal(ssfl$Simpson, ssfl$pre_post_thaw, group=TRUE, p.adj="bonferroni")$groups
# Utqiagvik 
kruskal.test(ssut$Simpson ~ pre_post_thaw, data = ssut)
kruskal(ssut$Simpson, ssut$pre_post_thaw, group=TRUE, p.adj="bonferroni")$groups
# Simpson by site 
kruskal.test(ssmeta$Simpson ~ site, data = ssmeta) # chi-squared = 9.5593, df = 2, p-value = 0.008399
dunnTest(ssmeta$Simpson ~ site, data = ssmeta)

# Using the base Kruskal test we find that it's the same as above with the FSA package Kruskal test 
#stats::kruskal.test(ssmeta$Simpson ~ site, data = ssmeta)
# Checking to see if Simpson is normally distributed
shapiro.test(ssmeta$Simpson) # this is not normally distributed, p = 1.541e-07

# Simpson by site 
kruskal.test(ssmeta$Simpson ~ site, data = ssmeta)
kruskal(ssmeta$Simpson, ssmeta$site, group=TRUE, p.adj="bonferroni")$groups
dunnTest(ssmeta$Simpson ~ site, data = ssmeta)
# Simpson by pre- post-thaw 
kruskal.test(ssmeta$Simpson ~ pre_post_thaw, data = ssmeta) #chi-squared = 18.564, df = 1, p-value = 1.643e-05
kruskal(ssmeta$Simpson, ssmeta$pre_post_thaw, group=TRUE, p.adj="bonferroni")$groups

# Simpson by pre-post thaw by site 
# CRREL
kruskal.test(ssmeta$Simpson ~ pre_post_thaw, site == "CRREL", data = ssmeta) # chi-squared = 6.5455, df = 1, p-value = 0.01052

# FL
kruskal.test(ssmeta$Simpson ~ pre_post_thaw, site == "FL", data = ssmeta) # chi-squared = 6, df = 1, p-value = 0.01431

# Utqiagvik
kruskal.test(ssmeta$Simpson ~ pre_post_thaw, site == "Utqiagvik", data = ssmeta) # chi-squared = 8.0769, df = 1, p-value = 0.004483
```


```{r}

#Alpha diversity take two (https://micca.readthedocs.io/en/latest/phyloseq.html)
# Alpha diversity: plotting Shannon and Chao1
plot_richness(dorm1rarefied, measures = c("Chao1", "Shannon"))
shannon<- merge(ssmeta,shan)

# THIS NEEDS TO BE FIXED FOR PUBLICATION, YOU DON'T WANT POST AND PRE TO HAVE THE SAME BOX
# Plot Shannon diversity/making the plot prettier
plot_richness(dorm1rarefied, x = "site", measures = c("Shannon"), color = "pre_post_thaw") +
  geom_boxplot(aes(color = "pre_post_thaw")) + # need to reorder so that pre comes before post +
  theme_classic() +
  xlab("Site") +
  theme(text = element_text(size =18)) +
  scale_color_discrete("Treatment")


# Code for box plots above (trying to fix the above code)
plot_richness(dorm1rarefied, x = "site", measures = c("Shannon"), color = "pre_post_thaw") +
  geom_boxplot() +
  theme_classic() +
  xlab("Site") +
  theme(text = element_text(size =12)) +
  scale_color_discrete("Treatment")
```


```


Extra code that was not used:
```{r}
#********************************************************************************
# BELOW IS EXTRA CODE (THAT HAS NOT BEEN USED) FOR MAKING A BAR PLOT
#Stacked bar plot (https://deneflab.github.io/MicrobeMiseq/demos/mothur_2_phyloseq.html)
dorm_phylum <- dorm1rarefied %>%
  tax_glom(taxrank = "Phylum") %>%                     # agglomerate at phylum level
  transform_sample_counts(function(x) {x/sum(x)} ) %>% # Transform to rel. abundance
  psmelt() %>%                                         # Melt to long format
  filter(Abundance > 0.02) %>%                         # Filter out low abundance taxa
  arrange(Phylum)                                      # Sort data frame alphabetically by phylum

phylum_colors <- c(
  "#CBD588", "#5F7FC7", "orange","#DA5724", "#508578", "#CD9BCD",
  "#AD6F3B", "#673770","#D14285", "#652926", "#C84248", 
  "#8569D5", "#5E738F","#D1A33D", "#8A7C64", "#599861"
)


# Plot with ggplot 
ggplot(dorm_phylum, aes(x = pre_post_thaw, y = Abundance, fill = Phylum)) + 
  facet_grid(site~.) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = phylum_colors) +
  scale_x_discrete(
    breaks = c("7/8", "8/4", "9/2", "10/6"),
    labels = c("Jul", "Aug", "Sep", "Oct"), 
    drop = FALSE
  ) +
  # Remove x axis title
  theme(axis.title.x = element_blank()) + 
  #
  guides(fill = guide_legend(reverse = TRUE, keywidth = 1, keyheight = 1)) +
  ylab("Relative Abundance (Phyla > 2%) \n") 
  #ggtitle("Phylum Composition of Lake Erie \n Bacterial Communities by Sampling Site") 

# Scale reads to even depth 
dorm_scale <- dorm1rarefied %>%
  scale_reads(round = "round") 

# Ordination 
# Ordinate via PCOA
dorm_pcoa <- ordinate(
  physeq = dorm1rarefied, 
  method = "PCoA", 
  distance = "bray"
)

# Plot the PCOA
plot_ordination(
  physeq = dorm1rarefied,
  ordination = dorm_pcoa,
  color = "site",
  shape = "pre_post_thaw",
  title = "PCoA of permafrost bacterial communities pre- and post thaw"
) + 
  scale_color_manual(values = c("#a65628", "red", "#ffae19",
                                "#4daf4a", "#1919ff", "darkorchid3", "magenta")
  ) +
  geom_point(aes(color = site), alpha = 0.7, size = 4) +
  geom_point(colour = "grey90", size = 1.5)
```
More extra code that was not used:
```{r}
# Create a new dataframe to hold the means and standard deviations of richness estimates
SampleID <- row.names(richness)
mean <- apply(richness, 1, mean)
sd <- apply(richness, 1, sd)
measure <- rep("Richness", nsamp)
rich_stats <- data.frame(SampleID, mean, sd, measure)

# Create a new dataframe to hold the means and standard deviations of evenness estimates
SampleID <- row.names(evenness)
mean <- apply(evenness, 1, mean)
sd <- apply(evenness, 1, sd)
measure <- rep("Inverse Simpson", nsamp)
even_stats <- data.frame(SampleID, mean, sd, measure)

alpha <- rbind(rich_stats, even_stats)
s <- data.frame(sample_data(dorm1rarefied))
alphadiv <- merge(alpha, s) 

ggplot(alphadiv, aes(x = site, y = mean, color = site, group = site, shape = site)) +
  geom_point(size = 2) + 
  geom_line(size = 0.8) +
  facet_wrap(~measure, ncol = 1, scales = "free") +
  scale_color_manual(values = c("#E96446", "#302F3D", "#87CEFA")) +
  scale_x_discrete(
    breaks = c("7/8", "8/4", "9/2", "10/6"),
    labels = c("Jul", "Aug", "Sep", "Oct"), 
    drop = FALSE
  ) +
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank()
  )

```

