---
title: "Respiration and copy number analysis"
author: "Joy O'Brien"
date: "2025-03-24"
output: html_document
---
Note: Deleted Fl1C4 R8 from the respiration data because we do not have the copy number counterpart 

1. Set working directory and load libraries
```{r setup, include=FALSE}
setwd("~/GitHub/thaw.response")

# Load the necessary libraries
library(vegan)
library(readxl)
library(dplyr)
library(ggplot2)
library(tidyr)
library(e1071) #for skewness function
library(FSA)

```

2. Load data
```{r}
# Read in the data from excel 
cpnumb <- read_excel("~/GitHub/thaw.response/Data/16Scopy_number.xlsx", sheet = "16S_copy")
resp4copy <- read_excel("~/GitHub/thaw.response/Data/cumulative_resp.xlsx", sheet = "resp")
thaw <- readRDS("~/GitHub/thaw.response/Data/thaw_phyloseq.rds")
```

3. Calculate absolute abundance using OTU table and 16S copy numbers
```{r}
# Extract raw OTU count table and make a matrix
otu_table <- as.data.frame(thaw@otu_table)
otu_mat <- as.matrix(otu_table)

# 2. Extract qPCR copy numbers into a named vector
qPCR_vec <- setNames(cpnumb$copy_numb_gsoil, cpnumb$Sample_ID)

# 3. Match the sample order to OTU matrix columns
qPCR_vec <- qPCR_vec[colnames(otu_mat)]

# 4. Calculate total sequencing reads per sample
total_reads <- colSums(otu_mat)

# 5. Calculate scaling factors
scaling_factors <- qPCR_vec / total_reads

# 6. Multiply each column (sample) in the OTU matrix by its scaling factor
otu_abs_abund <- sweep(otu_mat, 2, scaling_factors, FUN = "*")


# Checking the potential error / mismatch 
# Sample names in OTU table
otu_samples <- colnames(otu_mat)

# Sample names in qPCR data
qpcr_samples <- cpnumb$Sample_ID

# Find which OTU samples are missing in qPCR
missing_in_qpcr <- otu_samples[!otu_samples %in% qpcr_samples]

# Also check the reverse: which qPCR samples aren't in OTU table
missing_in_otu <- qpcr_samples[!qpcr_samples %in% otu_samples]

# Print to investigate
missing_in_qpcr
missing_in_otu

```

3A. Identifying and Removing samples that don't align in the two datasets
```{r}
# scaling_factors came from: qPCR_vec / total_reads
na_samples <- names(scaling_factors)[is.na(scaling_factors)]
na_samples

#–– If you’re working with plain matrices / data frames ––
otu_mat      <- otu_mat[ , !colnames(otu_mat)      %in% na_samples]
total_reads  <- total_reads[ !names(total_reads)   %in% na_samples]
qPCR_vec     <- qPCR_vec[  !names(qPCR_vec)        %in% na_samples]

scaling_factors <- qPCR_vec / total_reads          # now no NAs
otu_abs_abund   <- sweep(otu_mat, 2, scaling_factors, FUN = "*")



```
3B. Reset and filter
```{r}
# First, ensure both are named vectors
names(qPCR_vec)     # should return sample names
names(total_reads)  # should return sample names

# Align by intersecting names
shared_samples <- intersect(names(qPCR_vec), names(total_reads))

# Subset both to shared samples
qPCR_vec      <- qPCR_vec[shared_samples]
total_reads   <- total_reads[shared_samples]

# Now calculate scaling factors
scaling_factors <- qPCR_vec / total_reads

```

3C. Confirming that they align
```{r}
length(scaling_factors)
any(is.na(scaling_factors))  # should be FALSE

```
3D. Calculating for absolute_otu
```{r}
# Make sure samples in scaling_factors match the columns in otu_mat
otu_mat_subset <- otu_mat[, names(scaling_factors)]

# Multiply each sample column by its corresponding scaling factor
absolute_otu <- sweep(otu_mat_subset, 2, scaling_factors, `*`)
```

4. Subsetting copynumber data to analyze
```{r}
# Subsetting the data so that we can analyze 
library(dplyr)

# Subset data for 'pre' thaw samples
pre_thaw_copy <- cpnumb %>%
  filter(pre_post_thaw == "pre")

# Subset data for 'post' thaw samples
post_thaw_copy <- cpnumb %>%
  filter(pre_post_thaw == "post")

# Make sure that we have the right number of observations
unique <- setdiff(pre_thaw_copy$sample_ID, resp4copy$Sample_ID)
print(unique)
# Visualizing to start off
scatter.smooth(pre_thaw_copy$copy_numb_gsoil, y = resp4copy$Cumulative_Respiration)

# Checking for normality
shapiro.test(pre_thaw_copy$copy_numb_gsoil) #p-value = 0.007103 Data is not normal
shapiro.test(resp4copy$Cumulative_Respiration) #p-value = 0.02446 data is not normal 

```

1. Looking at pre-thaw copy number and cumulative respiration
```{r}
# Before merging the datasets, we want to drop the PRE_SOIL from the sample ID name in pre_thaw_copy so we can smoothly merge with the resp data

pre_thaw_copy$Sample_ID <- sub("_PRE_SOIL", "", pre_thaw_copy$Sample_ID)


# Merge datasets 
# Merge the data frames (pre-thaw copy number and resp)

pre_respcpnumb <- merge(pre_thaw_copy, resp4copy, by = "Sample_ID")

# Remove extra columns 
pre_respcpnumb<- pre_respcpnumb[ , !(names(pre_respcpnumb) %in% c("site.y", "core.y")) ]
# Remove the .x from site and core 
names(pre_respcpnumb) <- sub("\\.x$", "", names(pre_respcpnumb))
```


1A. Spearman Correlation for cumulative respiration and pre-thaw copy number
```{r}
# FIRST LET'S RUN A CORRELATION FOR RESPIRATION ~ PRE THAW COPY NUMBER 
resp_pre_corr <- cor.test(x =pre_respcpnumb$copy_numb_gsoil, y = pre_respcpnumb$Cumulative_Respiration, method = "spearman")
print(resp_pre_corr)
```
1B. Visualize the relationship here
```{r}
# Visualize here 
# ggplot(pre_respcpnumb, aes(x = copy_numb_gsoil, y = Cumulative_Respiration)) + 
#   geom_point(aes(color = site)) +
#   scale_x_log10() +
#   scale_y_log10() +
#   geom_smooth(method = "lm", color = "black", linetype = "dashed", fill = "gray", alpha = 0.2, se = TRUE) +
#   xlab("log10 Pre-thaw copy number g⁻¹ soil") +
#   ylab("log10 Cumulative Respiration (µg C-CO₂ g⁻¹ dry soil)") +
#   theme(text = element_text(size = 12)) +
#   scale_color_discrete("Site") +
#   theme_classic()

pre.plot <- ggplot(pre_respcpnumb, aes(x = copy_numb_gsoil, y = Cumulative_Respiration)) + 
  geom_point(aes(color = site)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(method = "lm", color = "black", linetype = "dashed", fill = "gray", alpha = 0.2, se = TRUE) +
  labs(
    x = expression("log"[10] * " Pre-thaw copy number g"^{-1} * " soil"),
    y = expression("log"[10] * " Cumulative Respiration (" * mu * "g C-CO"[2] * " g"^{-1} * " dry soil)")
  ) +
  scale_color_manual(
    values = c("#1b9e77", "#d95f02", "#7570b3"),
    name = "Site"
  ) +
  theme(text = element_text(size = 12)) +
  theme_classic()

```

2. Looking at post-thaw copy number and cumulative respiration

```{r}
# Before merging the datasets, we want to drop the POST_SOIL from the sample ID name in pre_thaw_copy so we can smoothly merge with the resp data

post_thaw_copy$Sample_ID <- sub("_POST_SOIL", "", post_thaw_copy$Sample_ID)


# Merge datasets 
# Merge the data frames (pre-thaw copy number and resp)

post_respcpnumb <- merge(post_thaw_copy, resp4copy, by = "Sample_ID")

# Remove extra columns 
post_respcpnumb<- post_respcpnumb[ , !(names(post_respcpnumb) %in% c("site.y", "core.y")) ]
# Remove the .x from site and core 
names(post_respcpnumb) <- sub("\\.x$", "", names(post_respcpnumb))
```

2A. Spearman Correlation for cumulative respiration and post-thaw copy number
```{r}
# FIRST LET'S RUN A CORRELATION FOR RESPIRATION ~ POST THAW COPY NUMBER 
resp_post_corr <- cor.test(x =post_respcpnumb$copy_numb_gsoil, y = post_respcpnumb$Cumulative_Respiration, method = "spearman")
print(resp_post_corr)
```
2B. Visualizing 

```{r}
ggplot(post_respcpnumb, aes(x = copy_numb_gsoil, y = Cumulative_Respiration)) + 
  geom_point(aes(color = site)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(method = "lm", color = "black", fill = "gray", alpha = 0.2, se = TRUE) +
  xlab("log10 Post-thaw copy number g⁻¹ soil") +
  ylab("log10 Cumulative Respiration (µg C-CO₂ g⁻¹ dry soil)") +
  theme(text = element_text(size = 12)) +
  scale_color_discrete("Site") +
  theme_classic()

post.plot <- ggplot(post_respcpnumb, aes(x = copy_numb_gsoil, y = Cumulative_Respiration)) + 
  geom_point(aes(color = site)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(method = "lm", color = "black", fill = "gray", alpha = 0.2, se = TRUE) +
  labs(
    x = expression("log"[10] * " Post-thaw copy number g"^{-1} * " soil"),
    y = expression("log"[10] * " Cumulative Respiration (" * mu * "g C-CO"[2] * " g"^{-1} * " dry soil)")
  ) +
  scale_color_manual(
    values = c("#1b9e77", "#d95f02", "#7570b3"),
    name = "Site"
  ) +
  theme(text = element_text(size = 12)) +
  theme_classic()

```
3. Bringing the two plots together
```{r}
# Facet 
library(patchwork)
# Combine plots
combined <- pre.plot + post.plot

# Display combined
combined

# Trying another way 
install.packages("gridextra")
library(gridExtra)


```




```{r}
respcpnumb <- factor(respcpnumb, levels = rev(levels(respcpnumb)))
# Hannah's code to visualize both pre and post as a response to respiration: 
pivot_longer(respcpnumb, cols = contains("copy_number"), names_to = "Pre_post_copynumber", values_to = "copy_number") %>% 
  ggplot(aes(respcpnumb, x = copy_number, y = Cumulative_Respiration, shape = Pre_post_copynumber, group = Pre_post_copynumber, color = site)) + 
           geom_point(size = 2.5) + 
  geom_smooth(method = "lm", se = TRUE, aes(color = Pre_post_copynumber)) + # okay this works but its coming out weird in the legend 
            scale_x_log10() +
            #scale_y_log10() +
            theme_classic() +
            xlab("log10 Copy number g-1 soil") +
            ylab("Cumulative respiration (µg C-CO2 g-1 dry soil)") +
            theme(text = element_text(size = 12)) +
            scale_shape_discrete("Copy number type", labels = c("Post-thaw", "Pre-thaw")) + #this line changes the name of the shape but doesnt take the data with it 
            scale_color_discrete("Site")
            # geom_smooth(aes(color = "pre_post_thaw")) trying to change the color of the regression lines
            # guide = guide_legend(reverse = TRUE)

```

```{r}
respcpnumb <- merge(pre_respcpnumb, post_respcpnumb, by = "Sample_ID")

library(dplyr)

# Suppose pre_respcpnumb and post_respcpnumb have these columns:
# Sample_ID, site, core, pre_post_thaw, copy_numb_gsoil, Cumulative_Respiration

# First join by Sample_ID, but exclude the columns you don't want duplicated

respcnumb_delta <- full_join(
  pre_respcpnumb %>% select(-pre_post_thaw),  # remove pre_post_thaw from pre dataset
  post_respcpnumb %>% select(-pre_post_thaw), # remove pre_post_thaw from post dataset
  by = c("Sample_ID", "site", "core", "Cumulative_Respiration"),
  suffix = c("_pre", "_post")  # keep suffixes only for differing columns
)


library(dplyr)

respcpnumb_change <- respcnumb_delta %>%
  mutate(
    # Calculate fold change (post / pre), avoid divide-by-zero issues with a small pseudocount
    fold_change = (copy_numb_gsoil_post + 1) / (copy_numb_gsoil_pre + 1),
    log2_fold_change = log2(fold_change),
    
    # Optional: difference in copy number
    diff_copy_number = copy_numb_gsoil_post - copy_numb_gsoil_pre
  )

# For not log2
library(dplyr)

respcpnumb_change_sub <- respcnumb_delta %>%
  mutate(
    # Difference in copy number (post - pre)
    diff_copy_number = copy_numb_gsoil_post - copy_numb_gsoil_pre
  )


library(ggplot2)

ggplot(respcpnumb_change_sub, aes(x = log2_fold_change, y = Cumulative_Respiration, color = site)) +
  #geom_point(size = 3, alpha = 0.9) +
  geom_smooth(method = "lm", se = TRUE, aes(color = site)) +
   geom_point(size = 3, alpha = 0.9) +
  scale_y_log10() +  # log scale respiration
  labs(
    x = "Log2 Fold Change in 16S Copy Number (" g"^{-1} * " dry soil)",
    y = expression("Cumulative Respiration (" * mu * "g C-CO"[2] * " g"^{-1} * " dry soil)",
    color = "Site") +
  scale_color_manual(values = c("#1b9e77", "#d95f02", "#7570b3")) +  # FIXED: added missing parenthesis
  theme_classic() +
  theme(text = element_text(size = 12))
```

```{r}
ggplot(respcpnumb_change_sub, aes(x = diff_copy_number, y = Cumulative_Respiration, color = site)) +
  geom_smooth(method = "lm", se = TRUE, aes(color = site)) +
  geom_point(size = 3, alpha = 0.9) +
  scale_y_log10() +# log scale respiration
  scale_x_log10() +
  labs(
    x = expression("Log2 Fold Change in 16S Copy Number (g"^{-1}~"dry soil)"),
    y = expression("Cumulative Respiration (" * mu * "g C-CO"[2] * " g"^{-1} * " dry soil)"),
    color = "Site"
  ) +
  scale_color_manual(values = c("#1b9e77", "#d95f02", "#7570b3")) +
  theme_classic() +
  theme(text = element_text(size = 12))

```


Stats
```{r}
# Is the relationship between the change in biomass and cumulative respiration significant? 

# For each site separately 
library(dplyr)
library(dplyr)

# Separate linear models per site
respcpnumb_change_sub %>%
  group_by(site) %>%
  do(model = summary(lm(log10(Cumulative_Respiration) ~ diff_copy_number, data = .)))

model_full <- lm(log10(Cumulative_Respiration) ~ diff_copy_number + site, data = respcpnumb_change)
summary(model_full)
anova(model_full)


```
```{r}
lm(log10(Cumulative_Respiration) ~ log2_fold_change + site)

```


```{r}
ancova_model <- aov(log10(Cumulative_Respiration) ~ log2_fold_change * site, data = respcpnumb_change)
summary(ancova_model)

respcpnumb_change %>%
  group_by(site) %>%
  summarize(r_squared = summary(lm(log10(Cumulative_Respiration) ~ log2_fold_change))$r.squared)

plot(model_full)
```


Now for stats on this 
```{r}
#linearmodel_respcopypost <- glm(copy_number_per_gsoil_POST ~ Cumulative_Respiration, family=gaussian, data = respcpnumb)
lm_change <- lm(Cumulative_Respiration ~ log2_fold_change + site + diff_copy_number:site, data = respcpnumb_change)
summary(lm_change)
AIC(lm_change)
# What is the r2 and p-value for the whole model, not split by site
lm_whole <- lm(Cumulative_Respiration ~ log2_fold_change, data = respcpnumb_change)
summary(lm_whole)
AIC(lm_whole)

library(rsq)
rsq.partial(lm_change)
```

```{r}
library(dplyr)
library(ggplot2)

# Combine the two datasets by row (assuming both have the same columns)
combined_respcpnumb <- bind_rows(pre_respcpnumb, post_respcpnumb)

# Check combined data
head(combined_respcpnumb)

# Plot with pre_post_thaw as a shape/color grouping factor
ggplot(combined_respcpnumb, aes(x = copy_numb_gsoil, y = Cumulative_Respiration, color = site, shape = pre_post_thaw)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, aes(linetype = pre_post_thaw, color = site)) +
  scale_x_log10() +
  scale_y_log10() +
  labs(
    x = expression("log"[10] * " Copy number g"^{-1} * " soil"),
    y = expression("log"[10] * " Cumulative Respiration (" * mu * "g C-CO"[2] * " g"^{-1} * " dry soil)"),
    color = "Site",
    shape = "Pre/Post Thaw",
    linetype = "Pre/Post Thaw"
  ) +
  scale_color_manual(values = c("#1b9e77", "#d95f02", "#7570b3")) +
  theme_classic() +
  theme(text = element_text(size = 12))

```
# Plotting for publication
```{r}

combined_respcpnumb$pre_post_thaw <- factor(combined_respcpnumb$pre_post_thaw, levels = c("pre", "post"))

ggplot(combined_respcpnumb, aes(x = copy_numb_gsoil, y = Cumulative_Respiration, color = site, shape = pre_post_thaw)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_smooth(
    method = "lm", 
    se = TRUE, 
    aes(group = pre_post_thaw, linetype = pre_post_thaw),  # map linetype properly
    color = "black"  # this will be overridden by linetype, so you can remove this or keep it for black lines
  ) +
  scale_linetype_manual(values = c("dashed", "solid")) +  # assign dashed for pre, solid for post
  scale_x_log10() +
  scale_y_log10() +
  labs(
    x = expression("log"[10] * " Copy number g"^{-1} * " soil"),
    y = expression("log"[10] * " Cumulative Respiration (" * mu * "g C-CO"[2] * " g"^{-1} * " dry soil)"),
    color = "Site",
    shape = "Copy number type",
    linetype = "Copy number type"
  ) +
  scale_color_manual(values = c("#1b9e77", "#d95f02", "#7570b3")) +
  theme_classic() +
  theme(text = element_text(size = 12))


```
```{r}
ggplot(combined_respcpnumb, aes(x = copy_numb_gsoil, y = Cumulative_Respiration, color = site, shape = pre_post_thaw)) +
  geom_smooth(
    method = "lm", 
    se = TRUE, 
    aes(group = pre_post_thaw, linetype = pre_post_thaw),
    color = "black"
  ) +
  geom_point(size = 3, alpha = 0.8, stroke = 1) +
  scale_shape_manual(
    values = c("pre" = 1, "post" = 16)  # hollow circle for pre, filled for post
  ) +
  scale_linetype_manual(
    values = c("pre" = "dotted", "post" = "solid")
  ) +
  scale_x_log10() +
  scale_y_log10() +
  labs(
    x = expression("log"[10] * " Copy number g"^{-1} * " soil"),
    y = expression("log"[10] * " Cumulative Respiration (" * mu * "g C-CO"[2] * " g"^{-1} * " dry soil)"),
    color = "Site",
    shape = "Copy number type",
    linetype = "Copy number type"
  ) +
  scale_color_manual(values = c("#1b9e77", "#d95f02", "#7570b3")) +
  theme_classic() +
  theme(text = element_text(size = 12))

```

Now adding inset -- this is not the vibe
```{r}
library(ggplot2)
library(patchwork)

# 1. Main plot (you already have this)
main_plot <- ggplot(combined_respcpnumb, aes(x = copy_numb_gsoil, y = Cumulative_Respiration, color = site, shape = pre_post_thaw)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_smooth(
    method = "lm", 
    se = TRUE, 
    aes(group = pre_post_thaw, linetype = pre_post_thaw),
    color = "black"
  ) +
  scale_linetype_manual(values = c("dashed", "solid")) +
  scale_x_log10() +
  scale_y_log10() +
  labs(
    x = expression("log"[10] * " Copy number g"^{-1} * " soil"),
    y = expression("log"[10] * " Cumulative Respiration (" * mu * "g C-CO"[2] * " g"^{-1} * " dry soil)"),
    color = "Site",
    shape = "Copy number type",
    linetype = "Copy number type"
  ) +
  scale_color_manual(values = c("#1b9e77", "#d95f02", "#7570b3")) +
  theme_classic() +
  theme(text = element_text(size = 12))

# 2. Pre-thaw-only plot
pre_plot <- combined_respcpnumb |>
  dplyr::filter(pre_post_thaw == "pre") |>
  ggplot(aes(x = copy_numb_gsoil, y = Cumulative_Respiration, color = site)) +
  geom_point(size = 1.8) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linetype = "dashed") +
  scale_x_log10() +
  scale_y_log10() +
  #theme_void() + 
  theme(legend.position = "none")

# 3. Post-thaw-only plot
post_plot <- combined_respcpnumb |>
  dplyr::filter(pre_post_thaw == "post") |>
  ggplot(aes(x = copy_numb_gsoil, y = Cumulative_Respiration, color = site)) +
  geom_point(size = 1.8) +
  geom_smooth(method = "lm", se = TRUE, color = "black", linetype = "solid") +
  scale_x_log10() +
  scale_y_log10() +
  #theme_void() +
  theme(legend.position = "none")

# 4. Add insets
final_plot <- main_plot +
  inset_element(pre_plot, left = 0.05, bottom = 0.6, right = 0.4, top = 0.95) +
  inset_element(post_plot, left = 0.6, bottom = 0.05, right = 0.95, top = 0.4)

# 5. Print
final_plot

```


```{r}
library(tidyverse)
library(ggplot2)

# Assuming your data frame is called 'respcpnumb' and has columns like:
# - Cumulative_Respiration
# - site
# - copy_number_pre_thaw
# - copy_number_post_thaw

# First pivot longer the pre/post copy number columns to one column, with type info
respcpnumb_long <- respcpnumb %>%
  pivot_longer(
    cols = c(copy_number_pre_thaw, copy_number_post_thaw),
    names_to = "Pre_post_copynumber",
    values_to = "copy_number"
  ) %>%
  # Optional: clean up the Pre_post_copynumber labels for nicer legend
  mutate(Pre_post_copynumber = recode(Pre_post_copynumber,
                                     copy_number_pre_thaw = "Pre-thaw",
                                     copy_number_post_thaw = "Post-thaw"))

# Plot
ggplot(respcpnumb_long, aes(x = copy_number, y = Cumulative_Respiration, 
                           color = site, shape = Pre_post_copynumber)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, aes(linetype = Pre_post_copynumber, color = site)) +
  scale_x_log10() +
  scale_y_log10() +  # optional, comment out if you don't want log y-axis
  labs(
    x = expression("log"[10] * " Copy number g"^{-1} * " soil"),
    y = expression("log"[10] * " Cumulative Respiration (" * mu * "g C-CO"[2] * " g"^{-1} * " dry soil)"),
    color = "Site",
    shape = "Copy Number Type",
    linetype = "Copy Number Type"
  ) +
  scale_color_manual(values = c("#1b9e77", "#d95f02", "#7570b3")) +
  theme_classic() +
  theme(text = element_text(size = 12))

```


Okay now we will look to see if change in copy number has any correlation with cumulative respiration
```{r}
library(dplyr)

respcpnumb_change <- respcpnumb %>%
  mutate(
    # Calculate fold change (post / pre), avoid divide-by-zero issues with a small pseudocount
    fold_change = (copy_number_post_thaw + 1) / (copy_number_pre_thaw + 1),
    log2_fold_change = log2(fold_change),
    
    # Optional: difference in copy number
    diff_copy_number = copy_number_post_thaw - copy_number_pre_thaw
  )
```
















```{r}
# ANALYZING RESPIRATION AND PRE THAW COPY NUMBERS HERE 


# Look at the density of cumulative respiration
# plot(density(resp$Cumulative_Respiration), main="Cumulative_Respiration", ylab="Frequency", 
     # sub=paste("Skewness:", round(e1071::skewness(resp$Cumulative_Respiration), 3)))
# polygon(density(resp$Cumulative_Respiration), col="darkorange")

# Look at the density of copy numbers pre
# plot(density(cpnumb$copy_number_per_gsoil_PRE), main="copy_number_per_gsoil_PRE", ylab="Frequency", 
     # sub=paste("Skewness:", round(e1071::skewness(cpnumb$copy_number_per_gsoil_PRE), 3)))

# polygon(density(cpnumb$copy_number_per_gsoil_PRE), col="darkorange")

# Merge the data frames
respcpnumb <- merge(resp4copy,cpnumb) 

# Okay let's run a Kruskall Wallis and then a dunn test (maybe) for cum resp and pre-thaw copy numbers
kruskal.test(copy_number_per_gsoil_PRE ~ Cumulative_Respiration, data = respcpnumb) # Kruskal-Wallis chi-squared = 21, df = 21, p-value = 0.4589
kruskal.test(copy_number_per_gsoil_PRE ~ site, data = cpnumb) # not significant p = 0.2062

kruskal.test(copy_number_per_gsoil_POST ~ site, data = cpnumb) #significant p-value = 0.03
dunn_test(copy_number_per_gsoil_POST ~ site, data = cpnumb)
kruskal.test(copy_number_per_gsoil_PRE ~ copy_number_per_gsoil_POST, data = cpnumb) # p-value = 0.4599
```

```{r}
# Okay let's run a Kruskall Wallis for cum respiration and post thaw copy numbers
kruskal.test(copy_number_per_gsoil_POST ~ Cumulative_Respiration, data = respcpnumb) # Kruskal-Wallis chi-squared = 21, df = 21, p-value = 0.4589 SAME P VALUE?

# FIRST LET'S RUN A CORRELATION FOR RESPIRATION ~ PRE THAW COPY NUMBER 
resp_pre_corr <- cor.test(x =respcpnumb$copy_number_per_gsoil_PRE, y = respcpnumb$Cumulative_Respiration, method = "spearman")
print(resp_pre_corr)

# Visualize here 
ggplot(respcpnumb, aes(x = copy_number_per_gsoil_PRE, y = Cumulative_Respiration)) + 
  geom_point(aes(color = site)) +
  #scale_x_log10() +
  #scale_y_log10() +
  geom_smooth(method = "lm")

respcpnumb <- factor(respcpnumb, levels = rev(levels(respcpnumb)))
# Hannah's code to visualize both pre and post as a response to respiration: 
pivot_longer(respcpnumb, cols = contains("copy_number"), names_to = "Pre_post_copynumber", values_to = "copy_number") %>% 
  ggplot(aes(respcpnumb, x = copy_number, y = Cumulative_Respiration, shape = Pre_post_copynumber, group = Pre_post_copynumber, color = site)) + 
           geom_point(size = 2.5) + 
  geom_smooth(method = "lm", se = TRUE, aes(color = Pre_post_copynumber)) + # okay this works but its coming out weird in the legend 
            scale_x_log10() +
            #scale_y_log10() +
            theme_classic() +
            xlab("log10 Copy number g-1 soil") +
            ylab("Cumulative respiration (µg C-CO2 g-1 dry soil)") +
            theme(text = element_text(size = 12)) +
            scale_shape_discrete("Copy number type", labels = c("Post-thaw", "Pre-thaw")) + #this line changes the name of the shape but doesnt take the data with it 
            scale_color_discrete("Site")
            # geom_smooth(aes(color = "pre_post_thaw")) trying to change the color of the regression lines
            # guide = guide_legend(reverse = TRUE)

```

```{r}
# NEXT LET'S RUN A CORRELATION FOR RESPIRATION ~ POST THAW COPY NUMBER 
resp_post_corr <- cor.test(x = respcpnumb$copy_number_per_gsoil_POST, y = respcpnumb$Cumulative_Respiration, method = "spearman")
print(resp_post_corr)

# Visualize 
ggplot(respcpnumb, aes(x =copy_number_per_gsoil_POST, y = Cumulative_Respiration)) + 
  geom_point(aes(color = site)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(method = "lm")

# Cleaning up the figure for publication 
ggplot(respcpnumb, aes(x = copy_number_per_gsoil_POST, y = Cumulative_Respiration)) + 
  geom_point(aes(color = site), size = 2.5) +
  theme_classic() +
  scale_x_log10() +
  scale_y_log10() +
  xlab("log10 Post-thaw copy number g-1 soil") +
  ylab("log10 Cumulative Respiration (µg C-CO2 g-1 dry soil)") +
  geom_smooth(method = "lm", color = "dark gray", se=TRUE) +
  theme(text = element_text(size = 12)) +
  scale_color_discrete("Site")
```

```{r}
# NOW WE WILL RUN KRUSKALL WALLIS ON BACTERIAL ABUMDANCE PRE THAW BY SITE 
kruskal.test(copy_number_per_gsoil_PRE ~ site, data = respcpnumb) # chi-squared = 3.1581, df = 2, p-value = 0.2062

# AND WE WILL RUN KRUSKALL ON BACTERIAL ABUNDANCE POST THAW BY SITE
kruskal.test(copy_number_per_gsoil_POST ~ site, data = respcpnumb) # significant: Kruskal-Wallis chi-squared = 6.7302, df = 2, p-value = 0.03456
dunnTest(copy_number_per_gsoil_POST ~ site, data = respcpnumb, method = 'bh') # not sure if I should be using bh or holms? 

# Now let's look to see if there is any significance pre and post thaw by site for copy numbers 
kruskal.test(copy_number_per_gsoil_PRE ~ copy_number_per_gsoil_POST, data = respcpnumb) # NOT SIGNIF Kruskal-Wallis chi-squared = 21, df = 21, p-value = 0.4589

```

```{r}
# *************************************************************************************************************************************************************
# Creating the glm THIS NEEDS TO BE EDITED
linearmodel_respcopypre <- glm(copy_number_per_gsoil_PRE ~ Cumulative_Respiration, family=gaussian, data = respcpnumb)

# Look at the results of the glm
print(linearmodel_respcopypre)
summary(linearmodel_respcopypre)

# Visualize with ggplot  ## Okay this is fine, BUT I want to show this by site so theoretically make 3 plots for pre and 3 for post 
ggplot(linearmodel_respcopypre, aes(x = copy_number_per_gsoil_PRE, y = Cumulative_Respiration)) +
  geom_point (aes(color = linearmodel_respcopypre[["data"]][["site"]])) + 
  #scale_x_log10() +
  #scale_y_log10() +
  geom_smooth(method = "lm")

#**************************************************************************

# ANALYZING POST-THAW COPY NUMBERS AND RESPIRATION

# Visualizing the data to start off 
scatter.smooth(x = cpnumb$copy_number_per_gsoil_POST, y = resp4copy$Cumulative_Respiration)

# Checking for normality in the post thaw copy number values 
shapiro.test(cpnumb$copy_number_per_gsoil_POST) #p-value = 5.065e-06, not normal

# We need to check if the dependent variable is close to normal (?)
par(mfrow=c(1, 2)) # divide graph area in 2 columns

# Look at the density of cumulative respiration
plot(density(resp$Cumulative_Respiration), main="Cumulative_Respiration", ylab="Frequency", 
     sub=paste("Skewness:", round(e1071::skewness(resp$Cumulative_Respiration), 3)))
polygon(density(resp$Cumulative_Respiration), col="darkorange")

# Look at the density of copy numbers pre
plot(density(cpnumb$copy_number_per_gsoil_POST), main="copy_number_per_gsoil_POST", ylab="Frequency", 
     sub=paste("Skewness:", round(e1071::skewness(cpnumb$copy_number_per_gsoil_POST), 3)))

polygon(density(cpnumb$copy_number_per_gsoil_PRE), col="darkorange")

# Run the glm, this needs to be edited
linearmodel_respcopypost <- glm(copy_number_per_gsoil_POST ~ Cumulative_Respiration, family=gaussian, data = respcpnumb)

# Look at the results of the glm
print(linearmodel_respcopypost)
summary(linearmodel_respcopypost)

# Visualize with ggplot # HAVE TO GET THIS BY SITE AGAIN, IT'S FREAKING OUT OTHERWISE 
ggplot(linearmodel_respcopypost, aes(x = copy_number_per_gsoil_POST, y = Cumulative_Respiration)) +
  geom_point(aes(color = linearmodel_respcopypost[["data"]][["site"]])) + 
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(method = "lm")

#*****************************************************************************

# Spearman correlation for copy numbers post thaw and copy numbers pre thaw

# First let's look at pre-thaw

# Make a QQ plot
qqnorm(x = copynumb$copy_number_per_gsoil_PRE, y = resp4copy$Cumulative_Respiration, main = "Q-Q Plot")

# Run the corr test 
copyprecorr <- cor.test(x = copynumb$copy_number_per_gsoil_PRE, y = resp4copy$Cumulative_Respiration, method = "spearman") #p value = 0.6012, rho = 0.1146245
print(copyprecorr)

precopyresp<- merge(copynumb, resp4copy)

# Visualize
ggplot(precopyresp, aes(x = copy_number_per_gsoil_PRE, y = Cumulative_Respiration)) + 
  geom_point(aes(color = site, shape = core)) 
scale_x_log10() +
  scale_y_log10()

# Now we will look at copy number post thaw 

# Visualize a QQ plot first
qqnorm(x = copynumb$copy_number_per_gsoil_POST, y = resp4copy$Cumulative_Respiration, main = "Q-Q Plot")

copypostcorr <- cor.test(x = copynumb$copy_number_per_gsoil_POST, y = resp4copy$Cumulative_Respiration, method = "spearman") # p-value = 0.001604, rho = 0.6304348
print(copypostcorr)

# Visualize
ggplot(precopyresp, aes(x = copy_number_per_gsoil_POST, y = Cumulative_Respiration)) + 
  geom_point(aes(color = site, shape = core)) +
  scale_x_log10() +
  scale_y_log10()

# END
```

