---
title: "Thaw responder analysis"
author: "Joy O'Brien"
date: "2025-05-02"
output: html_document
---
Need to fix/know: Are we working with C.ASVs? 

Thaw response code (response ratio analysis code from Hannah Holland-Moritz, modeled and modified from the following publication: Fan, K., Holland-Moritz, H., Walsh, C., Guo, X., Wang, D., Bai, Y., Zhu, Y., Fierer, N. & Chu, H. (2022). 
# Identification of the rhizosphere microbes that actively consume plant-derived carbon. Soil Biology and Biochemistry, 166, 108577. 
# https://doi.org/10.1016/j.soilbio.2022.108577)
1. Set working directory
```{r setup, include=FALSE}
setwd("~/GitHub/thaw.response")
```
2. Load libraries
```{r}
# Load the necessary libraries
#install.packages("mctoolsr")
#library(mctoolsr)
library(ggplot2)
library(tibble)
library(dplyr)
library(multcompView)
library(phyloseq)
library(readxl)
library(biomformat)
library(readr)
library(readxl)
library(ARPobservation)
```

Employing Hannah & Jessica's feedback:
- Make this analysis absolute abundance
- Take each of the sites--dataframes from each site, stack them together, run code on larger frame (site columm) - top 100 thaw responders from each site, bind rows together
- Plots will have 3 points for every organism, 1 point represents the response from one organism in one site
- Color = site , position dodge by site


Edit: Calculating absolute abundance
3. Calculating absolute abundance (copying from script 4)
```{r}
thaw <- readRDS("~/GitHub/thaw.response/Data/thaw_phyloseq.rds")
cpnumb <- read_excel("~/GitHub/thaw.response/Data/16Scopy_number.xlsx", sheet = "16S_copy")

# Extract raw OTU count table and make a matrix
otu_table <- as.data.frame(thaw@otu_table)
otu_mat <- as.matrix(otu_table)

# Extract qPCR copy numbers into a named vector
qPCR_vec <- setNames(cpnumb$copy_numb_gsoil, cpnumb$Sample_ID)

# Match the sample order to OTU matrix columns
qPCR_vec <- qPCR_vec[colnames(otu_mat)]

# Calculate total sequencing reads per sample
total_reads <- colSums(otu_mat)

# Calculate scaling factors
scaling_factors <- qPCR_vec / total_reads

# Multiply each column (sample) in the OTU matrix by its scaling factor
otu_abs_abund <- sweep(otu_mat, 2, scaling_factors, FUN = "*") # we get some NAs here because samples are missing 
# we do not get nas here
```

3A. Checking out the samples that have NA
```{r}
# Checking the potential error / mismatch 
# Sample names in OTU table
otu_samples <- colnames(otu_mat)

# Sample names in qPCR data
qpcr_samples <- cpnumb$Sample_ID

# Find which OTU samples are missing in qPCR
missing_in_qpcr <- otu_samples[!otu_samples %in% qpcr_samples]

# Also check the reverse: which qPCR samples aren't in OTU table
missing_in_otu <- qpcr_samples[!qpcr_samples %in% otu_samples]

# Print to investigate
missing_in_qpcr
missing_in_otu
```

```{r}
# # scaling_factors came from: qPCR_vec / total_reads
# na_samples <- names(scaling_factors)[is.na(scaling_factors)]
# na_samples
# 
# # Removing NA samples from all dataframes
# otu_mat      <- otu_mat[ , !colnames(otu_mat)      %in% na_samples]
# total_reads  <- total_reads[ !names(total_reads)   %in% na_samples]
# qPCR_vec     <- qPCR_vec[  !names(qPCR_vec)        %in% na_samples]
# 
# scaling_factors <- qPCR_vec / total_reads          # now no NAs

otu_abs_abund   <- sweep(otu_mat, 2, scaling_factors, FUN = "*")

# Obtain sample names
names(qPCR_vec)
names(total_reads)  

# Align by intersecting names
shared_samples <- intersect(names(qPCR_vec), names(total_reads))

# Subset both to shared samples
qPCR_vec      <- qPCR_vec[shared_samples]
total_reads   <- total_reads[shared_samples]

# Now calculate scaling factors
scaling_factors <- qPCR_vec / total_reads

length(scaling_factors)
any(is.na(scaling_factors))  # should be FALSE

# Samples in scaling_factors should match the columns in otu_mat
otu_mat_subset <- otu_mat[, names(scaling_factors)]

# Multiply each sample column by its corresponding scaling factor
absolute_otu <- sweep(otu_mat_subset, 2, scaling_factors, `*`)
# Now we have absolute OTU without NA

```



4. Make df from phyloseq object
```{r}
# Load data
#thaw <- read_rds("~/GitHub/thaw.response/Data/thaw_phyloseq.rds")
thaw.meta <- read_rds("~/GitHub/thaw.response/Data/thaw_metadata_phyloseq.rds")
thaw_OTU_phyloseq <- read_rds("~/GitHub/thaw.response/Data/thaw_OTU_phyloseq.rds")
thaw_tax <- read_rds("~/GitHub/thaw.response/Data/thaw_tax_phyloseq.rds")
thaw_response <- thaw_OTU_phyloseq %>%
  data.frame()

# We are going to scale each OTU count up by 0.01 so that 0 becomes 0.01 and the relationship stays the same
#thaw_response <- absolute_otu #+ 0.01 
thaw_response <- otu_abs_abund + 0.01

# Change the absolute abundance df to otu table object 
absolute_matrix <- as.matrix(absolute_otu)

# Convert to an otu_table object
otu_abs <- otu_table(absolute_matrix, taxa_are_rows = TRUE)  # Or FALSE if samples are rows

# Incorporating this back into phyloseq object 
thaw.abs <- phyloseq(otu_abs,thaw.meta,thaw_tax)
#abs_OTU_phyloseq <- thaw.abs@otu_table
#as.data.frame(abs_OTU_phyloseq)

# Create thaw_response based on new abs table
#thaw.response.abs <- abs_OTU_phyloseq %>%
 # data.frame()


#thaw.response.abs <- thaw.abs@otu_table
#thaw.response.abs <- as.data.frame(thaw.response.abs) # COME BACK AND PICK UP HERE I DONT KNOW WHY THERE IS CONFUSION
```

3A. Assessing thaw response (all 3 sites combined and all responders)
```{r}
# dont know why i had to remove samples 41 and 42 from the line of number code below 
ToF <- thaw_response; #changed from thaw_response to thaw.abs
ToF[ToF > 0] <- 1 #anything above 0 is 1 (Presence absence table) ##

# Which columns are post and which are pre (for the controls and treatment)
pre_col_indices <- grep("PRE", names(thaw_response))
post_col_indices <- grep("POST", names(thaw_response))

pre_col_indices
post_col_indices

# Using relative abundance table, creates average
# treatment_names <- c("sample1", "sample2"...)
ave1 = apply(thaw_response[,c(4,6,8,10,12,14,16,18,20,22,24,26,28,30,32,34,35,37,39,41,43)],1,mean,na.rm = TRUE);#Control
ave2 = apply(thaw_response[,c(1,2,3,5,7,9,11,13,15,17,19,21,23,25,27,29,31,33,36,38,40,42)],1,mean,na.rm = TRUE)# Treatment 

# Standard Deviation of ALL samples
sd1 = apply(thaw_response[,c(4,6,8,10,12,14,16,18,20,22,24,26,28,30,32,34,35,37,39,41,43)],1,sd,na.rm = TRUE);# Control #
sd2 = apply(thaw_response[,c(1,2,3,5,7,9,11,13,15,17,19,21,23,25,27,29,31,33,36,38,40,42)],1,sd,na.rm = TRUE)# treatment # 

# How many encounters for ASV number of observations
num1 = apply(ToF[,c(4,6,8,10,12,14,16,18,20,22,24,26,28,30,32,34,35,37,39,41,43)],1,sum,na.rm = TRUE); #Control#
num2 = apply(ToF[,c(1,2,3,5,7,9,11,13,15,17,19,21,23,25,27,29,31,33,36,38,40,42)],1,sum,na.rm = TRUE) # Treatment #
```

3B. Response ratio calculations (all sites)
```{r}
# Response ratio calculations
# Sample variance
v1 = sd1^2/(ave1^2*num1); # Control #
v2 = sd2^2/(ave2^2*num2) # treatment #

# Square root of variance 
sqv = sqrt(v1+v2)

# RR = Response ratio
RR = log(ave2/ave1)
```

3C. Confidence interval (all sites)
```{r}
# Confidence interval

ICl_R = RR-1.96*sqv
ICu_R = RR+1.96*sqv

# Error 
err_R = ICu_R-RR

# Significance
Sig_R = ICl_R*ICu_R
Sig_adjust_R <- p.adjust(Sig_R, method = "fdr") 

# Creating data frame w/ 3 coloums (RR, err, Sig) 
R = as.data.frame(cbind(RR,err_R,Sig_R, ICl_R, ICu_R))

R_adjust <- as.data.frame(cbind(RR,err_R,Sig_adjust_R, ICl_R, ICu_R))
# Create a new column from row names 
R$ID <- rownames(R)

#selecting rows (ASV's) with significance & values #
R$Significance <- ifelse(ICl_R > 0 & ICu_R > 0, "Up",
                   ifelse(ICl_R < 0 & ICu_R < 0, "Down", "Not Significant"))

sig_ASVs <- R[R$Significance != "Not Significant", ]

# Adjust for multiple comparisons 
z_scores <- RR / sqv
p_vals <- 2 * pnorm(-abs(z_scores))  # two-tailed test
p_adj <- p.adjust(p_vals, method = "fdr")

R$p_value <- p_vals
R$p_adj <- p_adj
R$Significant_FDR <- p_adj < 0.05

thaw.tax <- as.data.frame(thaw_tax)

tax_response <- left_join(R, thaw.tax %>% mutate(ID = rownames(thaw.tax)), by = "ID")

tax_sig <- tax_response %>% filter(Significance != "Not Significant")

tax_sig_top <- tax_sig %>%
  arrange(desc(abs(RR))) %>%
  slice(1:100)

ggplot(tax_sig_top, aes(x = reorder(ID, RR), y = RR, color = Phylum)) + 
  geom_pointrange(aes(ymin = ICl_R, ymax = ICu_R)) +
  geom_hline(yintercept = 0, linetype = 2) +
  facet_wrap(~Phylum, scales = "free_x") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 70, hjust = 1)) +
  labs(x = "ASV ID", y = "Log Response Ratio (Thaw/Post vs Pre)")



```
4. Significant thaw responders only (from all sites)
```{r}
#Significant thaw responders only 
tax_sig_up <- tax_response %>%
  filter(Significance == "Up")

top_up <- tax_sig_up %>%
  arrange(desc(RR)) %>%
  slice(1:100)  # change 20 to however many you want

ggplot(top_up, aes(x = reorder(ID, RR), y = RR, color = Phylum)) + 
  geom_pointrange(aes(ymin = ICl_R, ymax = ICu_R)) +
  geom_hline(yintercept = 0, linetype = 2) +
  facet_wrap(~Phylum, scales = "free_x") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 70, hjust = 1)) +
  labs(x = "ASV ID", y = "Log Response Ratio (Thaw/Post vs Pre)")

# Plotting to show the genus
#ggplot(top_up, aes(x = reorder(Class, RR), y = RR, color = Phylum)) + 
  #geom_pointrange(aes(ymin = ICl_R, ymax = ICu_R)) +
  #geom_hline(yintercept = 0, linetype = 2) +
  #facet_wrap(~Phylum, scales = "free_x") +
  #theme_bw() +
  #theme(axis.text.x = element_text(angle = 70, hjust = 1)) +
  #labs(x = "Class", y = "Log Response Ratio (Thaw/Post vs Pre)") 
  


# # Significant thaw decliners
# tax_sig_down <- tax_response %>%
#   filter(Significance == "Down")
# 
# down_up <- tax_sig_down %>%
#   arrange(desc(RR)) %>%
#   slice(1:20)  # change 20 to however many you want
# 
# ggplot(down_up, aes(x = reorder(ID, RR), y = RR, color = Phylum)) + 
#   geom_pointrange(aes(ymin = ICl_R, ymax = ICu_R)) +
#   geom_hline(yintercept = 0, linetype = 2) +
#   facet_wrap(~Phylum, scales = "free_x") +
#   theme_bw() +
#   theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
#   labs(x = "ASV ID", y = "Log Response Ratio (Thaw/Post vs Pre)")
```
5. Thaw response assessed by site (CRREL)
```{r}
# Subset the data out for CRREL sites
thaw_response <- as.data.frame(thaw_response)
t.r.crrel <- as.data.frame(thaw_response %>% select(matches("AT3|AT4")))
ToF_crrel <- t.r.crrel;
ToF_crrel[ToF_crrel > 0] <- 1 #anything above 0 is 1 (Presence absence table) ##

# Which columns are post and which are pre (for the controls and treatment)
pre_col_indices_crrel <- grep("PRE", names(t.r.crrel))
post_col_indices_crrel <- grep("POST", names(t.r.crrel))

pre_col_indices_crrel
post_col_indices_crrel

grep("PRE", colnames(t.r.crrel))
grep("POST", colnames(t.r.crrel))


# Using relative abundance table, creates average
# treatment_names <- c("sample1", "sample2"...)
ave1_crrel = apply(t.r.crrel[,c(4, 6, 8, 10, 12)],1,mean,na.rm = TRUE);# Control
ave2_crrel = apply(t.r.crrel[,c(1, 2, 3, 5, 7, 9, 11)],1,mean,na.rm = TRUE)# Treatment 

# Standard Deviation of ALL samples
sd1_crrel = apply(t.r.crrel[,c(4,6,8,10,12)],1,sd,na.rm = TRUE);# Control #
sd2_crrel = apply(t.r.crrel[,c(1, 2, 3, 5, 7, 9, 11)],1,sd,na.rm = TRUE)# treatment # 

# How many encounters for ASV number of observations
num1_crrel= apply(ToF_crrel[,c(4, 6, 8, 10, 12)],1,sum,na.rm = TRUE); #Control#
num2_crrel = apply(ToF_crrel[,c(1, 2, 3, 5, 7, 9, 11)],1,sum,na.rm = TRUE) # Treatment #

# Response ratio calculations
# Sample variance
v1_crrel = sd1_crrel^2/(ave1_crrel^2*num1_crrel); # Control 
v2_crrel= sd2_crrel^2/(ave2_crrel^2*num2_crrel) # treatment

# Square root of variance 
sqv_crrel = sqrt(v1_crrel+v2_crrel)

# RR = Response ratio
RR_crrel = log(ave2_crrel/ave1_crrel)

# Confidence interval

ICl_R_crrel = RR_crrel-1.96*sqv_crrel
ICu_R_crrel = RR_crrel+1.96*sqv_crrel

# Error 
err_R_crrel = ICu_R_crrel-RR_crrel

# Significance
Sig_R_crrel = ICl_R_crrel*ICu_R_crrel
Sig_adjust_R_crrel <- p.adjust(Sig_R_crrel, method = "fdr") # only 1 asv is significant ?? 

# Creating data frame w/ 3 coloums (RR, err, Sig) 
R_crrel = as.data.frame(cbind(RR_crrel,err_R_crrel,Sig_R_crrel, ICl_R_crrel, ICu_R_crrel))
# Creating data frame adjusted 
R_adjust_crrel <- as.data.frame(cbind(RR_crrel,err_R_crrel,Sig_adjust_R_crrel, ICl_R_crrel, ICu_R_crrel))

# Create a new column from row names 
R_crrel$ID <- rownames(R_crrel)

#selecting rows (ASV's) with significance & values #
R_crrel$Significance <- ifelse(ICl_R_crrel > 0 & ICu_R_crrel > 0, "Up",
                   ifelse(ICl_R_crrel < 0 & ICu_R_crrel < 0, "Down", "Not Significant"))

sig_ASVs_crrel <- R_crrel %>%
  filter(Significance != "Not Significant")


```

5A. Adjusting for multiple comparisons and subsetting for the top 100 thaw responders (positive and negative)
```{r}
# Adjust for multiple comparisons 
z_scores_crrel <- RR_crrel / sqv_crrel
p_vals_crrel <- 2 * pnorm(-abs(z_scores_crrel))  # two-tailed test
p_adj_crrel <- p.adjust(p_vals_crrel, method = "fdr")

R_crrel$p_value <- p_vals_crrel
R_crrel$p_adj <- p_adj_crrel
R_crrel$Significant_FDR <- p_adj_crrel < 0.05

thaw.tax <- as.data.frame(thaw_tax)
tax_response_crrel <- left_join(R_crrel, thaw.tax %>% mutate(ID = rownames(thaw.tax)), by = "ID")

tax_sig_crrel <- tax_response_crrel %>% filter(Significance != "Not Significant")
tax_sig_top_crrel <- tax_sig_crrel %>%
  arrange(desc(abs(RR_crrel))) %>%
  slice(1:100)

ggplot(tax_sig_top_crrel, aes(x = reorder(ID, RR_crrel), y = RR_crrel, color = Phylum)) + 
  geom_pointrange(aes(ymin = ICl_R_crrel, ymax = ICu_R_crrel)) +
  geom_hline(yintercept = 0, linetype = 2) +
  facet_wrap(~Phylum, scales = "free_x") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 70, hjust = 1)) +
  labs(x = "ASV ID", y = "Log Response Ratio (Thaw/Post vs Pre)")


#ggplot(tax_sig_top_crrel %>% slice(1:100), aes(x= ID, y = RR_crrel, color = Phylum)) + 
  #geom_pointrange(aes(ymin = RR_crrel - err_R_crrel, ymax = RR_crrel + err_R_crrel)) +
  #geom_hline(yintercept = 0, linetype = 2) +
  #facet_wrap(~Phylum,scales = "free_x") +
  #theme_bw() 
  #theme(axis.text.x = element_text(angle = 70, hjust = 1))



```

5A. Top thaw responders (positive response) (CRREL) 
Pick up here- get rid of NA taxa and use these plots in the manuscript for each site
```{r}
# Top thaw responders (positive response)
tax_sig_up_crrel <- tax_response_crrel %>%
  filter(Significance == "Up")

top_up_crrel <- tax_sig_up_crrel %>%
  filter(!is.na(Genus)) %>%              # <- Remove NA genera
  arrange(desc(RR_crrel)) %>%
  slice(1:50)  # changed to 100


crrel_plot <- ggplot(top_up_crrel, aes(x = reorder(Genus, RR_crrel), y = RR_crrel, color = Phylum)) + 
  geom_pointrange(aes(ymin = ICl_R_crrel, ymax = ICu_R_crrel), color = "#1b9e77") +
  geom_hline(yintercept = 0, linetype = 2) +
  facet_wrap(~Phylum, ncol = 3, scales = "free_x") +
  theme_classic() +
  theme(
    strip.text = element_text(size = 14),
    axis.text.x = element_text(angle = 70, hjust = 1, size = 12),
    axis.text.y = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.title.x = element_text(size = 12),
    legend.position = "none"
  ) +
  labs(x = "Genus", y = "Log Response Ratio")

crrel_plot
  
#ggsave("~/GitHub/thaw.response/Figures/crrel_top100plot.png", width = 12, height = 12, units = "in", dpi = 300)
# # Top thaw decliners (negative response)
# tax_sig_down_crrel <- tax_response_crrel %>%
#   filter(Significance == "Down")
# 
# down_up_crrel <- tax_sig_down_crrel %>%
#   arrange(desc(RR_crrel)) %>%
#   slice(1:100)  # change 20 to however many you want
# 
# ggplot(down_up_crrel, aes(x = reorder(ID, RR_crrel), y = RR_crrel, color = Phylum)) + 
#   geom_pointrange(aes(ymin = ICl_R_crrel, ymax = ICu_R_crrel)) +
#   geom_hline(yintercept = 0, linetype = 2) +
#   facet_wrap(~Phylum, scales = "free_x") +
#   theme_bw() +
#   theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
#   labs(x = "ASV ID", y = "Log Response Ratio (Thaw/Post vs Pre)")
```

# Take these top 100 and add to larger dataframe
- top_up_crrel
- top_up_fl
- top_up_ut

# Changing to top 50 from each site to see if that helps with the  crowding on all sites plot





6. Assessing thaw response by site (Farmer's Loop FL)
```{r}
# Subset the data out for FL sites
t.r.fl <- as.data.frame(thaw_response %>% select(matches("FL2C3|FL1C4")))
ToF_fl <- t.r.fl;
ToF_fl[ToF_fl > 0] <- 1 #anything above 0 is 1 (Presence absence table)

# Which columns are post and which are pre (for the controls and treatment)
pre_col_indices_fl <- grep("PRE", names(t.r.fl))
post_col_indices_fl <- grep("POST", names(t.r.fl))

pre_col_indices_fl
post_col_indices_fl
```

6A. Farmer's Loop thaw response (positive and negative)
```{r}
# Using relative abundance table, creates average
# treatment_names <- c("sample1", "sample2"...)
ave1_fl = apply(t.r.fl[,c(2, 4, 5, 7, 9, 11, 13)],1,mean,na.rm = TRUE);# Control
ave2_fl = apply(t.r.fl[,c(1, 3, 6, 8, 10, 12)],1,mean,na.rm = TRUE)# Treatment 

# Standard Deviation of ALL samples
sd1_fl = apply(t.r.fl[,c(2, 4, 5,7, 9, 11, 13)],1,sd,na.rm = TRUE);# Control #
sd2_fl = apply(t.r.fl[,c(1, 3, 6, 8, 10, 12)],1,sd,na.rm = TRUE)# treatment # 

# How many encounters for ASV number of observations
num1_fl= apply(ToF_fl[,c(2, 4, 5,7, 9, 11, 13)],1,sum,na.rm = TRUE); #Control#
num2_fl = apply(ToF_fl[,c(1, 3, 6, 8, 10, 12)],1,sum,na.rm = TRUE) # Treatment #


# Response ratio calculations
# Sample variance
v1_fl = sd1_fl^2/(ave1_fl^2*num1_fl); # Control 
v2_fl= sd2_fl^2/(ave2_fl^2*num2_fl) # treatment

# Square root of variance 
sqv_fl = sqrt(v1_fl+v2_fl)

# RR = Response ratio
RR_fl = log(ave2_fl/ave1_fl)


# Confidence interval

ICl_R_fl = RR_fl-1.96*sqv_fl
ICu_R_fl = RR_fl+1.96*sqv_fl

# Error 
err_R_fl = ICu_R_fl-RR_fl

# Significance
Sig_R_fl = ICl_R_fl*ICu_R_fl
Sig_adjust_R_fl <- p.adjust(Sig_R_fl, method = "fdr") # only 1 asv is significant ?? 

# Creating data frame w/ 3 coloums (RR, err, Sig) 
R_fl = as.data.frame(cbind(RR_fl,err_R_fl,Sig_R_fl, ICl_R_fl, ICu_R_fl))

R_adjust_fl <- as.data.frame(cbind(RR_fl,err_R_fl,Sig_adjust_R_fl, ICl_R_fl, ICu_R_fl))
# Create a new column from row names 
R_fl$ID <- rownames(R_fl)

#selecting rows (ASV's) with significance & values #
R_fl$Significance <- ifelse(ICl_R_fl > 0 & ICu_R_fl > 0, "Up",
                   ifelse(ICl_R_fl < 0 & ICu_R_fl < 0, "Down", "Not Significant"))

sig_ASVs_fl <- R_fl[R_fl$Significance_fl != "Not Significant", ]

# Adjust for multiple comparisons 
# Adjust for multiple comparisons 
z_scores_fl <- RR_fl / sqv_fl
p_vals_fl <- 2 * pnorm(-abs(z_scores_fl))  # two-tailed test
p_adj_fl <- p.adjust(p_vals_fl, method = "fdr")

R_fl$p_value <- p_vals_fl
R_fl$p_adj <- p_adj_fl
R_fl$Significant_FDR <- p_adj_fl < 0.05

tax_response_fl <- left_join(R_fl, thaw.tax %>% mutate(ID = rownames(thaw.tax)), by = "ID")

tax_sig_fl <- tax_response_fl %>% filter(Significance != "Not Significant")

tax_sig_top_fl <- tax_sig_fl %>%
  arrange(desc(abs(RR_fl))) %>%
  slice(1:100)

ggplot(tax_sig_top_fl, aes(x = reorder(ID, RR_fl), y = RR_fl, color = Phylum)) + 
  geom_pointrange(aes(ymin = ICl_R_fl, ymax = ICu_R_fl)) +
  geom_hline(yintercept = 0, linetype = 2) +
  facet_wrap(~Phylum, scales = "free_x") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 70, hjust = 1)) +
  labs(x = "ASV ID", y = "Log Response Ratio (Thaw/Post vs Pre)")

# Take the top 100 sig fl thaw responders 

```
6B. Top thaw responders (FL Farmer's Loop)
```{r}
# Top thaw responders (positive response)
tax_sig_up_fl <- tax_response_fl %>%
  filter(Significance == "Up")

# obtaining the top 100 by the highest response ratio
top_up_fl <- tax_sig_up_fl %>%
    filter(!is.na(Genus)) %>%  # removing data with NA at the genus level (for plotting)
  arrange(desc(RR_fl)) %>%
  slice(1:50)  # changed to 100

fl_plot <- ggplot(top_up_fl, aes(x = reorder(Genus, RR_fl), y = RR_fl, color = Phylum)) + 
  geom_pointrange(aes(ymin = ICl_R_fl, ymax = ICu_R_fl), color = "#d95f02") +
  geom_hline(yintercept = 0, linetype = 2) +
  facet_wrap(~Phylum, ncol = 3, scales = "free_x") +
  theme_classic() +
    theme(
    strip.text = element_text(size = 14),
    axis.text.x = element_text(angle = 70, hjust = 1, size = 12),
    axis.text.y = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.title.x = element_text(size = 12),
    legend.position = "none"
  ) +
  labs(x = "Genus", y = "Log Response Ratio")
fl_plot 
#ggsave("~/GitHub/thaw.response/Figures/fl_top100plot.png", width = 16.5, height = 12, units = "in", dpi = 300)


# # Top thaw decliners (negative response)
# tax_sig_down_fl <- tax_response_fl %>%
#   filter(Significance == "Down")
# 
# down_up_fl <- tax_sig_down_fl %>%
#   arrange(desc(RR_fl)) %>%
#   slice(1:100)  # change 100 to however many you want
# 
# ggplot(down_up_fl, aes(x = reorder(ID, RR_fl), y = RR_fl, color = Phylum)) + 
#   geom_pointrange(aes(ymin = ICl_R_fl, ymax = ICu_R_fl)) +
#   geom_hline(yintercept = 0, linetype = 2) +
#   facet_wrap(~Phylum, scales = "free_x") +
#   theme_bw() +
#   theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
#   labs(x = "ASV ID", y = "Log Response Ratio (Thaw/Post vs Pre)")

```
7. Assessing thaw responders by site (Utqiagvik, UT)
```{r}
# Subset the data out for CRREL sites
t.r.ut <- as.data.frame(thaw_response %>% select(matches("BEO1|BEO2")))
ToF_ut <- t.r.ut;
ToF_ut[ToF_ut > 0] <- 1 #anything above 0 is 1 (Presence absence table)

# Which columns are post and which are pre (for the controls and treatment)
pre_col_indices_ut <- grep("PRE", names(t.r.ut))
post_col_indices_ut <- grep("POST", names(t.r.ut))

pre_col_indices_ut
post_col_indices_ut
```

7A. UT Thaw response (positive and negative)
```{r}
# Using relative abundance table, creates average
ave1_ut = apply(t.r.ut[,c(2,4,6,8,10,12,14,16)],1,mean,na.rm = TRUE); # Control
ave2_ut = apply(t.r.ut[,c(1,3,5,7,9,11,13,15)],1,mean,na.rm = TRUE); # Treatment

# Standard Deviation of ALL samples
sd1_ut = apply(t.r.ut[,c(2,4,6,8,10,12,14,16)],1,sd,na.rm = TRUE); # Control
sd2_ut = apply(t.r.ut[,c(1,3,5,7,9,11,13,15)],1,sd,na.rm = TRUE); # Treatment

# How many encounters for ASV number of observations
num1_ut = apply(ToF_ut[,c(2,4,6,8,10,12,14,16)],1,sum,na.rm = TRUE); # Control
num2_ut = apply(ToF_ut[,c(1,3,5,7,9,11,13,15)],1,sum,na.rm = TRUE); # Treatment

# Response ratio calculations
v1_ut = sd1_ut^2 / (ave1_ut^2 * num1_ut); # Control
v2_ut = sd2_ut^2 / (ave2_ut^2 * num2_ut); # Treatment

# Square root of variance
sqv_ut = sqrt(v1_ut + v2_ut)

# RR = Response ratio
RR_ut = log(ave2_ut / ave1_ut)

# Confidence interval
ICl_R_ut = RR_ut - 1.96 * sqv_ut
ICu_R_ut = RR_ut + 1.96 * sqv_ut

# Error
err_R_ut = ICu_R_ut - RR_ut

# Significance
Sig_R_ut = ICl_R_ut * ICu_R_ut
Sig_adjust_R_ut <- p.adjust(Sig_R_ut, method = "fdr")

# Creating data frame
R_ut = as.data.frame(cbind(RR_ut, err_R_ut, Sig_R_ut, ICl_R_ut, ICu_R_ut))
R_adjust_ut <- as.data.frame(cbind(RR_ut, err_R_ut, Sig_adjust_R_ut, ICl_R_ut, ICu_R_ut))

# Add ID column
R_ut$ID <- rownames(R_ut)

# Determine direction of significance
R_ut$Significance <- ifelse(ICl_R_ut > 0 & ICu_R_ut > 0, "Up",
                     ifelse(ICl_R_ut < 0 & ICu_R_ut < 0, "Down", "Not Significant"))

# Select significant ASVs
sig_ASVs_ut <- R_ut[R_ut$Significance != "Not Significant", ]

# Adjust for multiple comparisons
z_scores_ut <- RR_ut / sqv_ut
p_vals_ut <- 2 * pnorm(-abs(z_scores_ut))  # two-tailed test
p_adj_ut <- p.adjust(p_vals_ut, method = "fdr")

R_ut$p_value <- p_vals_ut
R_ut$p_adj <- p_adj_ut
R_ut$Significant_FDR <- p_adj_ut < 0.05

# Merge with taxonomy table
tax_response_ut <- left_join(R_ut, thaw.tax %>% mutate(ID = rownames(thaw.tax)), by = "ID")

# Filter significant responses
tax_sig_ut <- tax_response_ut %>% filter(Significance != "Not Significant")

# Top responders
tax_sig_top_ut <- tax_sig_ut %>%
  arrange(desc(abs(RR_ut))) %>%
  slice(1:100)

# Plot
ggplot(tax_sig_top_ut, aes(x = reorder(ID, RR_ut), y = RR_ut, color = Phylum)) + 
  geom_pointrange(aes(ymin = ICl_R_ut, ymax = ICu_R_ut)) +
  geom_hline(yintercept = 0, linetype = 2) +
  facet_wrap(~Phylum, scales = "free_x") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 70, hjust = 1)) +
  labs(x = "ASV ID", y = "Log Response Ratio (Thaw/Post vs Pre)")

```
7B. Top thaw responders (Utqiagvik, UT)
```{r}
# Top thaw responders (positive response)
tax_sig_up_ut <- tax_response_ut %>%
  filter(Significance == "Up")

top_up_ut <- tax_sig_up_ut %>%
  filter(!is.na(Genus)) %>% 
  arrange(desc(RR_ut)) %>%
  slice(1:50)  # change 100 to however many you want


ut_plot <- ggplot(top_up_ut, aes(x = reorder(Genus, RR_ut), y = RR_ut, color = Phylum)) + 
  geom_pointrange(aes(ymin = ICl_R_ut, ymax = ICu_R_ut), color = "#7570b3") +
  geom_hline(yintercept = 0, linetype = 2) +
  facet_wrap(~Phylum, ncol = 3, scales = "free_x") +
  theme_classic() +
  theme(
    strip.text = element_text(size = 14),
    axis.text.x = element_text(angle = 70, hjust = 1, size = 12),
    axis.text.y = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.title.x = element_text(size = 12),
    legend.position = "none"
  ) +
  labs(x = "Genus", y = "Log Response Ratio")
  

ut_plot
#ggsave("~/GitHub/thaw.response/Figures/ut_top100plot.png", width = 12, height = 12, units = "in", dpi = 300)

```


```{r}
# # Top thaw decliners (negative response)
# tax_sig_down_ut <- tax_response_ut %>%
#   filter(Significance == "Down")
# 
# down_up_ut <- tax_sig_down_ut %>%
#   arrange(desc(RR_ut)) %>%
#   slice(1:100)  # change 100 to however many you want
# 
# ggplot(down_up_ut, aes(x = reorder(ID, RR_ut), y = RR_ut, color = Phylum)) + 
#   geom_pointrange(aes(ymin = ICl_R_ut, ymax = ICu_R_ut)) +
#   geom_hline(yintercept = 0, linetype = 2) +
#   facet_wrap(~Phylum, scales = "free_x") +
#   theme_bw() +
#   theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
#   labs(x = "ASV ID", y = "Log Response Ratio (Thaw/Post vs Pre)")

```
8. Common positive thaw responders found in all 3 sites
```{r}
# Really we only care about the positive thaw responders 
# Rename columns in each dataframe
R_fl_renamed <- R_fl %>%
  rename(RR = RR_fl,
         err = err_R_fl,
         Sig = Sig_R_fl,
         ICl_R = ICl_R_fl,
         ICu_R = ICu_R_fl) %>%
  mutate(Site = "FL")

R_crrel_renamed <- R_crrel %>%
  rename(RR = RR_crrel,
         err = err_R_crrel,
         Sig = Sig_R_crrel,
         ICl_R = ICl_R_crrel,
         ICu_R = ICu_R_crrel) %>%
  mutate(Site = "CRREL")

R_ut_renamed <- R_ut %>%
  rename(RR = RR_ut,
         err = err_R_ut,
         Sig = Sig_R_ut,
         ICl_R = ICl_R_ut,
         ICu_R = ICu_R_ut) %>%
  mutate(Site = "UT")

# Get IDs that are significant "Up" responders at all sites
common_up_ASVs <- Reduce(intersect, list(
  tax_sig_up_fl$ID,
  tax_sig_up_crrel$ID,
  tax_sig_up_ut$ID
))

# Bind rows from all sites for only the common "Up" responders
combined_common_up <- bind_rows(
  R_fl_renamed,
  R_crrel_renamed,
  R_ut_renamed
) %>%
  filter(ID %in% common_up_ASVs)

# Add taxonomy
combined_common_up <- left_join(
  combined_common_up,
  thaw.tax %>% mutate(ID = rownames(thaw.tax)),
  by = "ID"
)

ggplot(combined_common_up, aes(x = reorder(ID, RR), y = RR, color = Phylum)) +
  geom_pointrange(aes(ymin = ICl_R, ymax = ICu_R)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~Site, scales = "free_x") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 70, hjust = 1),
        strip.text = element_text(face = "bold")) +
  labs(x = "ASV ID", y = "Log Response Ratio (Thaw/Post vs Pre)",
       title = "Common Positive Thaw Responders Across Sites")

ggplot(combined_common_up, aes(x = reorder(Genus, RR), y = RR, color = Phylum)) +
  geom_pointrange(aes(ymin = ICl_R, ymax = ICu_R)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~Site, scales = "free_x") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 70, hjust = 1),
        strip.text = element_text(face = "bold")) +
  labs(x = "Genus", y = "Log Response Ratio (Thaw/Post vs Pre)",
       title = "Common Positive Thaw Responders Across Sites")
```


```{r}
library(dplyr)
library(tidyr)



ggplot(combined_common_up, aes(x = reorder(Genus, RR), y = RR, color = Family)) +
  geom_pointrange(aes(ymin = ICl_R, ymax = ICu_R)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~Site, scales = "free_x") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 70, hjust = 1),
        strip.text = element_text(face = "bold")) +
  labs(x = "Genus", y = "Log Response Ratio (Thaw/Post vs Pre)",
       title = "Common Positive Thaw Responders Across Sites")
```


```{r}

#Rename the sites for publication
library(dplyr)

combined_common_up <- combined_common_up %>%
  mutate(Site = recode(Site,
                       "CRREL" = "PT",
                       "UT" = "BEO",
                       "FL" = "FL"))
library(stringr)
# Recode site before plotting so site shows up correctly
# combined_common_up <- combined_common_up %>%
#   mutate(
#     Site = recode(Site,
#                   "CRREL" = "CRREL",
#                   "FL" = "Farmers Loop",
#                   "UT" = "Utqiagvik")
#   )

# Remove genus = NA 
combined_common_up <- combined_common_up %>%
  filter(!is.na(Genus))


combined_common_up <- combined_common_up %>%
mutate(Genus_wrapped = case_when(
Genus == "Candidatus Planktophila" ~ "Candidatus\nPlanktophila",
Genus == "Burkholderia-Caballeronia-Paraburkholderia" ~ "Burkholderia-Caballeronia-\nParaburkholderia",
TRUE ~ Genus
))
response.family <- ggplot(combined_common_up, aes(x = reorder(Genus_wrapped, RR), y = RR, color = Site)) +
  geom_pointrange(aes(ymin = ICl_R, ymax = ICu_R), position = position_dodge(width = 0.6)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~Family, scales = "free_x", ncol = 3) +
  theme_classic() +
  theme(axis.text.x = element_text(size = 10), # add angle = 70
        strip.text = element_text(face = "bold")) +
  labs(x = "Genus", y = "Log Response Ratio",
       title = "") +
  scale_color_manual(
    values = c("#1b9e77", "#d95f02", "#7570b3"),
    name = "Site")

ggsave("~/GitHub/thaw.response/Figures/response.family.png", response.family, width = 10, height = 4, dpi = 300)

```


```{r}
# Determining fold change
combined_common_up$fold_change <- exp(combined_common_up$RR)

```



Editing:
Okay right away this is wrong because it needs to come from the top up datasets 
```{r}
# Bringing all the top 100 positive thaw responders together to one dataset
# First, add a column to each dataset identifying the site
tax_sig_top_fl <- top_up_fl %>% mutate(Site = "FL")
tax_sig_top_crrel <- top_up_crrel %>% mutate(Site = "CRREL")
tax_sig_top_ut <- top_up_ut %>% mutate(Site = "UT")

# Rename the columns of each site dataset so we can merge smoothly

library(dplyr)

tax_sig_top_crrel <- tax_sig_top_crrel %>%
  rename(
    RR   = RR_crrel,
    err_R = err_R_crrel,
    Sig_R = Sig_R_crrel,
    ICl   = ICl_R_crrel,
    ICu   = ICu_R_crrel
  ) %>%
  mutate(original_Site = "CRREL")

tax_sig_top_fl <- tax_sig_top_fl %>%
  rename(
    RR   = RR_fl,
    err_R = err_R_fl,
    Sig_R = Sig_R_fl,
    ICl   = ICl_R_fl,
    ICu   = ICu_R_fl
  ) %>%
  mutate(original_Site = "FL")

tax_sig_top_ut <- tax_sig_top_ut %>%
  rename(
    RR   = RR_ut,
    err_R = err_R_ut,
    Sig_R = Sig_R_ut,
    ICl   = ICl_R_ut,
    ICu   = ICu_R_ut
  ) %>%
  mutate(original_Site = "UT")

metaRR <- bind_rows(tax_sig_top_crrel, tax_sig_top_fl, tax_sig_top_ut)



# Then bind them together
metaRR <- bind_rows(tax_sig_top_crrel, tax_sig_top_fl, tax_sig_top_ut)
library(dplyr)
library(tidyr)

# Rename existing Site column to avoid conflict
#metaRR2 <- metaRR %>%
  #rename(original_Site = Site)

#Rename the sites for publication
library(dplyr)

metaRR2 <- metaRR %>%
  mutate(original_Site = recode(original_Site,
                       "CRREL" = "PT",
                       "UT" = "BEO",
                       "FL" = "FL"))

# #metaRR_long <- metaRR2 %>%
#   pivot_longer(
#     cols = matches("RR_|ICl_|ICu_"),
#     names_to = c(".value", "Site"),
#     names_pattern = "(RR|ICl|ICu)_(.*)",
#     names_repair = "unique"  # just in case
#   ) %>%
#   filter(!is.na(RR))



ggplot(metaRR2, aes(x = Genus, y = RR, color = original_Site)) +
  geom_pointrange(aes(ymin = ICl, ymax = ICu), position = position_dodge(width = 0.6)) +
  geom_hline(yintercept = 0, linetype = 2) +
  facet_wrap(~Phylum, scales = "free_x") +
  theme_classic() +
    theme(axis.text.x = element_text(angle = 70, hjust = 1, size = 6),
        strip.text = element_text(face = "bold")) +
  labs(x = "Genus", y = "Log Response Ratio",
       title = "") +
  scale_color_manual(
    values = c("#7570b3", "#d95f02","#1b9e77"),
    name = "Site")

ggsave("~/GitHub/thaw.response/Figures/all_top50plot.png", width = 12, height = 8)
#theme(axis.text.x = element_text(angle = 70, hjust = 1)) +
  #labs(x = "Genus", y = "Log Response Ratio (Thaw/Post vs Pre)")
```


```{r}
# Trying to make the plot more readabable

library(dplyr)
library(ggplot2)
library(tidytext)   # for reorder_within() / scale_x_reordered()

# order genera within each phylum by mean RR (use median/|RR| if you prefer)
ord <- metaRR %>%
  group_by(Phylum, Genus) %>%
  summarize(meanRR = mean(RR, na.rm = TRUE), .groups = "drop")

plot_df <- metaRR %>%
  left_join(ord, by = c("Phylum","Genus")) %>%
  mutate(Genus_ord = reorder_within(Genus, meanRR, Phylum))

ggplot(plot_df, aes(x = Genus_ord, y = RR, color = original_Site)) +
  geom_pointrange(
    aes(ymin = ICl, ymax = ICu, group = original_Site),
    position = position_dodge(width = 0.4),
    linewidth = 0.3
  ) +
  geom_hline(yintercept = 0, linetype = 2) +
  coord_flip() +
  facet_wrap(~ Phylum, scales = "free_y") +
  scale_x_reordered() +
  theme_classic() +
  theme(
    strip.text = element_text(face = "bold"),
    axis.text.y = element_text(size = 10),
    panel.spacing.y = unit(6, "pt"),
    legend.position = "top"
  ) +
  guides(color = guide_legend(nrow = 1)) +
  labs(x = "", y = "Log Response Ratio") +
  scale_color_manual(values = c("#1b9e77", "#d95f02", "#7570b3"), name = "Site")
ggsave("~/GitHub/thaw.response/Figures/all_top100plot_readable.png", width = 12, height = 8)
```


# Come back and do the following:
- Check and vet all parts of this script (done
- Get significance values to report 
- Thaw responders for all sites --> positive --> common positive thaw responders across sites --> (top 100 from each site) --> plot the common thaw responders with the multi-point phylum plot 
- 
Redo: Okay bringing together the positive only datasets (not the top 100 of the positive)
```{r}
library(dplyr)
tax_sig_up_fl <- top_sig_up_fl %>% mutate(Site = "FL")
tax_sig_up_crrel <- top_sig_up_crrel %>% mutate(Site = "CRREL")
tax_sig_up_ut <- top_sig_up_ut %>% mutate(Site = "Utqiagvik")

```


```{r}
# Really we only care about the positive thaw responders 
# Rename columns in each dataframe
R_fl_renamed <- R_fl %>%
  rename(RR = RR_fl,
         err = err_R_fl,
         Sig = Sig_R_fl,
         ICl_R = ICl_R_fl,
         ICu_R = ICu_R_fl) %>%
  mutate(Site = "FL")

R_crrel_renamed <- R_crrel %>%
  rename(RR = RR_crrel,
         err = err_R_crrel,
         Sig = Sig_R_crrel,
         ICl_R = ICl_R_crrel,
         ICu_R = ICu_R_crrel) %>%
  mutate(Site = "CRREL")

R_ut_renamed <- R_ut %>%
  rename(RR = RR_ut,
         err = err_R_ut,
         Sig = Sig_R_ut,
         ICl_R = ICl_R_ut,
         ICu_R = ICu_R_ut) %>%
  mutate(Site = "UT")

# Get IDs that are significant "Up" responders at all sites
common_up_ASVs <- Reduce(intersect, list(
  tax_sig_up_fl$ID,
  tax_sig_up_crrel$ID,
  tax_sig_up_ut$ID
))

# Bind rows from all sites for only the common "Up" responders
combined_common_up <- bind_rows(
  R_fl_renamed,
  R_crrel_renamed,
  R_ut_renamed
) %>%
  filter(ID %in% common_up_ASVs)

# Add taxonomy
combined_common_up <- left_join(
  combined_common_up,
  thaw.tax %>% mutate(ID = rownames(thaw.tax)),
  by = "ID"
)

ggplot(combined_common_up, aes(x = reorder(ID, RR), y = RR, color = Phylum)) +
  geom_pointrange(aes(ymin = ICl_R, ymax = ICu_R)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~Site, scales = "free_x") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 70, hjust = 1),
        strip.text = element_text(face = "bold")) +
  labs(x = "ASV ID", y = "Log Response Ratio (Thaw/Post vs Pre)",
       title = "Common Positive Thaw Responders Across Sites")

ggplot(combined_common_up, aes(x = reorder(Genus, RR), y = RR, color = Phylum)) +
  geom_pointrange(aes(ymin = ICl_R, ymax = ICu_R)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  facet_wrap(~Site, scales = "free_x") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 70, hjust = 1),
        strip.text = element_text(face = "bold")) +
  labs(x = "Genus", y = "Log Response Ratio (Thaw/Post vs Pre)",
       title = "Common Positive Thaw Responders Across Sites")
```

# Pre and post thaw abundance plots 


```{r}
library(tidyverse)
absolute_otu <- as.data.frame(absolute_otu)


abund_long <- absolute_otu %>%
  rownames_to_column("Species") %>%
  pivot_longer(-Species, names_to = "Sample", values_to = "Abundance") %>%
  separate(Sample, into = c("Site","Rep","Time","Type"), sep = "_") %>%
  mutate(Site = case_when(
    str_detect(Site, "AT")  ~ "PT",
    str_detect(Site, "BEO") ~ "BEO",
    str_detect(Site, "FL")  ~ "FL",
    TRUE ~ Site   # keep as-is if none match
  ))

thaw.tax <- thaw.tax %>%
  rownames_to_column("Species")
abund_meta <- abund_long %>%
  left_join(thaw.tax, by = "Species")


top_phyla <- abund_meta %>%
  group_by(Phylum) %>%
  summarise(total = sum(Abundance)) %>%
  top_n(10, total) %>%
  pull(Phylum)

abund_top <- abund_meta %>%
  mutate(Phylum = ifelse(Phylum %in% top_phyla, Phylum, "Other"))

ggplot(top_phyla, aes(x = Time, y = Abundance, fill = Phylum)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ Site, scales = "free_y") +
  theme_minimal() +
  labs(y = "Absolute abundance", x = "Pre vs Post")



```
Pre post abundance plots - relative abundance 
```{r}
library(phyloseq)

# Transform to relative abundance (per sample, sums to 1)
thaw_relabund <- transform_sample_counts(thaw, function(x) x / sum(x))
thaw_phylum <- tax_glom(thaw_relabund, taxrank = "Phylum")
library(ggplot2)

plot_bar(thaw_phylum, 
         x = "Site",   # or whatever variable you have in sample_data(thaw)
         fill = "Phylum") +
  facet_wrap(~ Site, scales = "free_x") +   # split Pre vs Post
  theme_minimal() +
  ylab("Relative abundance") +
  scale_y_continuous(labels = scales::percent)



```

# Running response ratio analysis with the clr transformed data to see if rarefying makes any difference in the results (to respond to reviewer 1)
# need to do based on absolute abundance as well 

# Load in data
```{r}
clr_data <- read.csv("~/GitHub/thaw.response/Data/ASV_table_clr_transformed_wide_format.csv")
```


3A. Assessing thaw response (all 3 sites combined and all responders)
```{r}

# dont know why i had to remove samples 41 and 42 from the line of number code below 
ToF <- clr_data #changed from thaw_response to thaw.abs
#ToF[ToF > 0] <- 1 #anything above 0 is 1 (Presence absence table) ##

# Which columns are post and which are pre (for the controls and treatment)
pre_col_indices <- grep("PRE", names(clr_data))
post_col_indices <- grep("POST", names(clr_data))

pre_col_indices
post_col_indices

# Using relative abundance table, creates average
# treatment_names <- c("sample1", "sample2"...)
# Control (your first list)

ave1 = apply(clr_data[,c(3,5,7,9,11,13,15,17,19,21,23,25,27,29,31,33,35,37,39,41,43,46,48,50,52)],1,mean,na.rm = TRUE);#Control
ave2 = apply(clr_data[,c(2,4,6,8,10,12,14,16,18,20,22,24,26,28,30,32,34,36,38,40,42,44,45,47,49,51)],1,mean,na.rm = TRUE)# Treatment 


# Standard Deviation of ALL samples
sd1 = apply(clr_data[,c(3,5,7,9,11,13,15,17,19,21,23,25,27,29,31,33,35,37,39,41,43,46,48,50,52)],1,sd,na.rm = TRUE);# Control #
sd2 = apply(clr_data[,c(2,4,6,8,10,12,14,16,18,20,22,24,26,28,30,32,34,36,38,40,42,44,45,47,49,51)],1,sd,na.rm = TRUE)# treatment # 

# How many encounters for ASV number of observations
num1 = apply(ToF[,c(3,5,7,9,11,13,15,17,19,21,23,25,27,29,31,33,35,37,39,41,43,46,48,50,52)],1,sum,na.rm = TRUE); #Control#
num2 = apply(ToF[,c(2,4,6,8,10,12,14,16,18,20,22,24,26,28,30,32,34,36,38,40,42,44,45,47,49,51)],1,sum,na.rm = TRUE) # Treatment

```
3B. Response ratio calculations (all sites)
```{r}
# Response ratio calculations
# Sample variance
v1 = sd1^2/(ave1^2*num1); # Control #
v2 = sd2^2/(ave2^2*num2) # treatment #

# Square root of variance 
sqv = sqrt(v1+v2)

# RR = Response ratio
RR = log(ave2/ave1)
```

3C. Confidence interval (all sites)
```{r}
# Confidence interval

ICl_R = RR-1.96*sqv
ICu_R = RR+1.96*sqv

# Error 
err_R = ICu_R-RR

# Significance
Sig_R = ICl_R*ICu_R
Sig_adjust_R <- p.adjust(Sig_R, method = "fdr") 

# Creating data frame w/ 3 coloums (RR, err, Sig) 
R = as.data.frame(cbind(RR,err_R,Sig_R, ICl_R, ICu_R))

R_adjust <- as.data.frame(cbind(RR,err_R,Sig_adjust_R, ICl_R, ICu_R))
# Create a new column from row names 
R$ID <- rownames(R)

#selecting rows (ASV's) with significance & values #
R$Significance <- ifelse(ICl_R > 0 & ICu_R > 0, "Up",
                   ifelse(ICl_R < 0 & ICu_R < 0, "Down", "Not Significant"))

sig_ASVs <- R[R$Significance != "Not Significant", ]

# Adjust for multiple comparisons 
z_scores <- RR / sqv
p_vals <- 2 * pnorm(-abs(z_scores))  # two-tailed test
p_adj <- p.adjust(p_vals, method = "fdr")

R$p_value <- p_vals
R$p_adj <- p_adj
R$Significant_FDR <- p_adj < 0.05

thaw.tax <- as.data.frame(thaw_tax)

tax_response <- left_join(R, thaw.tax %>% mutate(ID = rownames(thaw.tax)), by = "ID")

tax_sig <- tax_response %>% filter(Significance != "Not Significant")

tax_sig_top <- tax_sig %>%
  arrange(desc(abs(RR))) %>%
  slice(1:100)

ggplot(tax_sig_top, aes(x = reorder(ID, RR), y = RR, color = Phylum)) + 
  geom_pointrange(aes(ymin = ICl_R, ymax = ICu_R)) +
  geom_hline(yintercept = 0, linetype = 2) +
  facet_wrap(~Phylum, scales = "free_x") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 70, hjust = 1)) +
  labs(x = "ASV ID", y = "Log Response Ratio (Thaw/Post vs Pre)")



```

